# Chapter Six - The Military After 9/11

On September 11, 2001, terrorists hijacked several commercial airplanes and crashed them into several sites, including the twin towers of the World Trade Center in New York City, the Pentagon in Washington D.C., and a field in Western Pennsylvania.^[Heroic Americans prevented this plane from reaching its intended target, the US Capitol or the White House, in Washington D.C. See @kean_911_2004, 14 for more details.] That same evening, members of Congress gathered on the steps of the US capitol to collectively express solidarity and resolve. After a moment of silence and short remarks by then speaker of the House Dennis Hastert (R-IL), those who were gathered sang the song, "God Bless America" [@us_congress_members_2001].

Seventeen months later, in February 2003, the Senate Armed Services Committee held a hearing on war plans for an imminent invasion of Iraq, which began the following month. One exchange was particularly noteworthy. General Eric Shinseki, the Army Chief of Staff, told the committee that he thought "something on the order of several hundred thousand..." troops were likely required to stabilize Iraq after the invasion [Shinseki, 2003 quoted in @fallows_karmic_2008]. Shinseki's comments ran counter to the views held by the Secretary of Defense, Donald Rumsfeld, and his second in command, Paul Wolfowitz, both of whom believed that a smaller, lighter military footprint was more than adequate for invading and stabilizing Iraq. In the days after the hearing, both Wolfowitz and Rumsfeld publicly criticized Shinseki's remarks. Wolfowitz, in fact, dismissed Shinseki's estimate as "wildly off the mark" [@schmitt_pentagon_2003]. General Shinseki, for his part, remained mostly silent, and retired from the Army shortly afterward.  

Eighteen and a half years later, the United States withdrew from Afghanistan after the resurgent Taliban rapidly displaced the Afghan National Government in the summer of 2021. During a Congressional hearing held on 28 September 2021 about the withdrawal, Tom Cotton, a veteran and the junior Republican Senator from Arkansas, asked the Chairman of the Joint Chiefs of Staff, General Mark Milley, "why haven't you resigned" [@singman_milley_2021]? 

The Senator's question alluded to the fact that Milley, weeks before the collapse of the Afghan Government, had recommended to President Biden that he maintain approximately 2,500 US Troops in Afghanistan, a recommendation the President did not heed. In response to Cotton's question, Milley stated, "It would be an incredible act of political defiance for a commissioned officer to just resign because my advice is not taken...this country doesn't want generals figuring out what orders we are going to accept and do or not. That's not our job [@singman_milley_2021]." 

In the same hearing, however, before ever even discussing the issue of Afghanistan, General Milley addressed allegations that he had inappropriately conversed with both a Chinese general and with the Speaker of the House, Representative Nancy Pelosi (D-CA), during the transition between Presidents Trump and Biden [@baldor_gen_2021]. These allegations, which are described in detail later in this chapter, provoked strong reactions from several Republican lawmakers in the days leading up to the hearing. 

This chapter identifies, compares, and contrasts the patterns of political behavior engaged in by military and civilian actors in the post-9/11 era. Similar to the previous chapter, one of the independent variables primarily varies more so than the other over the period of time examined. From approximately 2001 - 2021, the level of military prestige remained relatively high, but the level of political polarization rose sharply. 

The chapter proceeds in five parts. First, this chapter describes the predictions concerning the patterns of political behavior involving the military as a democracy moves from a domestic environment marked by low polarization and high military prestige to one marked by high polarization and high military prestige. In this section, I substantiate the claim that this change occurred in the United States from 2001 - 2021. Second, this chapter describes a major episode that occurred during the first portion of this twenty year period. This first episode examined is the 2006 so-called "Revolt of the Generals," which involved several retired senior military officers publicly castigating the Secretary of Defense, Donald Rumsfeld. 

The third part of the chapter then explores an episode that occurred during the second half of the period. This latter episode involves the political behaviors undertaken by military actors coinciding with a period of domestic civil unrest in June 2020 and political behaviors undertaken by military and civilian actors in the wake of the US withdrawal from Afghanistan in 2021. The fourth part of this chapter evaluates, compares, and contrasts the political behaviors involving the military undertaken in these two episodes (2006 compared to 2020 - 2021). The fifth and final part of this chapter briefly compares and contrasts portions of this case study to portions of the case study undertaken in the previous chapter.

Similar to the previous chapter, the central focus in this case study is on the tone, tenor, and characteristics of the political behaviors involving the military that occur during each of the episodes examined. The chief argument made in this chapter is that, in accordance with this dissertation's theoretical predictions, the rise in the level of political polarization from 2001 - 2021 contributed to several discernible changes in _how_ civilian and military actors politicized the military.

\singlespace
## Theoretical Predictions for the Post-9/11 Era, 2001-2021
\doublespace

###  Low Polarization and High Military Prestige

The theory presented in chapter two argued that a state's domestic environment strongly shapes the patterns of political behavior that occur involving the military. When the level of military prestige is high, both civilian and military actors are more likely to behave in ways that harness or leverage the prestige of the military, even if doing so violates one or more of the principles of civil-military relations. This theoretical expectation is rooted in the political communication literature, which argues that the public listens to voices who it believes are "knowledgeable" and "trustworthy" [@lupia_democratic_1998, 69-76; see also @zaller_nature_1992, 47]. The theory argued that as the level of military prestige rises, the public increasingly considers the military and its members to possess these traits. 

But the theory presented in chapter two also argued that the level of polarization shapes how civilian and military actors behave with respect to the military. When the level of polarization is low, the theory argued that there is relatively little basis for ideological division between and among military actors. This is because periods of relatively low political polarization indicate that few if any differences in the political space involve issues that are related to the moral values and/or worldviews that military actors hold to. This is not to say that military actors are entirely unified, or that there is little chance of conflict between military actors, when polarization is low. Rather, the theory predicts that absent high levels of polarization, there is little if any ideological basis, or _motive_, for military actors to behave in ways that violate the central principles of civil-military relations.

In environments marked by the presence of low polarization and high military prestige, the theory ultimately predicts that either military or civilian actors (or both) may induce politicization of the military by engaging in behaviors that violate the principles of civil-military relations in an effort to shape public opinion. The theory even noted that such an environment is far from tranquil, and that there is a high potential for civil-military conflict altogether, as shown in the bottom left quadrant of Table \@ref(tab:poss-environment3). 

```{r poss-environment3, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env_revised2 <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "No politicization; low levels of political activity; low potential for civil-military conflict", 
  "Military (induced) politicization; moderate levels of political activity; moderate potential for civil-military conflict",  
  "High Prestige", "Military and Civilian (induced) politicization; moderate levels of political activity; high potential for civil-military conflict", 
  "Comprehensive politicization; extreme levels of political activity; significant potential for civil-military conflict" 
  )
 

  kable(text_env_revised2, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors and the Conduct of Civil-Military Relations") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2:3, width="18em") 
 
```

Historically, there have been several significant civil-military relations disputes that have occurred under the domestic conditions of low polarization and high military prestige in the United States. For example, disputes between General Douglas MacArthur and President Harry Truman over Korea, General Matthew Ridgeway and President Dwight Eisenhower over Ike's New Deal policies in the early 1950s, and Colin Powell and President Bill Clinton over several developing post-Cold War crises in the early 1990s, all occurred in domestic environments marked by relatively low polarization and high military prestige. In short, the historical record indicates that a state's civil-military relationship will be far from free of discord when polarization is low and when military prestige is high. On the contrary, an environment marked by high military prestige and low polarization can produce significant civil-military tension. 

### High Polarization and High Military Prestige

The theory does predict, however, that as polarization rises, the potential for civil-military conflict rises even further. This prediction stems from the relationship between rising political polarization and growing ideological division that occurs among military actors. The theory argued that military actors, like the rest of society, become more polarized as the level of polarization grows. Therefore, holding all else equal, as the level of polarization rises, military actors are more likely to engage in political behaviors for ideological reasons that violate the principles of civil-military relations. 

This occurs, the theory argued, because of rising levels of polarization indicate. In chapter two, the theory argued that polarization, at its root, attests to a loss of consensus or agreement concerning the moral values and ideals that hold sway in society [@marsden_twilight_2014; @mann_its_2012; @allen_why_2020]. In highly polarizing times such as the contemporary age in the United States, the theory argued that many political conflicts that emerge are not simply political. That is, many contemporary polarizing conflicts do not simply involve questions concerning whether society should enact policy "X" or policy "Y", but rather involve conflicts over the worldviews, moral values, and philosophical foundations that underlie policy "X" and policy "Y". 

In making this claim, chapter two highlighted several polarizing issues in the contemporary United States. These include the question of what it means to be an American, whether America should be proud of its history, whether abortion is a "right," and whether the construct of gender is important anymore, and to what degree. What these and other polarizing issues have in common is that they cannot be solved, let alone reasonably and intelligently discussed in most cases, without an individual explicitly or implicitly referencing or alluding to some sort of moral worldview commitment. 

The theory further argued that polarization impacts the conduct of civil-military relations because a military actor's commitment to the worldviews or values underlying any number of polarizing issues may in fact be stronger than that same actor's level of commitment to the central principles of civil-military relations. Stated differently, rising levels of polarization eventually introduce situations that challenge ultimate conceptions of authority, especially for military actors. An officer in the military likely feels some level of commitment to the principles of civilian control, non-partisanship of the military institution, and non-interference. But the same officer likely also has a worldview that relies on moral values or commitments of some kind, even if that officer cannot articulate what that worldview is [@sire_universe_2020]. 

In highly polarizing environments, military actors more frequently encounter situations in which they have to choose between adhering to the principles of civil-military relations or advancing the values and principles that underlie any number of polarizing issues that are at stake in the political sphere. When confronted with such a decision, especially when the consequences are great and much is at stake, military actors will increasingly act in accordance with the values that are associated with the worldview to which they hold dear, and thus, increasingly behave in ways that violate the principles of civil-military relations.

As Table \@ref(tab:poss-environment3) notes, simultaneously high levels of polarization and military prestige lead to a domestic environment in which there is comprehensive politicization, extreme levels of political activity involving the military, and significant potential for civil-military conflict. While it is difficult to determine exactly what form these characteristics might take in practice, or how they might manifest themselves into distinct political behaviors, these characteristics suggest that there is greater potential for a state's civil-military relationship to be riddled with tension when the levels of polarization and military prestige are both high compared to when military prestige is high but polarization is low.  

### Substantiation of Environmental Characteristics, 2001 - 2021

To substantiate the claim that the level of polarization increased and the level of military prestige generally remained constant (but relatively high) over the period of 2001 - 2021, I rely on measurable quantitative data. This is different from the previous chapter, which relied on mainly qualitative data to substantiate that claim that the level of military prestige declined throughout the Reconstruction Era. Figure \@ref(fig:case2-polar) plots the level of polarization in both the US House and the US Senate, as well as the level of trust and confidence in the US military, from 2001 - 2021. The data for polarization comes from DW-NOMINATE scores, which use roll-call voting for members of Congress to measure polarization over time [@jeffrey_b_lewis_voteview_2020]. The data for trust in the military comes from public opinion measures of trust and confidence in the military over time [@noauthor_confidence_2020].

```{r case2-polar-prep, include=FALSE, warning=FALSE, echo=FALSE}

df_trust_polar_cs2 <- read_csv(here("data","polar-inst-trust-time.csv"))

df_trust_polar_cs2 <- df_trust_polar_cs2 %>% filter(Year >= 2001) %>% select(-Congress) %>% 
  mutate(
    house_polar_dim1 = house_polar_dim1 *100,
    sen_polar_dim1 = sen_polar_dim1 *100,
    ) %>% pivot_longer(Military:sen_polar_dim1, names_to = "value", values_to = "Score") %>% distinct()
```

```{r case2-polar, include=T, warning=FALSE, echo=F, message=F, fig.height=4, fig.width=6, fig.cap = "Public Trust (L) and Polarization (R), 2001-2021" }
df_trust_polar_cs2 %>% ggplot(aes(x=Year, y=Score, color=value)) + geom_line() + geom_point(size=0.7) + labs(x="Year", y="Trust in Institutions") + scale_color_discrete(name="Trust/\nPolarization",
                         breaks=c("Military", "house_polar_dim1", "sen_polar_dim1"),
                         labels=c("Trust in\nMilitary","House\nPolarization", "Senate\nPolarization")) + scale_y_continuous(sec.axis = sec_axis(~., name = "Polarization (Scaled)")) + scale_x_continuous(breaks=seq(2001, 2021, 2)) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE))
```

Figure \@ref(fig:case2-polar) shows that polarization and military prestige have changed over the period from 2001-2021, but in different ways. The polarization levels in both the US House and Senate have generally consistently increased. When polarization has dropped (2007, 2009, and 2019 in the House; 2003 in the Senate), these drops have been somewhat small, and more importantly, outpaced by the numerous increases that have occurred over the twenty year period.

The level of public trust in the military, on the other hand, has exhibited greater variation between years. For example, a relative spike in public trust in the military occurred in 2003, perhaps reflecting a boost in patriotic fervor as the war in Iraq began the same year. Similarly, noticeable declines in public trust in the military occurred in 2007, perhaps reflecting impatience over mounting US casualties in Iraq, and again in 2021, perhaps reflecting public frustration over the US withdrawal from Afghanistan. Yet even in the presence of this variability, the level of military prestige that has endured for the past twenty years is relatively high, and certainly higher than the level that existed in the late 1970s and early 1980s after the Vietnam War. Since data has been kept, for example, the lowest recorded level of trust in the military occurred in 1981, when only 50% of surveyed Americans said they trusted the military a "great deal or quite a lot" [@noauthor_confidence_2020]. In short, the period 2001 - 2021 broadly reflects a domestic environment in which the level of military prestige remained relatively constant (and high in historical terms), but in which the level of polarization increased. 

Further note that Figure \@ref(fig:case2-polar) does not, and indeed cannot, definitively conclude when the shift from a relatively moderate to relatively high level of polarization occurred. However, the figure does show a fairly obvious spike in the level of polarization in the US House (red line) beginning in 2011, a spike accompanied by a smaller yet still detectable spike in polarization in the US Senate (blue line) in the same year. From a methodological standpoint, then, the analysis which follows takes the year 2011 as a sort of natural breaking point separating relatively lower and higher levels of polarization. Therefore, this chapter will describe two episodes that each occurs well before and after the year 2011. I first describe and analyze the 2006 "Revolt of the Generals" through the lens of the three central principles of civil-military relations. I then do the same with respect to a number of political behaviors undertaken by military and civilian actors involving the military in 2020 and 2021. 

\singlespace
## Part 1 - The War in Iraq and The Revolt of the Generals, 2006
\doublespace

The 2006 "Revolt of the Generals" narrowly refers to the behavior of several recently retired senior military officers who, in the Spring of 2006, publicly criticized the Secretary of Defense, Donald Rumsfeld, and specifically, the way in which Rumsfeld had managed the war in Iraq up until that point. By 2006, the war in Iraq was three years old, and it had become abundantly clear by then that the war would not be quick or easy. 

In a broad sense, the "Revolt of the Generals" can be viewed as the military's backlash to their perception of not being heard, listened to, or valued by their civilian chain of command within the Department of Defense [@snider_strategic_2014; @binkley_revisiting_2020]. In this sense, the "Revolt" was reminiscent of similar critiques of civilian leadership by military actors, such as H.R. McMaster's _Dereliction of Duty_, which charged the top military brass during Vietnam with failing to adequately push back against the wartime policies of President Lyndon Johnson and Secretary of Defense Robert McNamara [@mcmaster_dereliction_1997]. 

The "Revolt of the Generals" also served as the means by which several military actors responded to what they perceived to be the disparaging of General Shinseki by his civilian superiors, which was referenced in this chapter's introduction. After telling Congress that he thought far more troops were required to stabilize Iraq than his civilian superiors thought were necessary, Shinseki's last months in uniform had proven somewhat turbulent. Rumsfeld, for instance, did not even attend Shinseki's retirement ceremony, a significant break in protocol that rankled some senior military officers [@shanker_retiring_2003].

The timing of the "Revolt of the Generals" also matters a great deal. Not only was the war in Iraq three years old in the Spring of 2006, but that same year, Congressional midterm elections were held. The behaviors conducted as part of the "Revolt of the Generals" ultimately elevated the status of the war in Iraq as an issue, and in all likelihood, contributed to the eventual dismissal of Secretary Rumsfeld by the President, George W. Bush, after the midterm elections.  

### Political Behaviors Undertaken, 2006

The immediate slate of military actors who participated in the "Revolt of the Generals" consisted of six retired military officers, including Army Major General Paul Eaton, Marine Lieutenant General Greg Newbold, Army Major General John Batiste, Marine Lieutenant General Paul Van Riper, Army Major General John Riggs, and Army Major General Charles Swannack. In chapter three of this dissertation, which examined public opinion commentary authored by retired military actors, I discussed the behavior of retired military officers in broad detail, and so I will not readdress those considerations here. 
But it is important to point out that the six retired generals who participated in the "Revolt of the Generals" were unique in several ways. 

The most important of these is that these officers had only very recently retired, and from fairly high profile positions in the military. The "revolting" officers, in other words, had directly participated in the war in Iraq, either on the ground or during deliberations involving the war. Eaton, for example, had been responsible for training the Iraqi military from 2003 - 2004, and both Swannack and Batiste had served as Army Division Commanders early during the war [@cloud_more_2006]. The broader implication from the recent retirements of the officers participating in the "Revolt" is that they had deliberately chosen to retire and speak out _rather than_ remain in uniform. 

Furthermore, the actions of these six retired officers spurred other military actors to similarly engage in political behaviors that addressed similar topics. The considerable press coverage generated by the "Revolt of the Generals" led to a number of civilian actors commenting on the allegations levied against Rumsfeld, and on the broader war effort underway in Iraq. By considering the behaviors undertaken both as part of the "Revolt" _and_ those undertaken directly in response to it, the reader is better positioned to understand the full impacts of this episode. 

### Public Appeals

The main political behaviors undertaken by the six military actors who participated in the "Revolt of the Generals" were numerous public appeals. These took various forms, including publishing opinion articles and giving interviews to the press. For example, in March and April of 2006, Generals Eaton, Batiste, and Newbold penned articles that appeared in _The New York Times_, _The Washington Post_, and _Time_ magazine, respectively. The articles levied multiple criticisms of Rumsfeld, including that he had micromanaged the war effort and failed to heed the advice of senior military leaders during the planning and execution of the war [@eaton_top-down_2006; @batiste_case_2006; @newbold_why_2006]. In conjunction with the publication of these articles, Generals Van Riper, Riggs, and Swannack then gave interviews in which they "concurred with the criticisms levied by Eaton, Newbold, and Batiste" [@binkley_revisiting_2020, 23].

In response, some military actors then defended Rumsfeld. And while Rumsfeld's defenders did not necessarily constitute a large number of military actors, they too made public appeals. For instance, four retired generals - Vallely, McInerney, Crosby, and Moore - penned an op-ed in _The Wall Street Journal_ in April of 2006 entitled, "In Defense of Donald Rumsfeld" [@crosby_defense_2006]. 

### Grandstanding

While public appeals were the most obvious type of political behavior involved in the "Revolt of the Generals," the episode also indirectly touched on the behavior of grandstanding. In chapter two, I noted that Risa Brooks defined grandstanding as the "threat or actual resignation in protest of policy" [@brooks_militaries_2009, 219]. 

To my knowledge, no military officer publicly resigned or threatened to resign as part of the "Revolt of the Generals." But there were strong indications that several of the participating generals had chosen to retire from active military service rather than continue to serve, ostensibly in order to publicly criticize the Secretary of Defense. For example, strong rumors circulated that Marine Lieutenant General Newbold had chosen to retire despite being considered a strong candidate for the position of Marine Corps Commandant [@kaplan_revolt_2006]. Similarly, Generals Swannack and Batiste, who had most recently served as Army Division Commanders in combat, retired from positions which are not normally positions from which officers enter retirement.

The entire episode therefore sparked somewhat of a debate among civil-military relations scholars as well as among military officers regarding the behavior of military officer resignation, the conditions under which resignation can and should be considered, and whether it is ever an appropriate act. Chapter two discussed the behavior of resignation in greater detail, and noted that even among scholars, there is a lack of consensus about if and when resignation should be considered a viable option by military officers.^[For example, see @burke_responsible_2009; @dubik_principled_2014; @snider_strategic_2014; and @milburn_breaking_2010 for general arguments in favor of military officers exercising dissent under strict conditions. See also, however, @m_shields_introduction_2017 and @golby_beyond_2015 for scholars who present alternate views.] Furthermore, this ensuing debate over resignation occurring in the wake of the "Revolt of the Generals" pointed back to the example involving General Shinseki and his testimony before Congress regarding troop levels in Iraq. Those who participated in or supported the "Revolt of the Generals" argued that General Shinseki would have been justified to have resigned in protest in 2003 ahead of the Iraq War, rather than acquiesce begrudgingly to a strategic plan that he believed was bound for failure.

Finally, it should be noted that the "Revolt of the Generals" was unique from other civil-military disputes in US history in that it essentially involved a coordinated effort by multiple retired generals to oust the Secretary of Defense. We usually think of the behavior of resignation from the standpoint of military officers resigning, but during the "Revolt," it was military actors insisting that Rumsfeld resign! After naming Rumsfeld explicitly in his public appeal, Batiste, for instance, argued that, "To move forward, we need a leader with the character and skills necessary to lead" [@batiste_case_2006]. Newbold, for his part, stated that, "We need fresh ideas and fresh faces. That means, as a first step, replacing Rumsfeld and many others unwilling to fundamentally change their approach" [@newbold_why_2006]. Eaton also stated, more bluntly, that "Rumsfeld must go" [@eaton_top-down_2006]. Though Rumsfeld did not heed such requests, the fact that numerous military actors explicitly called on their former civilian boss to depart marked the episode as unique from previous civil-military relations disputes in American history.  

### Impact of Undertaken Behaviors

The public appeals undertaken as part of the "Revolt of the Generals" were highly visible to the American public. Especially in a midterm election year, the behaviors undertaken as part of this episode likely contributed to shaping the views of Americans on the war in Iraq, though it is not possible to accurately assess just how important the "Revolt" was in shaping these views. 

Donald Rumsfeld went on to remain at his post in 2006 until one week after the mid-term elections. During these elections, Republicans incurred heavy losses. Republicans lost more than twenty seats in the House and more than five seats in the Senate, resulting in the Democratic Party gaining control of the House of Representatives for the first time since 1994 [@hulse_wave_2006]. In his memoirs, President George W. Bush briefly commented on the "Revolt of the Generals." Bush wrote that, "there was no way I was going to let a group of retired officers bully me into pushing out the Secretary of Defense. It would have looked like a military coup and would have set a disastrous precedent" [@bush_decision_2010, 93]. 

Thus, we should not conclude that the "Revolt of the Generals" served as the immediate cause of Rumsfeld's dismissal. At the same time, it's clear that the behavior of the revolting generals almost certainly contributed to politicians, lawmakers, and the public subsequently assessing the personal leadership of Rumsfeld and the broader war effort in Iraq. In partisan terms, Republicans mostly responded to the "Revolt of the Generals" by backing Rumsfeld, but taking the time to evaluate the broader strategy of the war in Iraq [@rutenberg_general_2006-1]. A few Democrats, on the other hand, explicitly signaled their support for the revolting generals, and called on Rumsfeld to step down [@rutenberg_general_2006-1]. 

\singlespace
## How Levels of Polarization and Prestige Shaped the Revolt of the Generals
\doublespace

Previous analyses of the "Revolt of the Generals" have examined the episode from the perspective of military dissent, the conditions under which dissent may be appropriate, and the behavioral forms that military dissent should and should not take [@cook_revolt_2008; @snider_strategic_2014; @binkley_revisiting_2020]. Each of these are very important questions in their own right, and I do not revisit these questions here. However, these previous analyses of the "Revolt" have not fully explored the role of polarization and military prestige in enabling the "Revolt" to occur in the first place _and_ to unfold in the manner in which it did. The following paragraphs take on this important task. 

As previously noted, the level of polarization in 2006 was relatively low, at least compared with the levels of polarization that would emerge in the years afterward. Furthermore, the root issue addressed during the "Revolt of the Generals" was the war in Iraq, even as the episode certainly addressed this issue through the lens of Rumsfeld's leadership abilities and qualities. This is an important note, as it signifies that the "Revolt" ultimately dealt with topics related to military strategy, the planning and execution of wars, and the proper role or "voice" of military officers in formulating national security policy [@rapp_civil-military_2015]. 

All of these topics are squarely related to the professional knowledge and expertise of senior military officers. This is not to suggest that issues such as the planning and execution of wars are issues that cannot become polarized; indeed, these issues most certainly can and do. But it is worth emphasizing that even most criticisms of the retired generals who participated in the "Revolt" took issue with the _manner_ in which these officers expressed dissent, and not with the _substance_ of their criticism. The planning and execution of wars and military operations are perhaps the most appropriate topic for military actors to address, even if military actors need to exercise great discretion in _how_ they weigh in on these topics. The "Revolt" occurred under conditions of relatively low polarization and high military prestige aligns with one of the results found in chapter three, which is that since 1979, retired military officers have most frequently addressed topics related to warfighting and military operations or foreign policy when authoring opinion commentary (see Figure \@ref(fig:topics-time)), a dynamic that began to change in the late 2010's. 

Furthermore, it is worth noting that the 2003 Iraq War began on what should be considered a relatively bipartisan footing. The Authorization for Use of Military Force Against Iraq Resolution, for instance, passed the House by a vote of 296 to 133, and passed the Senate by a vote of 77-23, in October of 2002 [@hastert_hj_2002]. And while the the connection between the September 11, 2001 attacks and Iraq proved to be weak, public support at the war's outset was nonetheless relatively high (72%) [@pew_research_center_public_2008]. Even after public opinion on the war began to sour, support for those fighting the war - US Troops - remained somewhat high, a key difference between the war in Iraq and the Vietnam War [@saad_republicans_2006].

The relatively low level of polarization that prevailed at the time of the "Revolt" can also be seen in the fact that none of the revolting officers made explicitly partisan statements that supported or denounced either of the major political parties by name. The criticisms levied by the "revolting" generals instead were directed towards Rumsfeld and at his perceived lack of leadership. For all of these reasons - that the root issue of the "Revolt" involved the war in Iraq, that the war in Iraq had began on bipartisan footing, and that the criticisms levied by the revolting generals did not invoke explicitly partisan sentiments - it is correct to interpret the "Revolt of the Generals" as perhaps contributing to or worsening the moderate level of political polarization that existed at the time, but not as occurring because of the level of political polarization.

Yet the level of military prestige was also relatively high in 2006. This helps explain the fact that the revolting officers undertook behaviors that were highly visible, such as the authoring of public appeals in popular newspapers/magazines, as opposed to engaging in less visible forms of political behavior. It also explains why the press was interested in interviewing and subsequently publicizing the remarks of officers who did not write public appeals themselves, as in the case of Generals Riggs, Swannack, and Van Riper. Furthermore, the high level of military prestige explains the fact that politicians noted and reacted to the dissenting officers and their behavior. Even President Bush's short acknowledgement in his memoirs suggests that he felt some modicum of pressure to respond. 

We cannot know for sure how the "Revolt of the Generals" might have looked had the dissenting officers been part of a military institution that was considerably less prestigious than it was at the time, but we can think about it. The theory offered in chapter two suggests that at the dissenting officers might have engaged in less visible forms of behavior. Perhaps the military officers would have pushed back harder against their civilian superiors but in private rather than public venues and settings. Perhaps some of the military officers, instead of leaving active duty, would have remained on active duty out of a desire to influence policy from the inside rather than from the retired officer ranks. 

### The Revolt and the Central Principles of Civil-Military Relations

The "Revolt of the Generals" clearly violated the principle of civilian control of the military. Furthermore, the episode also violated the principle of non-partisanship, but only indirectly, as I explain below. The episode did not violate the principle of non-interference of the military, however.

The "Revolt of the Generals" most clearly violated the principle of civilian control. By levying blatant and public criticism against Secretary of Defense Rumsfeld and by explicitly calling on him to step down, the officers undertaking the revolt applied friction to the Bush Administration to seriously reconsider its policy approach to the war in Iraq. This was, after all, the stated purpose of the "Revolt." Underneath the allegations of Rumsfeld's acerbic personality was a strategy favored by Rumsfeld and other civilian leaders at the Pentagon that was committed to the use of relatively few US Troops in Iraq.  

Thus, those who applauded the "revolting" generals pointed out that even as the participating military actors engaged in behavior that breached a significant principle of civil-military relations, their behavior contributed to the achievement of a greater good, namely, the evaluation and alteration of a failing strategic policy. One commentator who views the "Revolt of the Generals" as a positive episode, Frederick Kagan, notes that while the generals may have breached protocol, their actions nonetheless helped "ensure that the people of a democracy [were] sufficiently well informed to make sound decisions about their leaders in a time of war" [@kagan_let_2006, 15]. 

The "Revolt of the Generals" also indirectly violated the principle of non-partisanship, however. Here, it is important to consider that on one hand, the participating generals did not levy any explicitly partisan remarks, while on the other, their actions likely nonetheless contributed to clear partisan consequences. The closest thing to a partisan statement came from Eaton, who stated in his op-ed that, "Perhaps a proven Democrat like Senator Joe Lieberman could repair fissures that have arisen between both parties and between uniformed men and the Pentagon big shots" [@eaton_top-down_2006]. Eaton's statement is not a ringing endorsement of either political party, however. 

Yet Secretary Rumsfeld was not an apolitical figure, nor his position simply the equivalent to the CEO of a major company. Rumsfeld was the appointed and confirmed Secretary of Defense serving in the second administration of the Republican President, George W. Bush. Furthermore, as mentioned previously, the "Revolt" occurred ahead of the 2006 midterm elections, and at a time when the American people were in the process of evaluating the war in Iraq. In the wake of the "Revolt" and ahead of the 2006 midterms, Congressional candidates took time to reevaluate their positions on Rumsfeld and a number of other broader topics related to the war in Iraq, potential war with Iran, and broader efforts against terrorism [@howell_while_2007, 238-242]. In early 2007, for instance, just weeks after Rumsfeld's dismissal, the Bush Administration announced a new strategy for Iraq involving a substantial surge of military forces. Senator Lindsey Graham, a Republican, noted at the time that, "we never had enough troops to begin with...a month or two ago, we found out the Army is broken, and they [the American people] agreed that Shinseki was right" [Graham, 2007, quoted in @shanker_new_2007]. 

Perhaps it is impossible to accurately measure the influence of the "Revolt of the Generals" on the outcome of the 2006 midterm elections, or on the Bush decision to institute a surge the following year in 2007. But at the very least, the "Revolt" likely impacted both of these (and other) events. Therefore, we should conclude that the "Revolt" indirectly violated the principle of military non-partisanship because it facilitated partisan consequences.

The "Revolt" did not violate the principle of non-interference, however. The military officers who participated in the revolt addressed topics that were in accord with reasonable expectations of military expertise, as noted in the previous section. In other words, the revolting officers did not address topic areas or issues that are far outside of the relevant professional expertise of military officers. 

### Summary of the "Revolt of the Generals"

The 2006 "Revolt of the Generals" was a significant episode in US civil-military relations. Among the high profile civil-military conflicts that have occurred in US history, the "Revolt" was unique in that several recently retired military generals, all of whom had recently served either in Iraq or been directly involved in its planning, called on the Secretary of Defense to step down. The episode also sparked renewed interest in the topic of military officer resignation and the conditions under which resignation is ever appropriate, a topic that remains hotly contested to this day [@m_shields_introduction_2017]. Table \@ref(tab:pol-behav-cs2) summarizes the political behaviors undertaken by military actors as part of the "Revolt of the Generals" in 2006. 

```{r pol-behav-cs2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1_cs2 <- tribble(~"Actor", ~"Tactic of Political Behavior", ~"Date", ~"Instance", ~"Central Principle Violated",
  "Retired Generals Eaton, Swannack, Batiste, Van Riper, Newbold, and Riggs", "Public Appeals", "2006", "These retired officers wrote a number of op-eds and gave interviews that criticized Secretary of Defense Rumsfeld's handling of the Iraq War", "Civilian Control; Non-Partisanship (Indirectly)",
  "Retired Generals Eaton, Swannack, Batiste, Van Riper, Newbold, and Riggs", "Grandstanding", "1868", "The revolting officers explicitly called on Rumsfeld to resign or step down", "Civilian control; Non-partisanship (Indirectly)"
  )
 

text_tbl1_cs2 %>% kable(booktabs=T, caption="Political Behaviors During the Revolt of the Generals, 2006") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="8em") %>% 
  column_spec(3, width="4em") %>% 
  column_spec(4, width="14em") %>% 
  column_spec(5, width="8em")
  
```

The "Revolt of the Generals" violated the principles of civilian control, and to a lesser degree, the principle of non-partisanship. However, the episode did not violate the principle of non-interference. Critics of the "revolting" generals generally take issue not with the substance of the critiques levied by the outspoken officers, but with the manner of behavior undertaken by these officers [@cook_revolt_2008; @binkley_revisiting_2020]. Those who support the "revolting" generals acknowledge the severity of the behaviors employed by the military actors, but argue that the national good was ultimately better served as a result of the episode having occurred [@kagan_let_2006]. That the episode involved military actors undertaking visible political behaviors over an issue that is highly aligned with the professional expertise of military officers yet without engaging in nor provoking a highly partisan narrative reflects the relative levels of military prestige (high) and polarization (low) that existed in 2006. 

\singlespace
## Part 2 - Domestic Unrest and the US Withdrawal from Afghanistan, 2020-2021
\doublespace

Approximately 14 and 15 years after the "Revolt of the Generals," an incredible flurry of political activity involving the military again occurred, this time centering on several tumultuous political events that occurred in 2020 and 2021. These events included a period of domestic civil unrest in 2020, the 2020 Presidential Election and its aftermath, and the US withdrawal from Afghanistan in 2021. 

Before describing the political activities involving the military that occurred as part of these events, it is first necessary to trace a few major political developments that occurred in the 15 years between the "Revolt of the Generals" and 2020 - 2021. For obvious reasons, it is difficult to fully unpack 15 years of US history into a brief space. The next several paragraphs therefore describe several developments that show how the level of polarization rose, and over what types of issues, from 2006 - 2021. These paragraphs should not be considered the definitive guide to polarization in the United States over this period. Scholars continue (and likely will) to unfold the dynamics that led to the election, and the stunning presidency, of Donald Trump [for instance, see @skocpol_upending_2020]. This is in itself a major story that cannot be unpacked in this dissertation. Still, the following paragraphs argue that from 2006 - 2021, many of the political topics that generated substantial political discord in the United States did so because they revealed that policy preferences over these topics invoke significantly different morals, values, and worldviews.    

### Increasing Polarization Between 2006 - 2021

Occurring two and a half years after the "Revolt of the Generals," the election of President Obama in the fall of 2008 was significant for many reasons. During his first term in office, the Democratic President sought and ultimately succeeded at instituting a number of major policy reforms. One of the earliest and most consequential of these policy initiatives was "Obamacare," formally known as the _Affordable Care Act_, which significantly overhauled the nation's healthcare system. President Obama also sought to provide generous economic assistance by overseeing passage of the _American Recovery and Reinvestment Act_ in order to revitalize the national economy after a severe financial crisis in 2008 [@conlan_american_2016]. 

Both of these initiatives were highly controversial, as both touched on questions about the size and scope of the federal government, individual versus government responsibility, and in the case of Obamacare, whether individuals could be coerced to obtain medical insurance or pay a penalty, the so-called "individual mandate" [@rovner_why_2017]. Largely in response to these Obama Administration policy initiatives, the Republican Party shifted further to the political right [@mann_its_2012]. A number of conservative Republicans formed the Tea Party, an ultraconservative caucus within the GOP, and in 2010, Republicans went on to make major gains in the midterm Elections, both in Congress as well as in a number of statehouses and gubernatorial races [@mann_its_2012]. 

These Republican gains across a number of individual states were important, Conlan and Posner (2016) note, insofar as they resulted in the implementation of many of Obama's policies on very different terms, depending on whether a state was "red" or "blue" [@conlan_american_2016]. Conlan and Posner argue that this development, which they term "variable speed federalism," led to increasing polarization as some states (led by Republicans) strongly resisted Obama's policies, while others (led by Democrats) happily enacted them to the fullest extent possible [@conlan_american_2016, 283]. Legislative battles in Congress likewise reflected the escalating division between Republicans and Democrats. Mann and Ornstein (2012), for example, point to the debt ceiling crisis of 2011 as constituting a major political moment when Republicans significantly escalated the hostility of their political tactics, engaging in what they call "hostage taking" in order to obtain their desired outcome [@mann_its_2012, 3-30]. 

Simultaneously, the Obama Presidency, like other presidencies, coincided with a number of significant socio-cultural flashpoints. These were important in that while they offered Americans an opportunity to reconsider a number of their convictions, they also had the tendency to exacerbate many of the underlying tensions and divisions that were already present in the political fabric of the nation. Combined with legislative battles between Republicans and Democrats over significant policy initiatives, these cultural flashpoints further contributed to making the political environment highly charged. 

Several of these flashpoints in particular stand out. The 2012 death of Trayvon Martin, a 17 year old black teenager, and the subsequent acquittal of the defendant, George Zimmerman, birthed the _Black Lives Matter_ (BLM) organization. BLM argues that "Black Lives are systematically and intentionally targeted for demise," and the organization would become more prominent in 2014 after a police officer killed Michael Brown, a black man in Ferguson, MO [@noauthor_herstory_nodate]. BLM gained even greater national prominence in the Summer of 2020 after the killing of George Floyd, as chapter two of this dissertation noted. These events, and the BLM movement more broadly, have caused many Americans to at least consider questions such as whether systemic racism exists, to what degree, and what, if anything, should be done about it. These are important questions, and the views they generate are often highly controversial, at least as of this dissertation's writing in 2022.   

Other flashpoints occurred on the cultural fronts of sexuality, gender, and identity. For instance, in 2015, the Supreme Court ruled in _Obergefell v. Hodges_ that the 14th Amendment of the Constitution guarantees the right of two persons of the same sex to marry [@gilkis_obergefell_2018]. The following year, the Obama Administration ended the prohibition on transgender troops serving in the US military, a prohibition that Trump would later reinstate, and that President Biden would once again end in 2021 [@lopez_obama_2016; @jackson_trumps_2019; @feldscher_biden_2021]. 

In 2021 and into 2022, many debates involve this same set of issues - race, sexuality, and gender - even as the application of these issues involve different contexts. One recent example is a bill passed in Florida in March of 2022 related to what topics school instructors in kindergarten through grade three may (and may not) discuss with their students with respect to the concepts of sexuality and gender identity. Dana Goldstein of _The New York Times_ (2022) fascinatingly captures the moral division over this bill. She notes that supporters of this highly controversial legislation refer to the bill as the "Parental Rights in Education" bill, whereas opponents call it the "Don't Say Gay" bill [@goldstein_opponents_2022]. 

A second contemporary example involves the rights of transgender individuals, and for example, whether athletes who identify as female but who were born as biological males should be allowed to compete in organized women's sports.^[In March 2022, a University of Pennsylvania swimmer, Lia Thomas, has most recently awakened this discussion at the national level. See @henderson_caitlyn_2022 for more information.] One of the foundational debates over this issue and larger debates over transgender and LGBTQ rights more broadly seems to involve the concept of equality and in particular, how the concepts of equality and respect for women should be applied in specific situations. Does a high view of equality and respect for women mean that there should be a protected space for women's athletics, and thus that biological males should be excluded from women's sports, or do these concepts instead mean disposing of the constructs of gender and biological sex altogether [@kurtzleben_political_2021]? 

### Rising Polarization and the Military

The connection between these developments and the US military may at first seem foggy. Do military actors really care about a bill passed in Florida that governs what topics teachers may address with first graders, or who competes in collegiate swimming events?  

The theory offered in chapter two answers both of these questions with a resounding yes. This is because these topics ultimately are associated with a deeper set of values, morals, worldviews, and philosophical commitments that military actors cannot simply discard or ignore. Even if the military is not directly involved in the specific manifestations of these issues, rising levels of polarization imply that it is only a matter of time before the military eventually encounters a situation that is related to these issues.^[One clear example of this is the new Army physical fitness test (the Army Combat Fitness Test, or ACFT). The latest version of the test was just released in April 2022. After several years of trying to implement a gender-free scoring rubric, the Army has reversed course and, for the time being, decided to implement standards for the test that vary by gender and age. This was done presumably because a significant body of evidence gathered over the past several years pointed to the fact that certain groups, including women, were failing the gender-free standards at a higher rate than men. See @hardison_independent_2022 for more details, especially pp. 19-30. My point here is simple: the Army's adoption of a test that has different standards by gender and age is not a morally neutral statement in today's polarizing age.]

If in fact "the public square is nothing more than a battleground of gods, each vying to push the levers of power in its favor," then a highly polarized society signifies that the gods competing against each other stand for a set of values that are highly opposed to each other [@leeman_political_2016, 82]. With this understanding in mind, military actors care a great deal about these types of topics, even if they do not care about every one of these topics to the same degree. As polarization rises, then, military actors will increasingly encounter situations in which they have to choose between adhering to the principles of civil-military relations or obeying the moral values and philosophical commitments that they hold dear. As polarization rises, more military actors will obey the values and morals with which they identify, and in the process, violate the central principles of civil-military relations. 

### Political Behaviors Undertaken, 2020-2021

From June 2020 - September 21, I trace two distinct sets of political behavior, each of which occurred at the front and back of this roughly 16 month-long period. The first set includes the reaction of several high profile military officers, to include retirees as well as active duty officers, to racially fueled civil unrest in the aftermath of the murder of George Floyd, an unarmed black man at the end of May 2020. The second set of behaviors involves the response of civilian and military actors to the US withdrawal from Afghanistan and simultaneous revelations that the Chairman of the Joint Chiefs of Staff had engaged in several questionable phone calls before and after January 6, 2021. 

This date, January 6, 2021, matters significantly, especially in terms of understanding the political behaviors involving the military discussed in this section of the chapter occurring later in 2021. On January 6, 2021, several thousand supporters of President Donald Trump rallied in Washington D.C. to decry the pending certification of the 2020 Presidential Election results occurring the same day. After a protest led by President Trump, somewhere between 2,000 and 2,500 Trump supporters entered the US Capitol, some wielding rudimentary weapons, causing members of Congress to flee from their respective chambers and to shelter in place [@lucas_where_2022]. January 6, 2021 will long be remembered an important date in US history. However, in this dissertation, I do not take up the task of addressing the actions of the military (or lack thereof) on this date itself, mainly because the military's role on that day remains somewhat unclear (for now) [@goodman_crisis_2021].   

Through the lens of the theory offered earlier in this dissertation, the political activities involving the military undertaken in June of 2020 and in September 2021 are not merely two aberrations from a domestic political environment that was otherwise marked by relatively stable adherence to the central principles of civil-military relations. Instead, these two episodes clearly illustrate the tension and instability in civil-military relations that arise as a result of extremely high levels of polarization when the level of military prestige is also relatively high. 

### Civil Unrest in May-June, 2020

On May 25, 2020, Minneapolis police officer Derek Chauvin knelt on the neck of George Floyd, an unarmed black man, resulting in Floyd's death. Video footage of the incident understandably outraged many Americans. Within days, the incident ignited a national level conversation about racism in the US, and quickly led to a number of riots, protests, and a general period of civil unrest across many American cities. By June 8, 2020, at least 19 Americans had been killed in violence directly related to this unrest [@mcevoy_14_2020].

Passions were further aroused after June 1, 2020, when the President, Donald Trump, physically walked from the White House, through adjacent Lafayette Park, and to St. John's Episcopal Church. For a period of several minutes, the President posed for cameras while holding a Bible. Reports immediately surfaced that the President had personally ordered that protesters be dispersed from Lafayette Park in order to facilitate the photo opportunity.  

Some viewed the President's supposed actions as in violation of the right to peacefully protest, a direct affront to the first Amendment of the Constitution. According to public opinion polling conducted in June 2020, most Americans (82%) believed that peaceful protests were an appropriate response to the killing of an unarmed man by police, and most Americans (79%) agreed that the violence that accompanied some of the protests ultimately "undermined" the purpose of the protests [@noauthor_reutersipsos_2020]. But in other ways, public opinion over the protests was clearly split by party affiliation. For example, 62 percent of Democrats "strongly supported" the protests conducted in the aftermath of Floyd's death, compared to just 22 percent of Republicans [@noauthor_reutersipsos_2020]. American opinion was also split on just how pressing racism as an issue was, and what actions, if any, should be taken to address it. For example, one poll found that "Democrats, liberals, and Americans younger than 50" were "far more likely than Republicans, conservatives, and Americans older than 50...to think that Blacks are treated less fairly than whites by the police" [@noauthor_poll_2020]. At the extreme ends, cries of "defund the police" were pitted against cries for "law and order" [@noonan_defund_2021]. 

A year later, an independent investigator determined that Trump had _not_ ordered that protesters be dispersed from Lafayette Park to facilitate the photo opportunity [@montanaro_watchdog_2021]. But it was true that the President _had_ strongly considered the use of active duty military forces to quell the unrest, a potential move that many leaders, including Secretary of Defense Mark Esper and General Mark Milley, strongly disagreed with [@seligman_esper_2020]. Chapters two and three briefly mentioned that military actors made a number of public appeals in the wake of Floyd's death and the domestic unrest that followed. These particular public appeals are collectively described below.

### Public Appeals, 2020

The military actors who made public appeals in June 2020 addressed a number of topics. These included the issues of race and racism, but also the topic of what role, if any, active duty military forces should have in responding to unfolding domestic unrest across the country. In total, at least 19 different retired generals and admirals addressed these topics through a variety of mediums, ranging from high profile interviews with the media, written op-eds, and messages posted on personal twitter accounts [@maurer_generals_2020]. In addition to these retired officers making public appeals, all of the active duty military service chiefs, as well as the Chairman of the Joint Chiefs of Staff, made public comments that clearly reminded members of the military - and signaled to the broader American public - that the military follows and obeys the Constitution first, rather than any individual President [@england_top_2020; @milley_official_2020]. 

The public appeals undertaken by many of the military actors, particularly several of the retired generals, explicitly mentioned or strongly alluded to Donald Trump personally, and in so doing, violated the central principle of civilian control of the military. The public statements made by the active duty service chiefs carefully steered clear of doing so, and thus did not violate the principle of civilian control (in my view). Previous scholarship has pointed out that observers of these public appeals can make reasonable arguments both in support of and in opposition to the behavior undertaken by the retired officers [@brooks_let_2020]. 

### The US Withdrawal from Afghanistan, 2021

Just over a year later, in August 2021, the Taliban in Afghanistan handily and quickly overtook the remnants of the central Afghan government, and the US withdrew all military forces from Afghanistan. This withdrawal of military forces from Afghanistan was both unsurprising and very much a surprise, depending on how one viewed the context in which this development took place.

The collapse of the Afghan Government was unsurprising in the sense that the central government of Afghanistan had been losing territory to a resurgent Taliban for several years [@bloch_look_2021]. On the other hand, the speed at which the Afghan Army and other institutions collapsed in the weeks leading up to the final withdrawal of US forces shocked the world. Images of desperate Afghans clinging to the side of American planes and falling to their deaths as US forces departed Afghanistan were difficult to view [@gannon_after_2021]. The last of US forces withdrew from Afghanistan on August 31, 2021, just days after a suicide attack outside of the Kabul Airport killed dozens of Afghans and 13 US Troops [@shivaram_attack_2021]. In the weeks to follow, members of Congress undertook a series of interviews to understand how the United States could have withdrawn from Afghanistan in what was by all accounts a chaotic and embarrassing manner.

Polarizing domestic politics significantly heightened an already tense political moment in September 2021, however, when a book entitled _Peril_, authored by Bob Woodward and Robert Costa, revealed that General Mark Milley, the Chairman of the Joint Chiefs of Staff, had engaged in a series of questionable telephone calls. The first series of calls occurred with Chinese General Li Zuocheng, the chief of the Joint Staff of the People's Liberation Army, in October 2020 and again in January 2021, and the second involved one call with the US Speaker of the House, Nancy Pelosi, on January 8, 2021 [@woodward_peril_2021, xiii-xxviii]. These allegations were explosive both because of the content of these phone calls, and in a broader sense, because of what these phone calls intimated. 

In his calls with Li, Milley essentially sought to reassure China that the US was not considering attacking them as a means of boosting Trump's political capital and ultimately, propelling Trump to the White House [@woodward_peril_2021]. Milley confirmed this was in fact the intent of these calls. Milley told a Congressional Committee in September 2021 that he had told the Chinese General, in effect, "Hell, I'll call you. But we're not going to attack you. Trust me, we're not going to attack you" [Milley, 2021 quoted in @moore_milley_2021]. Milley's call with Pelosi, which was initiated by Pelosi, involved more direct questions concerning the protocols and procedures of launching a military attack in the direct aftermath of the events of January 6, 2021.

Together, the phone calls described by Woodward and Costa in _Peril_ suggest that Milley was genuinely concerned about Trump's state of mind after January 6th, 2021. Woodward and Costa go so far as to state that Milley "believed it was his job as the senior military officer to think the unthinkable," and to "take any and all necessary precautions" to ensure that Trump could not launch a military strike before January 20, the date of the inauguration of Joe Biden [@woodward_peril_2021, xxv]. The revelation of these phone calls, together, raised immediate questions concerning whether Milley had violated the principle of civilian control of the military or worse, either by sharing information with China, a stated competitor of the United States, and/or by discussing protocols and procedures related to the Commander in Chief with someone other than the Commander in Chief. As I detail in the paragraphs below, numerous civilian and military actors joined together to suggest that Milley had. 

A total of three different sets of actors engaged in political behaviors involving the military during this episode. The first was General Milley himself, whose actions are difficult to classify. In my view, the political behavior listed in Table \@ref(tab:pol-behav) that most closely aligns with Milley's actions is that of a form of selective obedience/compliance. This, I recognize, is a serious charge to make. The second and third sets of actors to engage in political behavior involving the military were civilian actors and retired military officers, respectively. Both sets of actors assailed Milley and the Defense Department, and forcefully called on him to resign, some in response to the withdrawal from Afghanistan, and others because of the phone calls Milley had engaged in, and still others as some combination of the two. Several of these calls for Milley to resign (though perhaps not all) violated the principle of civilian control and non-partisanship.   

### A Form of Selective Obedience/Compliance, 2020-21

While General Milley was never strictly forbidden from speaking with either the Chinese General or Speaker Pelosi, and in this sense, his actions were _not_ a form of selective obedience/compliance with an explicitly issued order, his actions can be viewed from another perspective. This perspective involves considering what Milley's civilian supervisors thought about his actions. On this point, the evidence, at least in retrospect, is clear. The acting Secretary of Defense, Christopher Miller, did _not_ approve of Milley engaging in conversations with General Li, and neither did the President. Upon learning about these developments, Miller stated that he "did not and would not ever authorize" General Milley to engage in the types of phone calls that he participated in with China [@elkind_trumps_2021].

Thus, observers of civil-military relations are left with an interesting question: if military actors engage in behavior that they know for sure violates the intent of their civilian bosses, does such behavior constitute violating the principle of civilian control? To phrase the question perhaps a little more directly, are military officers violating the principle of civilian control if they engage in conduct that they believe their bosses might disapprove of, even if their civilian bosses have not explicitly forbidden such conduct? If we truly conceive of the principle of civilian control of the military as consisting of the full enablement of civilian leaders to fulfill their policy goals, as chapter one of this dissertation argued, then General Milley's conduct at least came _very_ close to violating, and perhaps violated, the principle of civilian control.   

Woodward and Costa use different words to describe the gravity of Milley's behavior. They allege that Milley believed in his heart that given the events of January 6th, 2021, he had to "pull a Schlesinger" [@woodward_peril_2021, xxv]. Schlesinger refers to former Secretary of Defense James Schlesinger, who in the waning days of the Nixon Administration, issued order to senior military officers _not_ "to follow orders that came directly from President Nixon without first checking with Schlesinger" and General George Brown, the Chairman of the Joint Chiefs of Staff [@woodward_peril_2021, xxv-xxvi]. If Milley did the same thing as Schlesinger, it's hard to view his actions in any other way than that he deliberately inserted himself for the purposes of providing additional personal oversight to activities and events for a select period of time. Even to the veteran Washington insider, Bob Woodward, and his junior co-author, Robert Costa, the fact that Milley had engaged with a Chinese General and with Speaker Pelosi to discuss what was discussed intimated that in some respects, Milley had chosen to curtail the principle of civilian control, at least for the tumultuous two week period of time between January 6 and January 20, 2021. 

The critical question was whether such conduct deserved to be praised or denounced. Yet with few exceptions, answering this question depended on what side of the political aisle the observer sat.  

### Calls for Resignation (Grandstanding) 

In the wake of these revelations, civilian actors and retired military officers accused General Milley and the Defense Department of three separate offenses. First, these actors charged Milley and Secretary Austin with strategic ineptitude resulting in the humiliating withdrawal of US forces from Afghanistan. Second, these actors charged Milley in particular of treasonous behavior by revealing the intent of the United States to Chinese forces. Third, military and civilian actors accused Milley of violating the chain of command by speaking with Pelosi about the Commander in Chief's state of mind. Subsets of both groups of actors, civilian and military, explicitly called on Milley and in some cases, Lloyd Austin, the Secretary of Defense, to resign.  

In bruising Congressional testimony held on September 28 and 29, 2021, multiple GOP lawmakers assailed Milley. Senator Tom Cotton, as mentioned in this chapter's introduction, implied that Milley should have resigned if the General believed that the President's decision _not_ to leave 2,500 Troops in Afghanistan would result in strategic mission failure [@singman_milley_2021]. Others blasted Milley for conversing with a Chinese General. Representative Vicky Hartzler, for instance, told Milley that, "I think you articulating that, that you would tell him, you would give him a call, I think is worthy of your resignation" [@hartzler_icymi_2021]. Senator Marco Rubio, as another example, wrote a letter to President Biden, insisting that Milley's actions "threatened to tear apart our nation's longstanding principle of civilian control of the military" [@rubio_rubio_2021]. Several other GOP lawmakers called on Milley to at least be thoroughly investigated, including Congressmen Jim Banks of Indiana, Doug Lamborn of Colorado, Rob Wittman of Virginia, Mike Gallagher of Wisconsin, and Senator Joni Ernst of Iowa [@zempel_pack_2021]. 

Several military actors echoed these calls. Retired General Don Bolduc, for example, stated, "I believe his (Milley's) actions are irresponsible and they fall somewhere between treason and dereliction of duty" [@fox_news_staff_retired_2021]. And more than 180 retired flag officers (generals and admirals) known as the _Flag Officers 4 America_ signed an open letter calling on both General Milley and Secretary of Defense Lloyd Austin to resign [@flag_officers_4_america_retired_2021]. This group previously penned an open letter suggesting that the 2020 Election had been stolen from Donald Trump, further attesting to the high level of polarization that prevailed during 2020 and into 2021 [@bender_disturbing_2021].

### Impact of Undertaken Behaviors

It is important to note up front that at least in some ways, the full impact of the public appeals undertaken by military actors in June 2020, the form of selective compliance exercised by General Milley, and the calls by civilian and military actors for General Milley and Secretary of Defense Austin to resign in September 2021 cannot be fully assessed (yet). This is because these behaviors, and the events which precipitated them, will likely still reverberate in the months and years to come in ways that are not yet fully known. 

However, this does not mean that we cannot and should not think through a couple of possible impacts of these behaviors. One potential impact to think through is whether the public appeals written by retired military officers in June of 2020 swayed the 2020 Presidential Election, or any of the other elections (Congressional, Gubernatorial, etc.) that occurred in 2020. While researchers may never know the definitive answer to this question with statistical precision, the possibility alone is worth thinking through. In part, this is because such a question then raises a subsequent normative question: do citizens _want_ the voice of retired military officers to potentially sway the results of elections, and to influence public opinion more generally?

This has always been an important question, and previous civil-military relations scholars, as this dissertation has noted, have offered both the advantages and disadvantages of military actor public speech [@robinson_dissents_2020; @brooks_let_2020]. But it is worth thinking through the possibility that these advantages and disadvantages of military actor political behavior depend on the level of polarization that is present in society! If the retired military officer commentary published in June of 2020 hypothetically had even a small influence in turning votes away from Donald Trump and towards Joe Biden, can society also live with the possibility that the same behavior - by a different set of military actors - may have had the opposite effect, and turned votes from Biden towards Trump? If the answer to this corollary hypothetical question is "no," then I suggest that for all of the potential positive repercussions of military actor political behavior, Risa Brooks still has the right view when she suggests that we should "require officers to refrain from political activity and that we do so across the board" [@brooks_militaries_2009, 237]. Said differently, if we applaud some political behaviors that violate normative standards and not other behaviors, then the principles of civil-military relations will either no longer remain principles, or they will be replaced by other principles that have yet to be fully derived. 

Another potential impact that is worth thinking about the political behaviors undertaken in 2020 and in 2021 involves what lessons, if any, younger military actors have learned - and will continue to learn - as a result of these episodes. Here, General Milley's behavior is especially worth pondering. As Chairman of the Joint Chiefs of Staff, General Milley undoubtedly found himself in a difficult position in the final months of the Trump Administration, and particularly so after the events of January 6, 2021. Whether one thinks Milley's actions were heroic or treasonous is a discussion that this dissertation, due to high levels of polarization, cannot conclusively solve. Nonetheless, it is vital to point out that many younger military officers are likely to absorb the following lesson from Milley's behavior: obey the principle of civilian control until a situation arises that one cannot do so in good conscience, and at that point in time, follow one's conscience carefully, and be prepared to pay the consequences.

In many ways, this is an excellent lesson to learn. But this excellent lesson also contains an implicit warning. This warning is simply that it is not possible to fully predict what situation may spark an officer to consider deviating from the principle of civilian control, or from any central principle of civil-military relations for that matter. This is in fact the very warning offered by Finer (1962) when he commented on a statement made by General Douglas MacArthur in 1952. A year after being relieved by Truman, MacArthur stated, "I find in existence a new and heretofore unknown and dangerous concept that the members of our armed forces owe primary allegiance or loyalty to those who temporarily exercise the authority of the Executive Branch of Government rather than to the country and its constitution which they are sworn to defend. No proposition could be more dangerous" [MacArthur, 1952 quoted in @finer_man_1962, 22-23]. 

In its context, MacArthur's statement suggests that he believed he had acted in accordance with both the Constitution and the nation's best interest when sparring with Truman over Korean policy. But Finer makes this warning regarding MacArthur's comments: "The moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this, it is only a skip to the constrained substitution of this view for that of the civilian government" [@finer_man_1962, 23]. 

The implications of Finer's warning are both profound and highly relevant to today's political environment in the US. If military leaders such as General Milley find themselves in a position where they have to overlook what an elected leader says in order to choose what they personally think is best for democracy or in accord with the Constitution, that is a likely sign that the democracy and the country itself are in a seriously grave condition. There is no guarantee that a military leader in such circumstances will choose the necessarily _better_ outcome, nor be able to determine what such an outcome is. For this reason, the _Washington Post_ columnist David Ignatius perhaps sums up Milley's behavior rather well as a "paradox: the nation benefited from the actions he took, but they also threaten to set a dangerous precedent...edging close to violating the sacrosanct principle of civilian control of the military" [@ignatius_gen_2021].   

It is also worth pointing out that as significant as the Truman - MacArthur episode in the 1950s was, the political environment then was much less polarized than it is today. In highly polarized environments, the theory offered in this dissertation suggests that military leaders who decide to act contrary to the wishes of their commander in chief will instead obey in accordance with his or her highest moral commitments, whatever these might be. Thus, if future military leaders learn that following Milley's example is proper, Americans have no other recourse than to hope that the military leader who finds himself or herself in such a position will have the individual fortitude to make wise decisions, as Milley perhaps did, and as General Grant did more than 150 years ago. But determining what is wise in the first place in a polarized environment is likely to prove far from easy. That Americans should find their military leaders ever in a position to have to make these types of decisions is neither desirable nor sustainable for an extended period of time.   

\singlespace
### How Levels of Polarization and Prestige Shaped Political Behaviors after Domestic Civil Unrest in 2020 and the US Withdrawal from Afghanistan in 2021
\doublespace

It is vital to underscore how a high level of polarization contributed to military and civilian actors undertaking the political activities they engaged in during the events examined in 2020 - 2021. Through this lens, the death of George Floyd was a flashpoint resulting in national conversations about and reflections upon racism in the United States. Many of the military actors who made public appeals were willing to push the limits of, and in some cases violate, the principles of civil-military relations because they felt very strongly that in considering the use of active duty military forces to suppress domestic unrest, President Trump was considering a course of action that would have suppressed the right of citizens to peacefully protest, in violation of the Constitution. 

Like the "Revolt of the Generals," the political behaviors undertaken in the aftermath of unfolding domestic civil unrest in 2020 and the US withdrawal of Afghanistan were highly visible to the public. The retired officers who commented on the President's handling of domestic unrest did so, in many cases, by engaging high profile newspapers, magazines, and other forums where the public was likely to be interested in what these prestigious retired officers had to say. While it can never be known for sure, some baseline level of military prestige likely facilitated the choice of retired military officers not only to engage in public appeals in the first place, but also to seek the publication or airing of their views through relatively popular mediums and to be successful (their views were published or aired) in doing so. 

Unlike the 2006 "Revolt of the Generals," however, the public appeals made in the midst of unfolding domestic unrest occurred in an extremely polarized political atmosphere. For this reason, it is not necessarily surprising that the public appeals undertaken in June 2020 did not address traditional matters of military strategy, external conflict, and military operations - as the "Revolt of the Generals" did - but rather the highly controversial question of whether the military could and should be used in the non-traditional task of suppressing domestic unrest, and under what conditions. 

June 2020 occurred roughly three and a half years into the Presidency of an extremely divisive and polarizing President, Donald Trump. Importantly, it is also factually wrong to conclude that all of the retired military figures who undertook public appeals in June of 2020 did so for the first time after staying quiet throughout Trump's Presidency. In fact, the data collected and referenced in chapter three of this dissertation proves that this is not the case. Admiral Mike Mullen, General Michael Hayden, and General John Allen had engaged in _multiple_ public appeals before engaging undertaking public appeals in June of 2020. These authors had previously criticized a number of Trump's policies, including his ban of transgender Troops (Mullen), stance on immigration (Mullen and Hayden), and view on climate change (Allen). With this in mind, it is far more accurate to conclude that the events of May and June 2020 proved to be an exclamation point of sorts during a Trump Presidency that had already laid bare a significant degree of polarization in the country.

The political behaviors undertaken by civilian and military actors in 2021 likewise point to the high degree of polarization that prevailed in the country, even after Joe Biden assumed the Presidency. Genuine questions concerning the chaotic military withdrawal from Afghanistan were largely impacted, if not overshadowed entirely, by the simultaneous revelations of General Milley's previous phone calls with a Chinese General and the Speaker of the House, Nancy Pelosi, near or in the wake of the events of January 6, 2021. 

It is likewise important to note that General Milley never denied or refuted the allegations revealed by Bob Woodward's and Robert Costa's book, _Peril._ Rather, Milley sought to articulate the reasons _why_ he tried to decrease tensions with China, and why he had reassured Speaker Pelosi concerning the procedures and protocols regarding the use of nuclear weapons. At the root of the reasons offered by Milley to justify his conduct, however, lay issues that are inexorably polarizing. To staunch Democrats, President Trump's mind was in an altered if not degraded state, especially after January 6, 2021. Some Republicans, on the other hand, may not condone the events of January 6, 2021 entirely, but they nonetheless strongly support many of the policies Trump stood for. In short, how one views the former President - and the policies former President Trump stood for - is likely a strong indicator of how one views the political behaviors undertaken in the summer of 2020 and again in 2021 after the US withdrawal from Afghanistan. 

Recent public opinion polling indicates just how divided Americans were over some of the events that occurred in 2020 and 2021. This further attests to the powerful role of polarization in contributing to the political behaviors undertaken during these years. What is especially important to note is how trust and confidence in the military declined by party affiliation after each of the episodes described in 2020 and 2021. For example, trust in the military declined among Democrats after the protests in June 2020, likely in part because the use of the military "for protest control" is strongly opposed by Democrats (79%) compared to Republicans (17%) [@yougov_economistyougov_2020]. Similarly, after the US withdrawal from Afghanistan, confidence in the military dropped the most among Republicans, ostensibly because it had occurred under the watch of a Democrat President [@hadley_afghanistan_2021].   
  
\singlespace
### Political Behaviors in 2020 and 2021 and the Central Principles of Civil-Military Relations
\doublespace
 
The public appeals undertaken in June of 2020 directly violated the principle of civilian control, and indirectly violated the principle of non-partisanship. The political activities undertaken in 2020 and 2021 directly violated both the principles of civilian control _and_ non-partisanship.   

Collectively, the public appeals levied by military actors amid domestic civil unrest in June 2020 clearly violated the principle of civilian control. By strongly condemning President Trump's consideration of using active duty military forces to quell domestic unrest, the retired military officers undertaking these public appeals likely impacted public opinion against the President, and had the practical effect of making the option of using active military forces less viable for the President. Indeed, the military actors who engaged in these public appeals desired to have precisely this impact.  

However, controversial decisions regarding the use of military force are, at the end of the day, ultimately to be made by the civilian chain of command, and not by retired or active military officers. As Americans, we can (and should) all agree that violating the first Amendment rights of protesters is squarely against the Constitution. But it is not at all easy to agree under what circumstances and situations a particular protest is no longer peaceful, and at what point coercive force - even military force - deserves to be used to protect life and property. As the last chapter described, this has been done before, both during the Reconstruction Era and in several significant labor disputes that occurred later in the 19th Century. Many Americans were understandably and rightly impacted by the death of George Floyd, and viewed the resulting protests differently. Still, there is a process by which civilian and military leaders deliberate and consider policy response options. In June of 2020, for better or for worse, this process was augmented by a flurry of public appeals authored by retired military officers. 

General Milley's behavior in 2021 also violated the principle of civilian control. The main evidence of this assertion stems from the content of General Milley's October 8, 2020 and January 8, 2021 phone calls with Chinese General Li Zuocheng. In my view, the most troubling aspect of this phone call is not what Milley said or did not say, but rather the fact that the civilian chain of command at the time has publicly stated that they did not and would not have approved of what Milley told the Chinese [@elkind_trumps_2021]. Even if it is in the regular duties and responsibilities of the Chairman of the Joint Chiefs of Staff to regularly converse with the heads of other militaries, as Milley has reasonably claimed, one must also reasonably conclude that informing foreign militaries of the intent of the US military is not solely the purview of military leaders [@ignatius_gen_2021]. In other words, if making military plans and implementing these plans is subject to civilian authorities, it also stands to reason that communicating these plans - especially to an adversary - should have some degree of civilian input at the very least. General Milley's behavior, by his own account, ensured that this was not the case. 

However, some of the civilian and military actor calls for Milley's resignation also violated the principles of civil-military relations, including the principles of non-partisanship and civilian control. That "some" of the calls for Milley's resignation violated any principles is important, because it implies that other calls for his resignation did not. Here, the distinction lies between genuine calls for Milley's resignation versus those that were made for partisan showmanship rather than out of a deep normative conviction that Milley should resign. 

Of course, the trouble lies in observing this distinction. This ultimately prevents us from classifying these calls for Milley's resignation into these categories. Yet the tone used by some who called on Milley to resign, or intimated that he or Austin should resign, indicates that such calls were done, at least in substantial part, to obtain a partisan advantage rather than to genuinely determine what happened in the final days of America's involvement in Afghanistan. Congressman Matt Gaetz, for instance, told General Milley directly during the Committee hearing held on 29 September 2021, that "you seem to be very happy failing up over there (the Pentagon). But if we didn't have a President that was so addled, you all would be fired. Because that is what you deserve...you're far more interested in what your perception is, and what people think about you, and insider Washington books, than you care about winning..." [@gaetz_ending_2021]. 

In actuality, Gaetz may have genuinely been disappointed by the outcome of Afghanistan, and perhaps had real questions about Milley's role leading to the withdrawal of US Troops from Afghanistan. But Gaetz's remarks seem to have been largely geared towards firing up his political base rather than investigating the proximate causes that led to the withdrawal of US forces from Afghanistan. If in fact this was Mr. Gaetz's intent - to use the military solely for his own partisan advantage - his remarks violated the principle of military non-partisanship. 

These types of remarks, in a very real sense, also violated the principle of civilian control. This is the case because by primarily attempting to obtain a partisan advantage, the ability of the nation to genuinely study and learn from foreign policy failure - in this case, the US mission in Afghanistan - was likely degraded [@schultz_perils_2017]. Congress has a duty to perform oversight of the military. But exercising and implementing this oversight, a key component of the principle of civilian control, cannot and will not occur if the primary lens through which civilian actors look is one of attempting to obtain a perpetual partisan advantage.

### Summary of Political Behavior Involving the Military Undertaken in 2020-2021

The 15 month period from June 2020 - September 2021 included several significant periods of political activity involving the military. This period included weeks of tumultuous domestic unrest in 2020 and the US withdrawal from Afghanistan in 2021. The latter episode occurred alongside of revelations that the senior military officer in the US, the Chairman of the Joint Chiefs of Staff, had engaged in phone calls with a Chinese General and the Speaker of the House. In both 2020 and in 2021, the US military found itself front and center in much of the political conversation occurring in the United States.  Some were proud of how the military handled itself, and others, as this section of the chapter has noted, were highly alarmed with the actions of senior military leaders. 

Table \@ref(tab:pol-behav-cs2-part2) summarizes the political behaviors undertaken by civilian and military actors from 2020-2021. 

```{r pol-behav-cs2-part2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1_cs2_part2 <- tribble(~"Actor", ~"Tactic of Political Behavior", ~"Date", ~"Instance", ~"Central Principle Violated",
  "Multiple Retired Generals", "Public Appeals", "2020", "Multiple retired Generals and Admirals publicly address former President Trump's consideration of using military forces to quell protests and riots in multiple American cities", "Civilian Control; non-partisanship (indirectly)",
  "General Mark Milley", "Selective Obedience/Compliance (form of)", "2020-2021", "General Milley engages in at least two phone calls with a Chinese General in which he seeks to de-escalate tensions with China; the acting Secretary of Defense later says he would not have approved these calls.", "Civilian control",
  "Several Republican Lawmakers", "Grandstanding", "2021", "Multiple Republican Lawmakers demand that General Milley resign in the wake of allegations that he had spoken with both a Chinese General and the Speaker of the House before and after the 2020 Presidential Election. Some of these calls for Milley to resign seem to have been motivated by a desire to fire up the political base of the lawmakers.", "Non-Partisanship; civilian control",
  "Multiple Retired Generals and Admirals, including the group the Flag Officers 4 America", "Grandstanding", "2021", "Several dozen Retired General and Admirals call on General Milley and Secretary of Defense Austin to resign in the wake of the withdrawal from Afghanistan. The group Flag Officers 4 America also publishes a letter suggesting that the 2020 Election had been stolen from Donald Trump. ", "Non-partisanship",
  )
 

text_tbl1_cs2_part2 %>% kable(booktabs=T, caption="Political Behaviors Involving the Military, 2020-2021") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="8em") %>% 
  column_spec(3, width="4em") %>% 
  column_spec(4, width="14em") %>% 
  column_spec(5, width="8em")
  
```

## Comparison and Contrast of "The Revolt" and Political Activity Undertaken in 2020-2021

Both the "Revolt of the Generals" and the political activities involving the military undertaken during the Summer of 2020 and after the US withdrawal from Afghanistan in 2021 were significant. In both episodes, military and civilian actors engaged in visible political behaviors that violated the principles of civil-military relations. Yet the "Revolt of the Generals" and the political activities involving the military undertaken in 2020 - 2021 differ in important ways as well. 

The main difference between the political environment that prevailed in each episode is that during the "Revolt of the Generals," the level of polarization that prevailed was relatively lower than that which existed during the Summer of 2020 and after the withdrawal from Afghanistan in 2021. According to Table \@ref(tab:poss-environment3), the political activities undertaken amid a higher level of polarization should be marked by a more comprehensive type of politicization, more extreme levels of political activity, and significant potential for civil-military conflict than those undertaken when the level of polarization is lower, even as the level of military prestige remains relatively high. 

### Comprehensive Politicization

In my view, a comparison of the "Revolt of the Generals" in 2006 and the political activities undertaken in 2020 and 2021 reveals that the latter episode was in fact marked by comprehensive politicization, a more extreme level of political activity, and significant (rather than high) potential for civil-military conflict. First, the politicization of the military undertaken in 2020 and 2021 was more comprehensive than the "Revolt of the Generals." This is most clearly seen, perhaps, in the fact that the political activities comprising the "Revolt of the Generals" were voluntary acts committed primarily one type of actor (military), whereas the latter episode of 2020-2021 involved behaviors that were somewhat inescapable and committed by both types of actors.

By use of the term inescapable, I have in mind the fact that the political behaviors committed in 2020-2021 flowed out of a far more polarized domestic setting, whereas the "Revolt of the Generals" did not. To see this, a counterfactual thought experiment might help. We might consider, for instance, what may have resulted if the actors in each of the episodes described in this chapter had chosen to _not_ speak out, act, or otherwise engage in the political activity they undertook? 

Had the revolting generals never written the public appeals they actually authored, would any civilian or military actors accused them of failing to fulfill their duties, or violating their oath of office? We cannot know for sure, but given the relatively mild level of polarization prevalent in 2006, it is reasonable to conclude that such criticism would have been either non-existent or relatively mild. Indeed, many critics of the revolting officers believe that these officers actually should have kept their mouths shut. Through this lens, the "Revolt" truly was a deliberate, intentional, and voluntary episode initiated by the revolting officers themselves. 

On the other hand, had Milley _not_ de-escalated with the Chinese General, or had refused to speak with the Speaker of the House after January 6, 2021, would civilian and military actors accused him of failing to fulfill his responsibilities, or of violating his oath of office? Though it is conjecture only, the theory offered and evidence examined thus far in this dissertation suggest that Milley indeed would have been criticized, but from the opposite political direction than the criticism he endured for actually engaging in these calls. Indeed, just as several Republicans tore into Milley for actually participating in the phone calls, it is likely that several Democrats would have strongly criticized Milley for _not_ engaging in the same phone calls, or for _not_ expressing dissent in some way amid a very turbulent period of time in 2020 or in 2021. For instance, had General Milley not offered an apology for walking with President Trump through Lafayette Square in 2020, it is highly likely he would have encountered significant criticism from Democrats. 

Unlike the revolting generals in 2006, General Milley in 2020 and in 2021, to a certain extent, was in a foreordained lose-lose situation: by engaging in the actions that he did, he was largely venerated by one political side and castigated by the other. This intense level of partisanship points to the comprehensiveness of the politicization that can occur in highly polarized settings and when the military institution is relatively prestigious.  

### Extreme Political Activity

Second, the political behavior undertaken in 2020 and in 2021 involving the military yielded political activity that was more extreme than that undertaken in 2006 during the "Revolt of the Generals." The lens of partisanship provides the clearest evidence for this assertion, rather than the types of behaviors committed. 

As noted earlier in this case study, recall that the criticisms levied by the revolting officers in 2006 mainly addressed the personal leadership deficits of Secretary Rumsfeld rather than explicit partisan concerns. These officers called on Rumsfeld to step down or to resign, but their remarks never explicitly criticized, for instance, President George W. Bush's handling of the war in Iraq, even though Secretary Rumsfeld ultimately worked in the Bush Cabinet. 

During the political activities undertaken by military and civilian actors in 2020 and in 2021, the role of partisanship was far more evident. To begin with, the calls for resignation went the other way - from civilian actors to a military actor, General Milley. However, these civilian actors calling for Milley's resignation came primarily from one political party. Furthermore, in calling for Milley to resign, some voices _harshly_ criticized the President, Joe Biden. This is important especially insofar as at the time of the withdrawal from Afghanistan, President Biden had been in office for less than eight months, and had inherited a war that was nearly 20 years old and overseen by Presidents of both political parties. Congressman Matt Gaetz's remarks, referenced earlier in this chapter (he labeled President Biden, "addled"), clearly testify to this fact [@gaetz_ending_2021]. Others calling for or intimating that Milley resign, such as Senator Tom Cotton, were among those who, the previous year, had taken public positions regarding the use of military force to suppress domestic civil unrest that were aligned with those favored by President Donald Trump.^[Cotton, for instance, had published an op-ed in the _New York Times_ entitled "Send in the Troops," in which he had stated, "This venerable law (_The Insurrection Act_), nearly as old as our republic itself, doesn't amount to 'martial law' or the end of democracy, as some excitable critics...have comically suggested. In fact, the federal government has a constitutional duty to the states to 'protect each of them from domestic violence.'" See @cotton_send_2020 for more details. Cotton's views on the domestic use of force thus became the object of scorn by those retired military officers who spoke out against the use of domestic military force.]

### Significant Potential for Civil-Military Conflict

Finally, the political activity undertaken as part of the "Revolt of the Generals" differs from that undertaken in 2020 and in 2021 in that during the latter, the potential for civil-military conflict was greater. The clearest evidence for this assertion stems from the fact that the latter episode occurred amid the background of what can be considered approaching a constitutional crisis. The "Revolt of the Generals," on the other hand, while serious, did not rise to the level of a constitutional crisis. 

The months leading up to January 6, 2021, and the period of time immediately afterward, constitute at least approaching a domestic crisis. It is difficult to pinpoint exactly when this period began and ended, but a reasonable consideration is that it began sometime around the domestic unrest that began in May of 2020, and ended after President Biden's formal inauguration on January 20, 2021. The "Revolt of the Generals," on the other hand, was certainly not insignificant, but it ultimately involved a handful of retired generals. There are no indications that a significant portion of military officers serving at the time shared in the convictions of these dissenting officers, or that a significant number of active duty military officers were on the verge of resigning as a result of Secretary Rumsfeld's leadership, or engaging in activity that would have threatened the day to day operations of the military. 

On the other hand, during 2020 and 2021, the political environment was (and remains) much more highly charged. While there is no data to suggest a significant portion of military officers or troops were on the verge of resigning, there is some evidence to suggest that the military was, like the country itself, split on many of the unfolding events and issues that occurred during these couple of years. The strongest evidence for this is perhaps General Milley's announcement at the National Defense University, where he apologized for walking to St. John's Church with President Trump in June 2020 [@milley_official_2020]. In addition to apologizing, a deeper reading of Milley's remarks also draws out the fact that he is addressing issues of diversity, racial harmony, and the numerous backgrounds of all who wear the uniform. The point is that Milley seems to be addressing not just the country at large, but the military itself, with the implicit understanding that the events that had transpired in the previous weeks, after the death of George Floyd, had deeply impacted many who serve. 

### Impact on Foreign Policy

One final important way in which the two episodes differ is in their impact on the respective foreign policy problems at the time. The 2006 "Revolt of the Generals" contributed to an important discussion and ultimately, a re-evaluation of the US strategy in Iraq. Though the precise impact of the "Revolt" on policy adjustments cannot be fully measured, this chapter noted that following the 2006 midterm Elections, President George W. Bush replaced Donald Rumsfeld with Robert Gates. In 2007, a lighter military footprint, which had been designed and advocated for by Rumsfeld, was augmented by a substantial increase of US military forces. The "Revolt of the Generals" had, in essence, contributed to an important reevaluation and subsequent adjustment of a major US foreign policy, the war in Iraq. 

The political activity conducted in 2020 and in 2021, on the other hand, did little to advance, sharpen, or improve the status of a tangible American foreign policy issue. To begin with, the political activity undertaken by military actors in 2020 concerned the domestic rather than foreign or external use of the military. Yet even in 2021, when the catalyst for much of the political activity undertaken _was_ a major foreign policy event (the sudden collapse of Afghanistan), the simultaneous revelation of General Milley's phone calls resulted in a refocusing on polarizing domestic politics, at least in the short term. As a result, much of the national conversation in September 2021 steered towards domestic politics, rather than truly ascertaining the reasons for America's 20 year long mission to Afghanistan ending the way that it did. 

## Comparison of Case Studies

This section of the chapter compares and contrasts the case study undertaken in this chapter with that which was explored in the previous chapter. Because the level of military prestige was the main independent variable that changed in the previous chapter, and because polarization was the main independent variable that changed in this chapter, there should be portions of each case study that are similar to the other. 

In this chapter, both episodes involved civilian and/or military actors undertaking fairly public and visible forms of political behavior, to include writing public appeals. This was also true of the first episode explored in the previous case study, when several military officers, including Generals Grant, Sheridan, and Pope undertook several instances of engaging in public appeals (mainly leaked letters) to advocate against the views of President Johnson concerning Reconstruction policy. 

Simultaneous conditions of high polarization and high military prestige marked the American domestic political landscape both in the first episode explored in the previous case study (1865-1868), as well as the latter episode explored in this chapter (2020-2021). From 1865 - 1868, several senior military officers, including the generals named above, opposed a deeply unpopular president, Andrew Johnson, primarily because he held a view of Reconstruction that, in the view of these officers, would squander the very purposes for which the Civil War had been fought [@hyman_santon_1960]. Furthermore, these military officers engaged in a number of visible political activities that violated normative standards of civil-military conduct, even by historical standards. Eventually, they allied with Republicans in Congress while advocating that the Army become _more_ involved in the undertaking of several non-traditional domestic political tasks in the southern states. These tasks were essential, these military leaders believed, in order to facilitate the civil rights of African Americans after the Civil War and to prevent another rebellion similar to the Civil War. 

In 2020 and in 2021, many military actors (retired officers especially) similarly opposed a deeply unpopular President, Donald Trump. In 2020, however, it was the President who considered using military force in a non-traditional manner by contemplating whether the military should be used to suppress civil unrest unfolding in multiple American cities in May and June, 2020. For their part, the retired military officers who engaged in a number of public appeals denouncing the domestic use of military force were advocating for a _lesser_ domestic role for military forces. 

A further similarity to the domestic environments in 1867-1868 and in 2020-2021 is evident in the fact that in both periods, high polarization had contributed to constitutional crises (or closely approaching one) occurring in each era. During the earlier period, the previous chapter noted that Congress eventually passed the _The Tenure of Office Act_, the design of which was to explicitly curtail Johnson's authority as President. Johnson was ultimately impeached in 1868 for violating this act. In 2020-2021, the results of the 2020 Presidential Election were questioned, the US Capitol subsequently attacked, and President Trump impeached. In both instances, highly polarized environments strongly shaped the views and the behaviors of civilian and military actors.

### Conclusion

This chapter did not seek to test outright any of the five hypotheses presented previously in this dissertation. Yet the exploration of the two episodes explored in this chapter, both of which occurred in the post-9/11 environment, has helped further confirm both H4 and H5. H4 states that military and civilian actors engage in relatively more visible and public forms of political behavior when military prestige is high relative to periods of low military prestige. H5 states that the most extreme levels of political activity occur in environments characterized by simultaneously high levels of political polarization and military prestige. 

A high level of military prestige prevailed during both of the episodes explored in this chapter. The level of polarization, on the other hand, increased substantially during the twenty year period from 2001 - 2021. Both the "Revolt of the Generals" and the political activity involving the military undertaken in 2020-2021 constituted serious breaches of several of the central principles of civil-military relations. The main difference between the "Revolt of the Generals" and the political behavior involving the military undertaken in the wake of American civil unrest in 2020 and the US withdrawal from Afghanistan is that during the latter episodes, partisan undertones were far more evident. These partisan undertones, furthermore, were not reflective of simple policy disagreements, but instead of deep division involving a number of policies associated with the Trump Presidency that touched on issues involving one's moral commitments and philosophical bents.  

A comparison of the two episodes examined in this case study show that increasing toxicity in a state's political environment further heightens civil-military tension and the potential for civil-military conflict. This case study suggests that policymakers, practitioners, and scholars should, at a minimum, take seriously the notion that while civil-military disputes such as the "Revolt of the Generals" can erupt when the military is relatively popular, the level of polarization is also an important and perhaps a vital indicator of just how significant civil-military tension can become. 
\pagebreak