
# Chapter Six {-}

On September 11, 2001, terrorists hijacked several commercial airplanes and crashed them into several sites, including the twin towers of the World Trade Center in New York City, the Pentagon in Washington D.C., and a field in Western Pennsylvania.^[Heroic Americans prevented this plane from reaching its intended target, the US Capitol or the White House, in Washington D.C. See @kean_911_2004, 14 for more details.] The evening of the 9/11 attacks, members of Congress gathered on the steps of the US capitol to collectively express solidarity and resolve. After a moment of silence and short remarks by then speaker of the House Dennis Hastert, those who were gathered sang the song, "God Bless America" [@us_congress_members_2001].

Seventeen months later, in February 2003, the Senate Armed Services Committee held a hearing on war plans for an imminent invasion of Iraq, which began the following month. One exchange was particularly noteworthy. General Eric Shinseki, the Army Chief of Staff, told the committee that he thought "something on the order of several hundred thousand..." troops were likely required to stabilize Iraq after the invasion [Shinseki, 2003 quoted in @fallows_karmic_2008]. Shinseki's position, however, ran counter to the views held by the Secretary of Defense, Donald Rumsfeld, and his second in command, Paul Wolfowitz, both of whom believed that a smaller, lighter military footprint was more than adequate for invading and stabilizing Iraq. In the days after the hearing, both Wolfowitz and Rumsfeld publicly criticized Shinseki's remarks. Wolfowitz dismissed Shinseki's estimate entirely, and called it "wildly off the mark" [@schmitt_pentagon_2003]. General Shinseki, for his part, remained largely silent, and retired from the Army shortly afterward.  

Eighteen and a half years later, the United States withdrew from Afghanistan after the resurgent Taliban rapidly displaced the Afghan National Government in the summer of 2021. During a Congressional hearing held on 28 September 2021 about the US withdrawal, Tom Cotton, a veteran and the junior Republican Senator from Arkansas, asked the Chairman of the Joint Chiefs of Staff, General Mark Milley, "why haven't you resigned [@singman_milley_2021]?" 

The Senator's question alluded to the fact that Milley, weeks before the collapse of the Afghan Government, had recommended to President Biden that he maintain approximately 2,500 US Troops in Afghanistan, a recommendation the President did not heed. In response to Cotton's question, Milley stated, "It would be an incredible act of political defiance for a commissioned officer to just resign because my advice is not taken...this country doesn't want generals figuring out what orders we are going to accept and do or not. That's not our job [@singman_milley_2021]." In the same hearing, before ever even discussing the issue of Afghanistan, General Milley addressed allegations that he had inappropriately conversed with both a Chinese general and with the Speaker of the House, Representative Nancy Pelosi, during the transition between President Trump and Biden [@baldor_watch_2021]. These allegations, which are described later in this chapter, provoked strong partisan reactions in the days leading up to the hearing. 

This chapter identifies, compares, and contrasts the patterns of political behavior engaged in by military and civilian actors in the post-9/11 era. From approximately 2001 - 2021, the level of military prestige remained relatively high, but the level of political polarization rose sharply. 

The chapter proceeds in four parts. First, the chapter describes the predictions concerning the patterns of political behavior as a state moves from a domestic environment marked by low polarization and high military prestige to one marked by high polarization and high military prestige. In this section, I substantiate the claim that such a movement approximates the change in the domestic environment that occurred in the United States from 2001 - 2021. Second, this chapter describes several political behaviors involving the military that occurred during the first half of this nearly twenty year period. The episode that I focus on in this section of the chapter is the 2006 so-called "Revolt of the Generals," which involved several retired senior military officers publicly castigating the Secretary of Defense, Donald Rumsfeld. 

The third part of the chapter explores an episode that occurred during the second half of the period. In this section, the chapter explores several political behaviors involving the military undertaken in 2020 and 2021. In particular, I focus on political behaviors undertaken by military actors coinciding with a period of domestic civil unrest in June 2020, and behavior undertaken by military and civilian actors in the wake of the US withdrawal from Afghanistan in 2021. The fourth and final part of this chapter evaluates, compares, and contrasts the political behaviors involving the military undertaken in 2006 and in 2020 - 2021.  

Similar to the previous chapter, the central focus in this case study is on the tenor, characteristics, and underlying reasons for the occurrence of the political behaviors involving the military that are examined. The chief argument made in this chapter is that, in accordance with this dissertation's theoretical predictions, the rise in the level of political polarization from 2001 - 2021 contributed to discernible changes in _how_ civilian and military actors politicized the military.

# Theoretical Predictions for the Post-9/11 Era, 2001-2021

This section describe how political behaviors involving the military are predicted to change as a state's domestic environment shifts from one marked by relatively low polarization and high prestige to one characterized by high polarization and high military prestige. 

## Low Polarization and High Military Prestige

The theory presented in chapter two argued that a state's domestic environment strongly shapes the patterns of political behavior that occur involving the military. When the level of military prestige is high, both civilian and military actors are more likely to believe that the military can influence public opinion relative to when the level of military prestige is low. Furthermore, the theory argued that both types of actors, military and civilian, will engage in behaviors that seek to harness or leverage the prestige of the military in order to appeal to the public when military prestige is high, even if doing so violates one or more of the principles of civil-military relations. 

This theoretical expectation is rooted in the political communication literature, which argues that the public listens to voices who it believes are "knowledgeable" and "trustworthy" [@lupia_democratic_1998, 69-76; see also @zaller_nature_1992, 47]. The theory argued that as the level of military prestige rises, the public increasingly considers the military and its members to possess these traits. 

But the theory presented in chapter two also argued that the level of polarization is also an important factor that shapes political behaviors involving the military. When the level of polarization is low, the theory predicts that there will be relatively little basis for any substantive ideological division between and among military actors. This is because periods of relatively low political polarization indicate that few if any differences in the political space involve issues related to moral values and/or worldviews. This is not to say that military actors are entirely unified, or that there is little chance of conflict between military actors, when polarization is low. Rather, the theory predicts that absent high levels of polarization, there is little if any ideological basis, or _motive_, for military actors to behave in ways that violate the central principles of civil-military relations.

In environments marked by the presence of low polarization and high military prestige, the theory ultimately predicts that either military or civilian actors (or both) may induce politicization of the military, and engage in behaviors that violate the principles of civil-military relations in effort to shape public opinion. The theory even noted that such an environment is far from tranquil, and that there is a high potential for civil-military conflict altogether, as shown in the bottom left quadrant of Table \@ref(tab:poss-environment3). 

```{r poss-environment3, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env_revised2 <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "No politicization; low levels of political activity; low potential for civil-military conflict", 
  "Military (induced) politicization; moderate levels of political activity; moderate potential for civil-military conflict",  
  "High Prestige", "Military and Civilian (induced) politicization; moderate levels of political activity; high potential for civil-military conflict", 
  "Comprehensive politicization; extreme levels of political activity; significant potential for civil-military conflict" 
  )
 

  kable(text_env_revised2, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors and the Conduct of Civil-Military Relations") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2:3, width="18em") 
 
```

Historically, there have been several significant civil-military relations disputes that have occurred under the domestic conditions of low polarization and high military prestige in the United States. For example, disputes between General Douglas MacArthur and President Harry Truman over Korea, General Matthew Ridgeway and President Dwight Eisenhower over Ike's New Deal policies in the early 1950s, and Colin Powell and President Bill Clinton over several post-Cold War crises in the early 1990s, all occurred in domestic environments marked by relatively low polarization and high military prestige. Therefore, we should not conclude that civil-military relations is free of tension when polarization is low and when military prestige is high. 

## High Polarization and High Military Prestige

The theory does predict, however, that as polarization rises, the potential for civil-military conflict rises even more. This prediction stems from the expectation that rising political polarization is associated with increasing ideological divisions that occur, particularly within and among military actors who, like the rest of society, become more polarized as the level of polarization grows. Therefore, holding all else equal, as the level of polarization rises, military actors are more likely to engage in behaviors that violate the principles of civil-military relations.

The theory argued that rising levels of polarization shape behavior because of what polarization indicates. In chapter two, the theory argued that polarization at its root attests to a loss of consensus or agreement concerning the moral values and ideals that hold sway in society [@marsden_twilight_2014; @mann_its_2012; @allen_why_2020]. In highly polarizing times such as the contemporary age in the United States, this dissertation's theory noted that the political conflicts that emerge are not simply political. That is, many contemporary polarizing conflicts do not simply involve questions concerning whether society should enact policy "X" or policy "Y", but rather involve conflicts over the worldviews and moral values that underlie policy "X" or policy "Y". 

In making these claims, chapter two highlighted several polarizing issues in the contemporary United States, including the question of what it means to be an American, whether America should be proud of its history, whether abortion is a right, and whether the construct of gender is important anymore. What these and many other polarizing issues have in common is that they cannot be solved nor even intelligently discussed without an individual referencing or alluding to, whether explicitly or implicitly, some sort of moral worldview commitment. 

The theory argued, furthermore, that the reason the level of polarization impacts the conduct of civil-military relations is because a military actor's commitment to the worldviews or values underlying any number of polarizing issues may in fact be stronger than that same actor's level of commitment to the central principles of civil-military relations. In other words, rising polarization challenges military actors in particular because questions of authority ultimately come into play. 

An officer in the United States military likely feels a very strong commitment to the principles of civilian control, non-partisanship of the military institution, and non-interference. But the same officer likely also has a worldview that relies on moral values or commitments of some kind, even if that officer cannot articulate what that worldview is [@sire_universe_2020]. In highly polarizing environments, military actors more frequently encounter situations in which they have to choose between adhering to the principles of civil-military relations or advancing the values and principles that underlie any number of polarizing issues that are at stake in the political sphere. When confronted with such a decision, especially when the consequences are great and much is at stake, military actors will increasingly side with the values that are associated with the worldview to which they hold dear, and thus, behave in ways that violate the principles of civil-military relations.

As Table \@ref(tab:poss-environment3) notes, simultaneously high levels of polarization and military prestige lead to a domestic environment in which there is comprehensive politicization, extreme levels of political activity involving the military, and significant potential for civil-military conflict. These characteristics are even more severe than those that exist in a domestic environment marked by low polarization and high military prestige. 

## Substantiation of Environmental Characteristics, 2001 - 2021

To substantiate the claim that the level of polarization increased and the level of military prestige generally remained constant (but relatively high) over the period of 2001 - 2021, I rely on measurable quantitative data. This is different from the previous chapter, which relied on mainly qualitative data to substantiate levels of polarization _and_ military prestige after the US Civil War. Figure \@ref(case2-polar) plots the level of polarization in both the US House and the US Senate, as well as the level of trust and confidence in the US military, from 2001 - 2021. Data for polarization uses roll-call voting data on members of Congress over time [@jeffrey_b_lewis_voteview_2020], and data for trust in the military comes from Gallup public opinion data over time regarding the level of public trust and confidence in the military [@noauthor_confidence_2020].  

```{r case2-polar-prep, include=FALSE, warning=FALSE, echo=FALSE}

df_trust_polar_cs2 <- read_csv(here("data","polar-inst-trust-time.csv"))

df_trust_polar_cs2 <- df_trust_polar_cs2 %>% filter(Year >= 2001) %>% select(-Congress) %>% 
  mutate(
    house_polar_dim1 = house_polar_dim1 *100,
    sen_polar_dim1 = sen_polar_dim1 *100,
    ) %>% pivot_longer(Military:sen_polar_dim1, names_to = "value", values_to = "Score") %>% distinct()
```

```{r case2-polar, include=T, warning=FALSE, echo=F, message=F, fig.height=4, fig.width=6, fig.cap = "Public Trust (L) and Polarization (R), 2001-2021" }
df_trust_polar_cs2 %>% ggplot(aes(x=Year, y=Score, color=value)) + geom_line() + geom_point(size=0.7) + labs(x="Year", y="Trust in Institutions") + scale_color_discrete(name="Trust/\nPolarization",
                         breaks=c("Military", "house_polar_dim1", "sen_polar_dim1"),
                         labels=c("Trust in\nMilitary","House\nPolarization", "Senate\nPolarization")) + scale_y_continuous(sec.axis = sec_axis(~., name = "Polarization (Scaled)")) + scale_x_continuous(breaks=seq(2001, 2021, 2)) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE))
```

Figure \@ref(fig:case2-polar) shows that polarization and military prestige have changed differently over the period from 2001-2021. With the exception of a select few years, polarization levels in both the US House and Senate have consistently increased over time. In other words, polarization did not substantially drop from 2001 - 2021 during any point in time. When polarization has dropped (2007, 2009, and 2019 in the House; 2003 in the Senate), these drops have been substantively small, and more importantly, outpaced by the numerous increases that have occurred. 

The level of public trust in the military, on the other hand, has exhibited greater variation between years. For example, a relative spike in public trust in the military occurred in 2003, perhaps reflecting a boost in patriotic fervor as the war in Iraq began the same year. Similarly, noticeable declines in public trust in the military occurred in 2007, perhaps reflecting impatience over mounting US casualties in Iraq, and again in 2021, perhaps reflecting public frustration over the US withdrawal from Afghanistan. Yet even in the presence of this variability, the level of military prestige that has endured for the past twenty years is relatively high, and certainly higher than the level that existed in the late 1970s and early 1980s after the Vietnam War. Since data has been kept, for example, the lowest recorded level of trust in the military occurred in 1981, when only 50% of surveyed Americans said they trusted the military a "great deal or quite a lot" [@noauthor_confidence_2020]. In short, the period 2001 - 2021 reflects a domestic environment in which the level of military prestige remained relatively high, but in which the level of polarization increased. 

Note, however, that Figure \@ref(case2-polar) does not, and indeed cannot, definitively conclude when the shift from a relatively moderate to relatively high level of polarization occurred. What we do see, however, is a fairly noteworthy spike in the level of polarization in the US House (red line) beginning in 2011, and that this spike was accompanied by a smaller yet still detectable spike in polarization in the US Senate (blue line) the same year. Furthermore, we know that this sharp rise in polarization occurred midway through the first term of President Barack Obama, and for reasons that are associated with a number of issues that I discuss later in this chapter. 

From a methodological standpoint, the analysis which follows takes the year 2011 as a sort of natural breaking point. I first describe and analyze the 2006 "Revolt of the Generals" through the lens of the three central principles of civil-military relations. I then do the same with respect to a number of political behaviors undertaken by military and civilian actors involving the military in 2020 and 2021. Each of these episodes occurred well before and after the year 2011.

\singlespace
# Part 1 - The War in Iraq and The Revolt of the Generals, 2004 - 2006
\doublespace

The 2006 "Revolt of the Generals" narrowly refers to the behavior of several recently retired senior military officers who, in the Spring of 2006, publicly criticized the Secretary of Defense, Donald Rumsfeld, and specifically, the way in which Rumsfeld had managed the war in Iraq up until that point. By 2006, the war in Iraq was three years old, and it had become clear that any hopes for a quick and relatively easy war had been dashed. 

In a broader sense, the "Revolt of the Generals" can be viewed as the military's backlash to their perception of not being heard, listened to, or valued by their civilian chain of command within the Department of Defense [@snider_strategic_2014; @binkley_revisiting_2020]. In this sense, the "Revolt" was reminiscent of similar critiques of civilian leadership by military actors, such as H.R. McMaster's _Dereliction of Duty_, which charged the top military brass during Vietnam with failing to adequately push back against the wartime policies of President Lyndon Johnson and Secretary of Defense Robert McNamara [@mcmaster_dereliction_1997]. 

The "Revolt of the Generals" also signified a reaction by several military actors to the way that civilians had disparaged General Shinseki for presenting Congress with his honest views on the force levels required to stabilize Iraq, as noted in this chapter's introduction. Shinseki's last months in uniform had proven turbulent, and Rumsfeld did not attend Shinseki's retirement ceremony, a significant break in protocol that rankled some senior military officers [@shanker_retiring_2003].

The timing of the "Revolt of the Generals" also matters a great deal. Not only was the war in Iraq three years old in the Spring of 2006, but also that same year, Congressional midterm elections were held. The behaviors conducted as part of the "Revolt of the Generals" ultimately elevated the status of the war in Iraq, and in all likelihood, partially contributed to the eventual dismissal of Secretary Rumsfeld by the President, George W. Bush.  

## Political Behaviors Undertaken, 2006

The immediate slate of military actors who participated in the "Revolt of the Generals" include a total of six retired military officers, including Army Major General Paul Eaton, Marine Lieutenant General Greg Newbold, Army Major General John Batiste, Marine Lieutenant General Paul Van Riper, Army Major General John Riggs, and Army Major General Charles Swannack. In chapter three of this dissertation, which examined public opinion commentary authored by retired military actors, I discussed the behavior of retired military officers in broad detail, and so I will not readdress those considerations here. 

But it is important to point out that the six retired generals who participated in the "Revolt of the Generals" were unique in several ways. To begin with, these officers had only recently retired. They were not officers whose military experience was decades removed from current events, but rather ones who had participated themselves in the very subject matter they were addressing by undertaking the "Revolt." Eaton, for example, had been responsible for training the Iraqi military from 2003 - 2004, and both Swannack and Batiste had served as Army Division Commanders on the ground in Iraq [@cloud_more_2006]. The recent retirements of the military actors participating in the "Revolt" is importantly mostly because it signified that at least for several of the other actors, they had chosen to retire and speak out _rather than_ remain in uniform. 

Moreover, the actions of these six retired officers spurred other military actors to similarly engage in political behaviors that addressed the same topics. By extension, the considerable press coverage generated by the "Revolt of the Generals" led to a number of civilian actors commenting on the allegations levied against Rumsfeld, and on the broader war effort underway in Iraq. By considering the behaviors undertaken both as part of the "Revolt" _and_ those undertaken directly in response to it, the reader is better positioned to understand the full impacts of this episode. 

### Public Appeals

The main political behavior undertaken by the six military actors who participated in the "Revolt of the Generals" was public appeals. These took various forms, including publishing opinion articles and giving interviews to the press. For example, in March and April of 2006, Generals Eaton, Batiste, and Newbold penned articles that appeared in _The New York Times_, _The Washington Post_, and _Time_ magazine, respectively [@eaton_top-down_2006; @batiste_case_2006; @newbold_why_2006]. The articles levied multiple criticisms of Rumsfeld, including that he had micromanaged the war effort and failed to heed the advice of senior military leaders during the planning and execution of the war. In conjunction with the publication of these articles, Generals Van Riper, Riggs, and Swannack gave interviews in which they "concurred with the criticisms levied by Eaton, Newbold, and Batiste" [@binkley_revisiting_2020, 23].

In response, some military actors then defended Rumsfeld. And while Rumsfeld's defenders did not necessarily constitute a large number of military actors, they too made public appeals. For instance, four retired generals - Vallely, McInerney, Crosby, and Moore - penned an op-ed in _The Wall Street Journal_ in April of 2006 entitled, "In Defense of Donald Rumsfeld" [@john_crosby_defense_2006]. 

### Grandstanding

While public appeals were the most obvious type of political behavior involved in the "Revolt of the Generals," the episode also indirectly involved the behavior of grandstanding. In chapter two, I noted that Risa Brooks defined grandstanding as the "threat or actual resignation in protest of policy" [@brooks_militaries_2009, 219]. 

To my knowledge, no military officer publicly resigned or threatened to resign as part of the "Revolt of the Generals." But there were strong indications that several of the participating generals had chosen to retire from active military service rather than continue to serve, ostensibly in order to publicly criticize the Secretary of Defense. For example, strong rumors circulated that Marine Lieutenant General Newbold had chosen to retire despite being considered a strong candidate for the position of Marine Corps Commandant [@kaplan_revolt_2006]. Similarly, Generals Swannack and Batiste had most recently served as Army Division Commanders in combat, positions which are not normally positions from which officers enter retirement. 

The entire episode sparked somewhat of a debate among civil-military relations scholars as well as among military officers regarding the political behavior of military officer resignation and the conditions under which such behavior can and should be considered. Chapter two discussed resignation in greater detail, and emphasized that even among scholars, there is a lack of consensus about if and when resignation should be considered a viable option by military officers.^[For example, see @burke_responsible_2009; @dubik_principled_2014; @snider_strategic_2014; and @milburn_breaking_2010 for general arguments in favor of military officers exercising dissent under strict conditions. See also, however, @m_shields_introduction_2017 and @golby_beyond_2015 for scholars who present alternate views.] However, it is important to acknowledge the debate over resignation that occurred in the wake of the "Revolt of the Generals" because the episode pointed back to the example involving General Shinseki and his testimony before Congress regarding troop levels in Iraq. Those who participated in or supported the "Revolt of the Generals" argued that General Shinseki would have been justified to have resigned in protest in 2003 ahead of the Iraq War, rather than acquiesce begrudgingly to a strategic plan that he believed was bound for failure.

Finally, it should be noted that the "Revolt of the Generals" was unique from other civil-military disputes in US history in that it essentially involved a coordinated effort by multiple retired generals to oust the Secretary of Defense. All of the revolting generals called on Rumsfeld to resign or step down from his role! After naming Rumsfeld explicitly, Batiste, for instance, argued that, "To move forward, we need a leader with the character and skills necessary to lead" [@batiste_case_2006]. Newbold, for his part, stated that, "We need fresh ideas and fresh faces. That means, as a first step, replacing Rumsfeld and many others unwilling to fundamentally change their approach" [@newbold_why_2006]. Eaton also stated, more bluntly, that "Rumseld must go" [@eaton_top-down_2006]. Though Rumsfeld did not heed such requests, the fact that numerous military actors explicitly called on their former civilian boss to depart marked the episode as unique from previous civil-military relations disputes in American history.  

### Impact of Undertaken Behaviors

The public appeals undertaken as part of the "Revolt of the Generals" were visible to the American public. Especially in a midterm election year, the behaviors undertaken as part of this episode likely contributed to shaping the views of Americans on the war in Iraq, though it is not possible to accurately assess just how important the "Revolt" was in shaping these views.  

Donald Rumsfeld went on to remain at his post in 2006 until one week after the mid-term elections. During these elections, the Republicans incurred heavy losses. Republicans lost more than twenty seats in the House and more than five seats in the Senate, resulting in the Democratic Party gaining control of the House of Representatives for the first time since 1994 [@hulse_wave_2006].

In his memoirs, President George W. Bush briefly commented on the "Revolt of the Generals." Bush wrote that, "there was no way I was going to let a group of retired officers bully me into pushing out the Secretary of Defense. It would have looked like a military coup and would have set a disastrous precedent" [@bush_decision_2010, 93]. Thus, we should not conclude that the "Revolt of the Generals" served as the immediate cause of Rumsfeld's dismissal. 

At the same time, it's clear that the behavior of the revolting generals nonetheless at least contributed to politicians evaluating and staking claims on the personal leadership of Rumsfeld and on the broader war effort in Iraq. Republicans mostly responded to the "Revolt of the Generals" by backing Rumsfeld, but they also seem to have responded to the incident by evaluating the broader strategy of the war in Iraq [@rutenberg_general_2006-1]. Some Democrats, on the other hand, explicitly signaled their support for the revolting generals, and called on Rumsfeld to step down [@rutenberg_general_2006-1].

## How Levels of Polarization and Prestige Shaped the Revolt of the Generals

Previous analyses of the "Revolt of the Generals" have examined the episode from the perspective of military dissent, the conditions under which dissent may be appropriate, and the behavioral forms that military dissent should and should not take [@cook_revolt_2008; @snider_strategic_2014; @binkley_revisiting_2020]. Each of these are very important questions in their own right, and I do not revisit these questions here. However, previous analyses of the "Revolt" have not fully explored the role of polarization and military prestige in enabling the "Revolt" to occur in the first place _and_ to unfold in the manner in which it did. I take up this important task in the following paragraphs. 

As previously noted, the level of polarization in 2006 was not necessarily low in absolute terms, but it was relatively low compared to the levels that would emerge in the years afterward, and relatively higher than the level that had existed years before. The war in Iraq was the root issue of the "Revolt of the Generals," even if the episode clearly touched on Secretary Rumsfeld's personal leadership qualities and how these impacted the war in Iraq. This is an important note, as it signifies that the "Revolt" dealt with topics and issue areas such as military strategy, the planning and execution of wars, and the proper role or "voice" of military officers in formulating national security policy [@rapp_civil-military_2015]. All of these topics are squarely related to the professional knowledge and expertise of senior military officers. 

This is not to suggest that issues such as the planning and execution of wars are issues that cannot become polarizing; indeed, these issues most certainly can. But it is worth emphasizing that most criticisms of the retired generals who participated in the "Revolt" took issue with the manner in which these officers expressed dissent, and not with the substantive topics of their criticism. The planning and execution of wars and military operations are perhaps the most appropriate topic for military actors to address, even if military actors need to exercise great discretion in _how_ they weigh in on these topics. On a related note, chapter three found that since 1979, retired military officers have most frequently addressed topics related to warfighting and military operations or foreign policy when authoring opinion commentary, as shown in Figure \@ref(fig:topics-time), but that this dynamic changed in the late 2010's. 

Furthermore, the 2003 Iraq War began on what should be considered a relatively bipartisan footing. The Authorization for Use of Military Force Against Iraq Resolution, for instance, passed the House by a vote of 296 to 133, and passed the Senate by a vote of 77-23, in October of 2002 [@hastert_actions_2002]. And while the the connection between the September 11, 2001 attacks and Iraq proved to be weak, public support at the war's outset was nonetheless relatively high (72%) [@pew_research_center_public_2008]. Even after public opinion on the war began to sour, support for those fighting the war - US Troops - remained somewhat high, a key difference between the war in Iraq and the Vietnam War [@saad_republicans_2006]. 

Furthermore, none of the revolting officers made explicitly partisan statements that supported or denounced either of the major political parties by name. The criticism levied by the "revolting" generals instead was personally directed towards Rumsfeld, and at his perceived lack of leadership. For all of these reasons - that the root issue of the "Revolt" involved the war in Iraq, that the war in Iraq had began on bipartisan footing, and that the criticisms levied by the revolting generals did not invoke explicitly partisan sentiments - it is correct to interpret the "Revolt of the Generals" as perhaps contributing to or worsening the moderate level of political polarization that existed at the time, but not as occurring primarily because of the level of political polarization. 

Yet the level of military prestige was also relatively high in 2006. This helps explain the fact that the revolting officers undertook behaviors that were highly visible, such as the authoring of public appeals in popular newspapers/magazines, as opposed to engaging in less visible forms of political behavior. It also explains why the press was interested in interviewing and subsequently publicizing the remarks of officers who did not write public appeals themselves, as in the case of Generals Riggs, Swannack, and Van Riper. Furthermore, the high level of military prestige explains the fact that politicians noted and reacted to the dissenting officers and their behavior. Even President Bush's short acknowledgement in his memoirs suggests that he felt some modicum of pressure and at least considered whether or not to respond. 

We cannot know for sure how the "Revolt of the Generals" might have looked different had the dissenting officers been part of a military that was considerably less prestigious than it was at the time, but we can think about it. The theory offered in chapter two suggests that at a minimum, the dissenting officers might have engaged in less public forms of behavior. Perhaps the military officers would have pushed back harder against their civilian superiors, but in private rather than public venues and settings. Perhaps some of the military officers, instead of leaving active duty, would have remained on active duty out of a desire to influence policy from the inside rather than from the retired officer ranks. 

### The Revolt and the Central Principles of Civil-Military Relations

With respect to the central principles of civil-military relations, the "Revolt of the Generals" clearly violated the principle of civilian control of the military. Furthermore, the episode also violated the principle of non-partisanship, but only indirectly. The episode did not violate the principle of non-interference of the military whatsoever.

The "Revolt of the Generals" most clearly violated the principle of civilian control. In levying blatant and public criticism against Secretary of Defense Rumsfeld, the Bush Administration encountered friction and had to seriously reconsider its policy approach to the war in Iraq. This was, after all, the stated purpose of the "Revolt," and the reality was that Secretary of Rumsfeld was a leader who represented a particular type of strategy in Iraq.  

Those who applaud the "revolting" generals are likely to point out that even as the participating military actors engaged in behavior that breached a significant principle of civil-military relations, their behavior nonetheless contributed to the achievement of a greater good, namely, the evaluation and alteration of a failing strategic policy. One commentator who views the "Revolt of the Generals" as a positive episode, Frederick Kagan, notes that while the generals may have breached protocol, their actions nonetheless helped "ensure that the people of a democracy [were] sufficiently well informed to make sound decisions about their leaders in a time of war" [@kagan_let_2006, 15]. 

The "Revolt of the Generals" also indirectly violated the principle of non-partisanship, however. Here, it is important to consider that on one hand, the participating generals did not levy any explicitly partisan remarks, while on the other, their actions nonetheless resulted in very clear partisan consequences. The closest thing to a partisan statement came from Eaton, who stated in his op-ed that, "Perhaps a proven Democrat like Senator Joe Lieberman could repair fissures that have arisen between both parties and between uniformed men and the Pentagon big shots" [@eaton_top-down_2006]. Even so, Eaton's statement does not constitute a ringing endorsement of either party. Instead, the criticism he levies against Rumsfeld is squarely directed at Rumsfeld's acerbic personality, micromanaging tendencies, and other perceived personal leadership deficits. 

Yet Donald Rumsfeld was not an apolitical figure, nor his position simply the equivalent to the CEO of a major company. Rumsfeld was the appointed and confirmed Secretary of Defense serving in the second term of the Republican President, George W. Bush. Furthermore, as mentioned previously, the "Revolt" occurred ahead of the 2006 midterm elections, and at a time when the American people were in the process of evaluating the war in Iraq. In the wake of the "Revolt" and ahead of the 2006 midterms, Congressional candidates took time to reevaluate their positions on Rumsfeld and a number of other broader topics related to the war in Iraq, potential war with Iran, and broader efforts against terrorism [@howell_while_2007, 238-242]. 

In early 2007, just weeks after Rumsfeld's dismissal, the Bush Administration announced a new strategy for Iraq involving a substantial "surge" of military forces. Senator Lindsey Graham, a Republican, noted at the time that, "we never had enough troops to begin with...a month or two ago, we found out the Army is broken, and they [the American people] agreed that Shinseki was right" [Graham, 2007, quoted in @shanker_new_2007]. Perhaps it is impossible to accurately measure the influence of the "Revolt of the Generals" on the outcome of the 2006 midterm elections, or on the Bush decision to institute a surge the following year in 2007. But at the very least, the "Revolt" likely impacted both of these (and other) consequential decisions. Therefore, we should conclude that the "Revolt" indirectly violated the principle of military non-partisanship.

The "Revolt" did not violate the principle of non-interference, however. The military officers who participated in the revolt addressed topics that were in accord with reasonable expectations of military expertise, as noted in the previous section. Their remarks did not address topic areas or issues that are far outside of the relevant expertise of military officers. 

### Summary

The 2006 "Revolt of the Generals" was a significant episode in US civil-military relations. Among the high profile civil-military conflicts that have occurred in US history, the "Revolt" was unique in that several recently retired military generals, all of whom had served either in Iraq or been directly involved in its planning, called on the Secretary of Defense to step down. Additionally, the episode sparked renewed interest in the topic of military officer resignation and the conditions under which resignation is ever appropriate, a topic that remains hotly contested to this day [@m_shields_introduction_2017].  

The "Revolt of the Generals" violated the principles of civilian control, and to a lesser degree, the principle of non-partisanship. However, the episode did not violate the principle of non-interference. Critics of the "revolting" generals generally take issue not with the substance of the critiques levied by the outspoken officers, but with the manner and forms of behaviors used by these officers [@cook_revolt_2008; @binkley_revisiting_2020]. Supporters of these officers, on the other hand, tend to acknowledge the severity of the behaviors employed by the military actors who undertook the "Revolt", but nonetheless argue that the national good was ultimately better served as a result of the episode having occurred [@kagan_let_2006]. 

\singlespace
# Part 2 - Domestic Unrest and the US Withdrawal from Afghanistan, 2020-2021
\doublespace

Approximately 14 and 15 years after the "Revolt of the Generals," an incredible flurry of political activity involving the military again occurred, this time centering on several tumultuous political events in 2020 and 2021. These events included a period of domestic civil unrest that occurred in the Spring and Summer of 2020, the 2020 Presidential Election and its aftermath, and the US withdrawal from Afghanistan in 2021. Before describing the political activities involving the military that occurred as part of these events, it is first necessary to trace some of the major political developments that occurred in the 15 years between the "Revolt of the Generals" and 2020 - 2021. 

For obvious reasons, it is difficult to fully unpack 15 years of US history into a brief space. The next several paragraphs therefore describe several selective developments that show how the level of polarization rose, and over what types of issues, from 2006 - 2021. Scholars continue (and likely will) to unfold the dynamics that led to the election, and the stunning presidency, of Donald Trump [for instance, see @skocpol_upending_2020]. This is in itself a major story that cannot be unpacked in this dissertation. Still, the following paragraphs argue that from 2006 - 2021, many of the political topics that generated controversy did so in part because they involved a significant difference in morals, values, and worldviews.   

## Increasing Polarization Between 2006 - 2021

The election of President Obama in the fall of 2008 was significant for many reasons. During his first term in office, the Democratic President sought and ultimately succeeded at instituting a number of major policy reforms. One of the earliest and most consequential of these policy initiatives was "Obamacare," formally known as the _Affordable Care Act_, which significantly overhauled the nation's healthcare system. President Obama also sought to provide generous economic assistance by overseeing passage of the _American Recovery and Reinvestment Act_ in order to revitalize the national economy after a severe financial crisis in 2008 [@conlan_american_2016]. 

Both of these initiatives were highly controversial, as both touched on questions about the size and scope of the federal government, individual versus government responsibility, and in the case of Obamacare, whether individuals could be coerced to obtain medical insurance or pay a penalty, the so-called "individual mandate" [@rovner_why_2017]. Largely in response to these Obama Administration policy initiatives, the Republican Party shifted further to the political right [@mann_its_2012]. A number of conservative Republicans formed the Tea Party, an ultraconservative caucus within the GOP, and in 2010, Republicans went on to make major gains in the midterm Elections, both in Congress as well as in a number of statehouses and gubernatorial races [@mann_its_2012]. 

These Republican gains across a number of individual states were important, Conlan and Posner (2016) note, insofar as they resulted in the implementation of many of Obama's policies on very different terms, depending on whether a state was "red" or "blue" [@conlan_american_2016]. Conlan and Posner argue that this development, which they term "variable speed federalism," led to increasing polarization as some states (led by Republicans) strongly resisted Obama's policies, while others (led by Democrats) happily enacted them to the fullest extent possible [@conlan_american_2016, 283]. Legislative battles in Congress also reflected the significant and escalating divisions between Republicans and Democrats. Mann and Ornstein (2012), for example, point to the debt ceiling crisis of 2011 as constituting a major political moment when Republicans significantly escalated the hostility of their political tactics, engaging in what they call "hostage taking" in order to obtain their desired outcome [@mann_its_2012, 3-30]. 

The Obama Presidency, like other presidencies, also coincided with a number of significant socio-cultural flashpoints. These were important in that while they offered Americans an opportunity to reconsider a number of their convictions and thus marked perhaps an important degree of change, they also had the tendency to exacerbate many of the underlying tensions and divisions that were already present in the political fabric of the nation. Combined with numerous legislative battles between Republicans and Democrats, these cultural flashpoints helped turn the political environment highly charged. 

Several of these flashpoints stand out. The 2012 death of Trayvon Martin, a 17 year old black teenager, and the subsequent acquittal of the defendant, George Zimmerman, birthed the _Black Lives Matter_ (BLM) organization. BLM argues that "Black Lives are systematically and intentionally targeted for demise," and the organization would become more prominent in 2014 after a police officer killed Michael Brown, a black man in Ferguson, MO [@noauthor_herstory_nodate]. BLM gained even greater national prominence in the Summer of 2020 after the killing of George Floyd, as chapter two of this dissertation noted. These events, and the BLM movement more broadly, have caused many Americans to at least consider questions such as whether systemic racism exists, to what degree, and what, if anything, should be done about it. These are important questions, and the views they generate are often highly controversial.   

Other flashpoints occurred on the cultural fronts of sexuality, gender, and identity. For instance, in 2015, the Supreme Court ruled in _Obergefell v. Hodges_ that the 14th Amendment of the Constitution guarantees the right of two persons of the same sex to marry [@gilkis_obergefell_2018]. The following year, the Obama Administration ended the prohibition on transgender troops serving in the US military, a prohibition that Trump would later reinstate, and that President Biden would once again end in 2021 [@lopez_obama_2016; @jackson_trumps_2019; @feldscher_biden_2021]. 

In 2022, many debates involve this same set of issues - race, sexuality, and gender - even as the application of these issues involve contemporary contexts. One recent example is a bill passed in Florida in March of 2022 related to what topics school instructors for kindergarten through grade three may (and may not) discuss with their students with respect to sexuality and gender identity. As Dana Goldstein of _The New York Times_ (2022) fascinatingly notes, supporters of this highly controversial legislation refer to the bill as the "Parental Rights in Education" bill, whereas opponents call it the "Don't Say Gay" bill [@goldstein_opponents_2022]. A second contemporary example involves transgender athletes, and whether, for example, athletes who identify as female but who were born as biological males should be allowed to compete in organized women's sports.^[In March 2022, a University of Pennsylvania swimmer, Lia Thomas, has most recently awakened this discussion at the national level. See @henderson_caitlyn_2022 for more information.]

The connection between some of these developments and the US military may at first seem foggy. Do military actors really care about a bill passed in Florida that governs what topics teachers may address with first graders? 

But the theory offered in chapter two of this dissertation argued that it is not necessarily a polarizing or controversial topic itself, but rather the values, morals, and worldviews that underlie them that cause military actors to take notice of these types of issues. This occurs because at its heart, increasing polarization denotes the contestation of values, moral judgments, and worldviews that reign in society. In other words, if "the public square is nothing more than a battleground of gods, each vying to push the levers of power in its favor," then a highly polarized society signifies that the gods competing against each other stand for a set of values that are highly opposed to each other [@leeman_political_2016, 82]. With this understanding in mind, military actors care a great deal about precisely this topic and many similar kinds of topics. As polarization rises, then, military actors will increasingly violate the central principles of civil-military relations in order to advance the values and morals in which they believe. 

## Political Behaviors Undertaken, 2020-2021

From June 2020 - September 21, I trace two distinct episodes, each of which occurred at the front and back of this roughly 16 month-long period. The first episode includes the reaction of several high profile military officers, to include retirees as well as active duty officers, to racially fueled civil unrest in the aftermath of the murder of George Floyd, an unarmed black man at the end of May 2020. The second episode involves the response of civilian and military actors to the US withdrawal from Afghanistan as well as simultaneous revelations that the Chairman of the Joint Chiefs of Staff had engaged in two questionable phone calls before and after January 6, 2021. 

On January 6, 2021, several thousand supporters of President Donald Trump rallied in Washington D.C. to decry the pending certification of the 2020 Presidential Election results, which was happening the same day at the US Capitol. After a protest led by President Trump, somewhere between 2,000 and 2,500 Trump supporters entered the US Capitol, some wielding rudimentary weapons, causing members of Congress to shelter in place [@lucas_where_2022]. For reasons that I do not directly take up in this dissertation, January 6, 2021 will long be remembered an important date in US history. The US military's actions - or lack thereof - continue to be a polarizing topic of conversation, and the narrative surrounding the military's actions on that particular day is still not entirely settled [@goodman_crisis_2021]. For this reason, I do not address the actions of the military directly on January 6, 2021, and focus instead on incidents that occurred before and after this date. 

Through the lens of the theory offered earlier in this dissertation, the political activities involving the military undertaken in June of 2020 and in September 2021 are not merely two aberrations from an environment that was otherwise marked by relatively stable adherence to the central principles of civil-military relations. Instead, these two episodes clearly reflect the instability in civil-military relations that arose as a result of rising levels of polarization. 

## Civil Unrest in June of 2020

On May 25, 2020, Minneapolis police officer Darek Chauvin knelt on the neck of George Floyd, an unarmed black man, resulting in Floyd's death. Video footage of the incident understandably outraged many Americans. Within days, the incident ignited a national level conversation about racism in the US, and quickly led to a number of riots, protests, and a general period of civil unrest across many American cities. By June 8, 2020, at least 19 Americans had been killed in violence directly related to this unrest [@mcevoy_14_2020].

Passions were further aroused after June 1, 2020, when the President, Donald Trump, physically walked from the White House, through adjacent Lafayette Park, and to St. John's Episcopal Church. For a period of several minutes, the President posed for cameras while holding a Bible. Reports immediately surfaced that the President had personally ordered that protesters be dispersed from Lafayette Park in order to facilitate the photo opportunity.  

Some viewed the President's supposed actions as in violation of the right to peacefully protest, a direct affront to the first Amendment of the Constitution. According to public opinion polling conducted in June 2020, most Americans (82%) believed that peaceful protests were an appropriate response to the killing of an unarmed man by police, and most Americans (79%) agreed that the violence that accompanied some of the protests ultimately "undermined" the purpose of the protests [@noauthor_reutersipsos_2020]. But in other ways, public opinion over the protests was clearly split by party affiliation. For example, 62 percent of Democrats "strongly supported" the protests conducted in the aftermath of Floyd's death, compared to just 22 percent of Republicans [@noauthor_reutersipsos_2020]. American opinion was also split on just how pressing racism as an issue was, and what actions, if any, should be taken to address it. For example, one poll found that "Democrats, liberals, and Americans younger than 50" were "far more likely than Republicans, conservatives, and Americans older than 50...to think that Blacks are treated less fairly than whites by the police" [@noauthor_poll_2020]. At the extreme ends, cries of "defund the police" were pitted against cries for "law and order" [@noonan_defund_2021]. 

A year later, an independent investigator determined that Trump had _not_ ordered that protesters be dispersed from Lafayette Park to facilitate the photo opportunity [@montanaro_watchdog_2021]. But it was true that the President _had_ strongly considered the use of active duty military forces to quell the unrest, a potential move that many, including Secretary of Defense Mark Esper and General Mark Milley, disagreed with [@seligman_esper_2020]. Chapters two and three briefly mentioned that military actors made a number of public appeals in the wake of Floyd's death and the domestic unrest that followed. These particular public appeals are collectively described below. 

### Public Appeals, 2020

The military actors who made public appeals in June 2020 engaged a number of topics, including the issues of race and racism, but also the topic of what role, if any, active duty military forces should have in responding to the unfolding domestic unrest. In total, at least 19 different retired generals and admirals addressed these topics through a variety of mediums, ranging from high profile interviews with the media, written op-eds, and messages posted on personal twitter accounts [@maurer_generals_2020]. In addition to these retired officers making public appeals, all of the active duty military service chiefs, as well as the Chairman of the Joint Chiefs of Staff, made public comments that clearly reminded members of the military - and signaled to the broader American public - that the military follows and obeys the Constitution first, rather than any individual President [@england_top_2020; @milley_official_2020]. 

The public appeals undertaken by many of the military actors, particularly several of the retired generals, explicitly mentioned or strongly alluded to Donald Trump personally, and in so doing, violated the central principle of civilian control of the military. The public statements made by the active duty service chiefs carefully steered clear of doing so, and thus did not violate the principle of civilian control. Previous scholarship has pointed out that observers of these public appeals can make reasonable arguments both in support of and in opposition to the behavior undertaken by these retired officers [@brooks_let_2020]. As Brooks and Robinson (2020) note, "there is no 'clear' answer" whether the behavior undertaken by retired military officers is normatively right or wrong [@brooks_let_2020].

## The US Withdrawal from Afghanistan, 2021

The US withdrawal from Afghanistan was both unsurprising and very much a surprise. It was unsurprising in the sense that the central government of Afghanistan had been literally losing territory to a resurgent Taliban for several years [@bloch_look_2021]. On the other hand, the speed at which the Taliban overtook the remnants of the Afghan Army in the weeks leading up to the final withdrawal of US forces shocked the world. Images of desperate Afghans clinging to the side of American planes and falling to their deaths as US forces departed Afghanistan were difficult to watch [@gannon_after_2021]. 

The last of US forces withdrew from Afghanistan on August 31, 2021, just days after a suicide attack outside of the Kabul Airport killed dozens of Afghans and 13 US Troops [@shivaram_attack_2021]. In the weeks to follow, members of Congress undertook a series of interviews to understand how the United States could have withdrawn from Afghanistan in what was by all accounts a chaotic and embarrassing manner.

Polarizing domestic politics significantly heightened an already tense political moment in September 2021, however. A book entitled _Peril_, authored by Bob Woodward and Robert Costa, revealed that General Mark Milley, the Chairman of the Joint Chiefs of Staff, had held separate telephone calls, first with Chinese General Li Zuocheng, the chief of the Joint Staff of the People's Liberation Army, in October 2020 and again in January 2021, and later with the US Speaker of the House, Nancy Pelosi, on January 8, 2021 [@woodward_peril_2021].   

These allegations were explosive both because of the content of the phone calls Milley had participated in, and in a larger sense, because of what these phone calls intimated more broadly. In phone calls with General Li of Chine, Milley sought to reassure China that the United States was not considering attacking them as a means of boosting Trump's political capital and ultimately, propelling Trump to the White House [@woodward_peril_2021]. Milley confirmed the intent of his phone calls with the Chinese General to a Congressional Committee, stating that he had told the General, in effect, "Hell, I'll call you. But we're not going to attack you. Trust me, we're not going to attack you" [Milley, 2021 quoted in @moore_milley_2021].

Furthermore, the phone calls suggest that Milley was genuinely concerned about Trump's state of mind after January 6th, 2021, and that Milley "believed it was his job as the senior military officer to think the unthinkable," and to "take any and all necessary precautions" to ensure that Trump could not launch a military strike before January 20, the date of the inauguration of Joe Biden [@woodward_peril_2021]. The revelation of these phone calls, together, raised immediate questions concerning whether Milley had violated the principle of civilian control of the military, and numerous civilian and military actors joined together to suggest that he had. 

A total of three different sets of actors engaged in political behaviors involving the military during this entire episode. The first was General Milley, whose actions are difficult to classify. In my view, the political behavior listed in Table \@ref(tab:pol-behav) that most closely aligns with Milley's actions is that of a form of selective obedience/compliance. This, I recognize, is a serious charge to make, and one that is not made lightly. The second and third sets of actors to engage in political behavior involving the military were civilian actors and retired military officers, respectively. Both sets of actors assailed Milley and the Defense Department, and forcefully called on him to resign, some in response to the withdrawal from Afghanistan, and others because of the phone calls Milley had engaged in. Several of these phone calls violated the principle of civilian control and military non-partisanship.   

### A Form of Selective Obedience/Compliance, 2020-21

While General Milley was never strictly forbidden from speaking with either the Chinese General or Speaker Pelosi, and in this sense, his actions were _not_ a form of selective obedience or compliance with an explicitly given order, his actions can be explored in another way. One imperfect but still reasonable way of classifying his actions is by considering what Milley's civilian supervisors thought about his actions. On this point, the evidence in retrospect is clear: the acting Secretary of Defense, Christopher Miller, did _not_ approve of Milley engaging in conversations with General Li. In fact, upon learning about these developments, Miller stated that he "did not and would not ever authorize" General Milley to engage in the types of phone calls that he participated in [@elkind_trumps_2021].

Thus, observers of civil-military relations are left with an interesting question: if military actors engage in behavior that they know would for sure land them in hot water with their civilian bosses, does such behavior constitute violating the principle of civilian control? A healthy view of and respect for the principle of civilian control, in my view, suggests that it does.  

Woodward and Costa use different words to essentially describe the same phenomenon. They allege that Milley believed he had to "pull a Schlesinger," referring to former Secretary of Defense James Schlesinger, who in the waning days of the Nixon Administration, issued order to senior military officers _not_ "to follow orders that came directly from President Nixon without first checking with Schlesinger" and General George Brown, the Chairman of the Joint Chiefs of Staff [@woodward_peril_2021]. In short, Milley's phone calls with both the Chinese General and Nancy Pelosi intimated that in some respects, Milley had chosen to curtail the principle of civilian control, at least for the roughly two week period of time between January 6 and January 20, 2021. 

### Calls for Resignation (Grandstanding) 

In the wake of these revelations, civilian actors and retired military officers accused General Milley and the Defense Department of three separate offenses, all of which had emerged simultaneously in August and September 2021. First, these actors charged Milley and Secretary Austin with strategic ineptitude resulting in the humiliating withdrawal of US forces from Afghanistan. Second, these actors charged Milley in particular of treasonous behavior by revealing military intent to Chinese forces. Third, military and civilian actors accused Milley of violating the chain of command by speaking with Pelosi about the Commander in Chief's state of mind. Both sets of actors, civilian and military, explicitly called on Milley and in some cases, Lloyd Austin, the Secretary of Defense, to resign.  

In bruising Congressional testimony held on September 28 and 29, 2021, multiple GOP lawmakers assailed Milley. Senator Tom Cotton, as mentioned in this chapter's introduction, implied that Milley should have resigned if the General believed that the President's decision _not_ to leave 2,500 Troops in Afghanistan would result in strategic mission failure [@singman_milley_2021]. Others blasted Milley for conversing with a Chinese General. Representative Vicky Hartzler, for instance, told Milley that, "I think you articulating that, that you would tell him, you would give him a call, I think is worthy of your resignation" [@hartzler_icymi_2021]. Senator Marco Rubio, as another example, wrote a letter to President Biden, insisting that Milley's actions "threatened to tear apart our nation's longstanding principle of civilian control of the military" [@rubio_rubio_2021]. Several other GOP lawmakers called on Milley to at least be thoroughly investigated, including Congressmen Jim Banks of Indiana, Doug Lamborn of Colorado, Rob Wittman of Virginia, Mike Gallagher of Wisconsin, and Senator Joni Ernst of Iowa [@zempel_pack_2021].   
Several military actors echoed these calls for General Milley to resign. Retired General Don Bolduc, for example,  stated, "I believe his (Milley's) actions are irresponsible and they fall somewhere between treason and dereliction of duty" [@fox_news_staff_retired_2021]. And more than 180 retired flag officers (generals and admirals) known as the _Flag Officers 4 America_ signed an open letter calling on both General Milley and Secretary of Defense Lloyd Austin to resign [@america_retired_2021]. This group previously penned an open letter suggesting that the 2020 Election had been stolen from Donald Trump, further attesting to the high level of polarization that prevailed at the time [@bender_disturbing_2021].

### Impact of Undertaken Behaviors

It is important to note that at least in some ways, the full impact of the public appeals undertaken by military actors in June 2020, the form of selective compliance exercised by General Milley, and the calls by civilian and military actors for General Milley and Secretary of Defense Austin to resign in September 2021 cannot yet be assessed. This is because these behaviors, and the events which precipitated them, are likely too recent. However, this does not mean that we cannot and should not think through a couple of possible impacts of these political behaviors.   

One potential impact to think through is whether the public appeals written by retired military officers in June of 2020 swayed the 2020 Presidential Election, or any of the Congressional elections that occurred in 2020. While researchers may never know the definitive answer to this question, the possibility alone is worth thinking about, because it raises a subsequent normative question: do citizens _want_ the voice of retired military officers to potentially sway the results of elections, or to influence public opinion more generally?

This has always been an important normative question, and previous civil-military relations scholars, as this dissertation has noted, have offered both the advantages and disadvantages of military actor public speech [@robinson_dissents_2020; @brooks_let_2020]. But it is worth thinking through the possibility that these advantages and disadvantages of military actor political behavior depend on the level of polarization that is present in society! In other words, if the retired military officer commentary published in June of 2020 hypothetically had even a small measure of impact in turning votes away from Donald Trump and Joe Biden, can society also live with the possibility that the same behavior - by a different set of military actors - may have had the opposite effect, and turned votes from Biden towards Trump? If the answer to this hypothetical question is "no," then I suggest that for all of the potential positive repercussions of military actor political behavior, Risa Brooks still has the right view when she suggests that we should "require officers to refrain from political activity and that we do so across the board" [@brooks_militaries_2009, 237].

Another potential impact that is worth thinking about involves what lessons, if any, younger military actors have learned - and will continue to learn - as a result of the episodes that described above. Here, General Milley's behavior is especially worth pondering. As Chairman of the Joint Chiefs of Staff, General Milley undoubtedly found himself in a difficult position in the final months of the Trump Administration, and particularly so after the attack on the US Capitol on January 6, 2021. Whether one thinks Milley's actions were heroic or treasonous is a discussion that this dissertation, due to high levels of polarization, cannot conclusively solve. Nonetheless, it is vital to point out that many younger military officers are likely to absorb the following lesson from Milley's behavior: obey the principle of civilian control completely until a situation arises that one cannot do so in good conscience, and at that point in time, follow one's conscience carefully, and be prepared to pay the consequences. 
In many ways, this is an excellent lesson to learn. But this excellent lesson also contains an implicit warning. This warning is simply that is not possible to fully predict what situation may spark an officer to consider deviating from the principle of civilian control, or from any central principle of civil-military relations for that matter. This is in fact the very warning first offered by Finer (1962) when he commented on the behavior of General Douglas MacArthur in the early 1950s. Finer stated, "The moment the military draw this distinction between nation and the government in power, they begin to invent their own private notion of the national interest, and from this, it is only a skip to the constrained substitution of this view for that of the civilian government" [@finer_man_1962, 23]. 

The implications of Finer's statement are both profound and highly relevant to today's political environment in the US. If military leaders such as General Milley find themselves in a position where they have to overlook what an elected leader says in order to choose what they personally think is best for democracy, that is a sign that the democracy itself lies in a seriously grave condition. There is no guarantee that the military leader facing such a condition will choose the necessarily _better_ outcome. More than likely, owing to the strong impact of highly polarized environments, the military leader in such a situation will make a choice in accordance with his or her highest commitments. For this reason, the _Washington Post_ columnist David Ignatius perhaps sums up Milley's behavior rather well as a "paradox: the nation benefited from the actions he took, but they also threaten to set a dangerous precedent...edging close to violating the sacrosanct principle of civilian control of the military" [@ignatius_gen_2021]. 

If future military leaders learn that following Milley's example is proper, Americans have no other recourse than to hope that the military leader who finds himself or herself in such a position will have the individual fortitude to make wise decisions, as Milley perhaps did, and as General Grant did more than 150 years ago. However, that Americans should find their military leaders in a position to have to make these types of decisions is neither desirable nor sustainable. 

\singlespace
## How Levels of Polarization and Prestige Shaped Political Behaviors after Domestic Civil Unrest in 2020 and the US Withdrawal from Afghanistan in 2021
\doublespace

It is vital to underscore how a high level of polarization contributed to military and civilian actors undertaking the political activities they engaged in during the events examined in 2020 - 2021. Through this lens, the death of George Floyd was a flashpoint resulting in nationwide conversations about and reflections upon racism in the United States. Many of the military actors who made public appeals were willing to push the limits of, and in some cases violate, the principles of civil-military relations because they felt very strongly that in considering the use of active duty military forces to suppress domestic unrest, President Trump was considering a course of action that would have suppressed the right of citizens to peacefully protest, in violation of the Constitution. 

Like the "Revolt of the Generals," the political behaviors undertaken in the aftermath of unfolding domestic civil unrest in 2020 and the US withdrawal of Afghanistan were highly visible to the public. The retired officers who commented on the President's handling of domestic unrest did so, in many cases, by engaging high profile newspapers, magazines, and other forums where the public was likely to be interested in what these retired officers had to say. While it can never be known for sure, some baseline level of military prestige likely facilitated the choice of retired military officers not only to engage in public appeals in the first place, but also to seek publication or airing of these appeals through fairly popular mediums. 

Unlike the 2006 "Revolt of the Generals," however, the public appeals made in the midst of unfolding domestic unrest occurred in an extremely polarized political atmosphere. For this reason, it is not necessarily surprising that the public appeals undertaken in June 2020 did not address traditional matters of military strategy, external conflict, and military operations, but rather the highly controversial question of whether the military could and should be used in the non-traditional task of suppressing domestic unrest. 

June 2020 occurred roughly three and a half years into the Presidency of an extremely divisive and polarizing President, Donald Trump. Importantly, it is factually wrong to conclude that all of the retired military figures who undertook public appeals in June of 2020 did so for the first time after staying quiet throughout Trump's presidency. In fact, the data collected and referenced in chapter three of this dissertation proves that this is not the case. Admiral Mike Mullen, General Michael Hayden, and General John Allen had engaged in _multiple_ public appeals before engaging undertaking public appeals in June of 2020. These authors had previously criticized a number of Trump's policies, including his ban of transgender Troops (Mullen), stance on immigration (Mullen and Hayden), and view on climate change (Allen). With this in mind, it is far more accurate to conclude that the events of May and June 2020 proved to be a sort of exclamation point on a Trump presidency that had already laid bare the significant degree of polarization in the country.

The political behaviors undertaken by civilian and military actors in 2021 likewise point to the high degree of polarization that prevailed in the country, even after Joe Biden assumed the Presidency. Genuine questions concerning the chaotic military withdrawal from Afghanistan were largely impacted if not overshadowed by the simultaneous revelations of General Milley's previous phone calls with a Chinese General and the Speaker of the House, Nancy Pelosi, in the wake of the events of January 6, 2021. 

It is important to note, however, that General Milley never denied or refuted the allegations revealed by Bob Woodward's and Robert Costa's book, _Peril._ Rather, Milley sought to articulate the reasons _why_ he tried to decrease tensions with China, and why he had reassured Speaker Pelosi concerning the procedures and protocols for the use of nuclear weapons. At the root of the reasons offered by Milley to justify his conduct, however, lay issues that are inexorably polarizing. To staunch Democrats, President Trump's mind was in an altered if not degraded state, especially after the events of January 6, 2021. Many Republicans, on the other hand, may not condone the events of January 6, 2021 entirely, but they nonetheless strongly support many of the policies Trump stood for. In short, how one views the former President - and the policies former President Trump stood for - is likely a strong indicator of how one views the political behaviors undertaken in the summer of 2020 and again in 2021 after the US withdrawal from Afghanistan. 

Future research can explore this in greater depth, but already, recent public opinion polling has indicated just how divided Americans were over some of the events described in this chapter. What is especially important to note is how trust and confidence in the military declined by party affiliation after each of the episodes referenced in 2020 and 2021. For example, trust in the military declined among Democrats after the protests in June 2020, likely because the use of the military "for protest control" is strongly opposed by Democrats (79%) compared to Republicans (17%) [@yougov_economistyougov_2020]. Similarly, after the US withdrawal from Afghanistan, confidence in the military dropped the most among Republicans [@hadley_afghanistan_2021]. These bifurcated opinions of the military along party affiliation further underscore the high level of polarization that prevailed when the episodes of June 2020 and September 2021 occurred.   
  
\singlespace
### Political Behaviors in 2020 and 2021 and the Central Principles of Civil-Military Relations
\doublespace
 
The public appeals undertaken in June of 2020 directly violated the principle of civilian control, and indirectly violated the principle of non-partisanship. The political activities undertaken in 2020 and 2021 directly violated both the principles of civilian control _and_ non-partisanship.   

Collectively, the public appeals levied by military actors amid domestic civil unrest in June 2020 clearly violated the principle of civilian control. By strongly condemning President Trump's consideration of using active duty military forces to quell domestic unrest, the retired military officers who engaged in public appeals likely impacted public opinion against the President, and had the practical effect of making the option of using active military forces less attractive for the President. The military actors who engaged in these public appeals desired to have this impact.  

In this vein, however, it deserves to be stressed that in the United States, controversial decisions regarding the use of military force are, at the end of the day, ultimately to be made by the civilian chain of command, and not by retired or active military officers. We can all agree that violating the first Amendment rights of protesters is squarely against the Constitution. But it not at all easy to agree under what circumstances and situations a particular protest is no longer peaceful, and at what point coercive force - even military force - deserves to be used to protect life and property. After the death of George Floyd, many Americans were understandably and rightly emotional, and viewed the resulting protests differently. Still, there is a process by which civilian and military leaders deliberate and consider policy options. In June of 2020, for better or for worse, this process was augmented by a flurry of public appeals authored by retired military officers. 

General Milley's behavior in 2021 also violated the principle of civilian control. The main evidence of this assertion stems from the content of General Milley's October 8, 2020 and January 8, 2021 phone calls with Chinese General Li Zuocheng. In my view, the most troubling aspect of this phone call is not what Milley said or did not say, but rather the fact that the civilian chain of command at the time has publicly stated that they did not and would not have approved of what Milley told the Chinese [@elkind_trumps_2021]. Even if it is in the regular duties and responsibilities of the Chairman of the Joint Chiefs of Staff to regularly converse with the heads of other militaries, as Milley has claimed and which seems reasonable, one must conclude that informing foreign militaries of US military plans is not solely the purview of military leaders [@ignatius_gen_2021]. In other words, if making military plans and implementing these plans is subject to civilian authorities, it also stands to reason that communicating these plans - especially to an adversary - should have some degree of civilian input. General Milley's behavior ensured that this was not the case. 

Some of the civilian and military actor calls for Milley's resignation also violated the principles of civil-military relations, including the principles of non-partisanship and civilian control. That "some" of the calls for Milley's resignation violated any principles is important, because it implies that other calls for his resignation did not. Here, the distinction lies between genuine calls for Milley's resignation versus those that outwardly stated the same thing, but were made for partisan showmanship rather than out of a deep normative conviction that Milley should resign. 

The trouble lies in observing this distinction, which ultimately prevents us from classifying calls for Milley's resignation into one of these categories. Yet the tone used by some who called on Milley to resign, or intimated that he or Austin should resign, indicates that these were done, at least in substantial part, to obtain a partisan advantage rather than to genuinely determine what happened in the final days of America's involvement in Afghanistan. Congressman Matt Gaetz, for instance, told General Milley directly during the Committee hearing held on 29 September 2021, that "you seem to be very happy failing up over there (the Pentagon). But if we didn't have a President that was so addled, you all would be fired. Because that is what you deserve...you're far more interested in what your perception is, and what people think about you, and inside Washington books, than you care about winning..." [@gaetz_full_2021]. In actuality, Gaetz may have genuinely been disappointed by the outcome of Afghanistan, and perhaps had real questions about Milley's role leading to the withdrawal from Afghanistan. But these types of remarks made by Gaetz seem to have been primarily geared towards firing up his political base rather than investigate the end of the US mission in Afghanistan. In using the military squarely for partisan advantage, if in fact that was Mr. Gaetz genuine intention, his remarks violated the principle of military non-partisanship. 

Similarly, these types of remarks violated the principle of civilian control. This is the case because by primarily focusing on obtaining a partisan advantage, the ability to genuinely study and learn from foreign policy failure - in this case, the US mission in Afghanistan - likely became degraded [@schultz_perils_2017]. Said differently, Mr. Gaetz's remarks likely served as a roadblock to Congress exercising oversight of the military by discovering why the US withdrawal from Afghanistan occurred so suddenly and quickly. 

### Summary

The 15 month period from June 2020 - September 2021 included several significant periods of political activity involving the military. This period included weeks of tumultuous domestic unrest in June 2020 and the US withdrawal from Afghanistan in 2021. The latter episode occurred alongside of revelations that the senior military officer in the US, the Chairman of the Joint Chiefs of Staff, had engaged in phone calls with a Chinese General and the Speaker of the House. The US military featured largely throughout this period. 

The American political landscape that prevailed in 2020 - 2021 was, in many ways, eerily reminiscent of the political landscape that had existed more than 150 years earlier after the US Civil War, a period that was described in the previous chapter. From 1865 - 1868, several senior military officers, including Generals Grant, Sheridan, and Pope, opposed a deeply unpopular president, Andrew Johnson, primarily because he held a view of Reconstruction that, in the view of these military officers, would squander the very purposes for which the Civil War had been fought in the first place [@hyman_santon_1960]. These military officers engaged in a number of political activities that violated normative standards of civil-military conduct. Essentially, they allied with Republicans in Congress while advocating that the Army become _more_ involved in the undertaking of several non-traditional domestic political tasks. These tasks were essential, these military leaders believed, in order to facilitate the civil rights of African Americans after the Civil War and to prevent another rebellion. 

In 2020, military leaders similarly opposed a deeply unpopular President, Donald Trump. This time, however, it was the President who thought about using military forces in a non-traditional manner by considering whether they should be used to suppress civil unrest unfolding in multiple American cities in May and June, 2020. For their part, the retired military officers who undertook a number of public appeals denouncing the domestic use of military force had advocated for _less_ domestic involvement of military forces. 

High polarization, furthermore, had contributed to constitutional crises occurring in both 1867-1868 and again in 2020-2021. During the former era, President Johnson had defied _The Tenure of Office Act_, which had been intended by Congress to curtail Johnson's presidential power in the first place. Johnson was ultimately impeached in 1868. In 2020-2021, the results of the 2020 Presidential Election were questioned, and the US Capitol subsequently attacked.In both instances, highly polarized environments strongly shaped the views and the behavior of the military actors who challenged the President.

# Comparison and Contrast of Behaviors, 2006 and 2020-2021

Both the "Revolt of the Generals" and the political activities undertaken involving the military during the Summer of 2020 and after the US withdrawal from Afghanistan in 2021 were serious and troubling. Observers and scholars are right to see both episodes as including political behaviors undertaken by civilian and military actors that were in violation of the central principles of civil-military relations.

Yet the 2006 "Revolt of the Generals" and the political activity involving the military undertaken in 2020 - 2021 differ in important ways as well. The main difference between the political environment that prevailed in each episode is that during the "Revolt of the Generals," the level of polarization that prevailed was relatively lower than that which existed during the Summer of 2020 and after the withdrawal from Afghanistan in 2021. According to Table \@ref(tab:poss-environment3), the political activities undertaken amid a higher level of polarization should be marked by a more comprehensive type of politicization, more extreme levels of political activity, and significant potential for civil-military conflict, than those undertaken when polarization is lower, even as the level of military prestige remains high. 

In my view, these three characteristics are seen when comparing the "Revolt" to the political behavior of civilian and military actors undertaken in 2020 and 2021 described in this case study. First, politicization in 2020 and 2021 was more comprehensive in the sense that partisanship proved a much more salient factor in 2020 and 2021 than during the "Revolt of the Generals." As noted in this case study, the criticisms levied by the revolting officers mainly addressed the personal leadership deficits of Secretary Rumsfeld rather than explicitly partisan concerns. These officers called on Rumsfeld to step down or to resign, but their remarks never explicitly criticized, for instance, President George W. Bush's handling of the war in Iraq, even though Secretary Rumsfeld ultimately worked for him. 

During the political activities undertaken by military and civilian actors in 2020 and in 2021, the role of partisanship was far more salient. One way to see this is to contrast the direction and the voices calling for another's resignation in 2021. Indeed, those who called on General Milley to resign in September 2021 included a relatively small number of mostly very conservative Republicans. 

Some of these Republicans, such as Senator Tom Cotton, were among those who, the previous year, had taken public positions regarding the use of military force to suppress domestic civil unrest that were aligned with President Donald Trump.^[Cotton, for instance, had published an op-ed in the _New York Times_ entitled "Send in the Troops," in which he had stated, "This venerable law (_The Insurrection Act_), nearly as old as our republic itself, doesn't amount to 'martial law' or the end of democracy, as some excitable critics...have comically suggested. In fact, the federal government has a constitutional duty to the states to 'protect each of them from domestic violence.'" See @cotton_opinion_2020 for more details. Cotton's views on the domestic use of force thus became the object of scorn by those retired military officers who spoke out against the use of domestic military force.] Unlike the "Revolt of the Generals," several of the Republicans used the withdrawal from Afghanistan to bash President Biden, as Congressman Matt Gaetz remarks during congressional testimony, referenced previously in this chapter, made clear. 

Second, the political activities that were undertaken in 2020 and in 2021 were more extreme than those taken in 2006 as part of the "Revolt of the Generals." This comes through most clearly not in the sense that the types of behaviors undertaken in one episode were necessarily more severe than those undertaken during the other, but rather in a broader sense in that the "Revolt of the Generals" did not occur amid anything resembling a larger constitutional crisis, whereas the political activities undertaken in 2020 and 2021 occurred amid a level of divisiveness that after January 6, 2021, erupted into a genuine crisis. One way of further thinking through this distinction involving a more extreme level of political activity involves thinking through a counterfactual thought experiment. We might consider, for instance, what may have resulted if the actors in each of the episodes described in this chapter had chosen to _not_ speak out, act, or otherwise engage in the political activity that they did? 

In particular, it is worth asking whether the revolting generals had never written the public appeals they actually authored. Would any civilian or military actors accused them of failing to fulfill their duties, or violating their oath of office? We cannot know for sure, but given the relatively mild level of polarization prevalent in 2006, it is reasonable to conclude that such criticism would have been either non-existent or relatively mild. Indeed, many critics of the revolting officers believe that these officers should have kept their mouths shut. 

On the other hand, General Milley's tenure as the Chairman of the Joint Chiefs of Staff in 2020 and 2021 occurred in the midst of a far more polarizing environment. After the events of January 6, 2021, had Milley _not_ de-escalated with the Chinese General, or had refused to speak with the Speaker of the House, would civilian and military actors accused him of failing to fulfill his responsibilities, or of violating his oath of office? 

Though it is a conjecture, the theory and cases examined in this dissertation suggest that Milley would have been criticized, but from the opposite political direction. Just as several Republicans tore into Milley for actually engaging in these phone calls, it is likely that several Democrats would have strongly criticized Milley for _not_ engaging in the same phone calls, or for not expressing dissent in some way. This suggests that in a highly polarizing environment and when the level of military prestige is relatively high, the line between a perceived traitor and hero is very fine, depending on the ideology of the person making this distinction. 

Finally, the domestic environment that prevailed in 2020 and in 2021 contained greater potential for civil-military conflict than the environment that prevailed in 2006. The "Revolt of the Generals" was not insignificant, but it ultimately involved a handful of retired generals. There are no indications that a significant portion of military officers serving at the time shared in the convictions of these dissenting officers, or that a significant portion of active duty military officers was on the verge of resigning. During 2020 and 2021, the political environment was (and remains) much more highly charged. 

While there is no data to suggest a significant portion of military officers or troops were on the verge of resigning, there is some evidence to suggest that the military was, like the country itself, split on many of the unfolding events and issues that occurred during these couple of years. The strongest evidence for this is perhaps General Milley's announcement at the National Defense University, where he apologized for walking to St. John's Church with President Trump in June 2020 [@milley_official_2020]. In addition to apologizing, a deeper reading of Milley's remarks also draws out the fact that he is addressing issues of diversity, racial harmony, and the numerous backgrounds of all who wear the uniform. The point is that Milley seems to be addressing not just the country at large, but the military itself, with the implicit understanding that the events that had transpired in the previous weeks, after the death of George Floyd, had deeply impacted many who serve. 

One final important way in which the two episodes differ is in their impact on the respective foreign policy problem at the time. The 2006 "Revolt of the Generals" contributed to an important discussion and ultimately, a re-evaluation of the US strategy in Iraq. Though the precise impact of the "Revolt" on policy adjustments cannot be fully measured, this chapter noted that following the 2006 midterm Elections, President George W. Bush replaced Donald Rumsfeld with Robert Gates. In 2007, a lighter military footprint, which had been designed and advocated for by Rumsfeld, was augmented by a substantial "surge" of military forces. The "Revolt of the Generals" had, in essence, contributed to an important reevaluation and subsequent adjustment of a major US foreign policy, the war in Iraq. 

The political activity conducted in 2020 and in 2021, on the other hand, did little to advance, sharpen, or improve the status of American foreign policy. To begin with, the political activity undertaken by military actors in 2020 concerned the domestic rather than foreign or external use of military forces. But even in 2021, when the catalyst for much of the political activity undertaken was a major foreign policy event (the sudden withdrawal of US forces from Afghanistan), the simultaneous revelation of General Milley's phone calls resulted in the inclusion of polarizing domestic politics. As a result, much of the national conversation in September 2021 steered towards domestic politics, rather than truly ascertaining the reasons for America's 20 year long mission to Afghanistan ending the way that it did. 

## Conclusion

This case study explored two episodes that occurred in the post 9/11 environment. The "Revolt of the Generals" and the political activity involving the military undertaken in 2020 and 2021 differed most substantially in that the former occurred during an era of relatively low polarization, whereas the latter occurred amid an era of much higher polarization. The "Revolt of the Generals" serves as a civil-military relations episode that mainly centered on questions related to the issue of civilian control of the military. The political activities undertaken in 2020 and 2021 touched on the issue of civilian control, but much more clearly incorporated considerations of moral ideology, values, and worldviews, which is precisely the definition of polarization used throughout this dissertation.




\pagebreak