# Chapter Five {-}

On August 19, 1867, _The New York Times_ published a letter sent to Ulysses S. Grant from General John Pope, who at the time commanded one of five military districts that had recently been established by Congressional action in spite of a veto by President Johnson. In the letter, Pope respectfully but clearly warned Grant of the significant consequences the nation would suffer if the military failed to see the task of Reconstruction through. Pope wrote in part, "Now is the time and this is the opportunity to complete the work so that it will stand. Every dictate of wisdom and patriotism demands this work at our hands. If hastily or partially done, Reconstruction will drag with it a train of evils to this country which can never be remedied"[@pope_reconstruction_1867]. 

That Pope's letter was published (likely leaked by either Grant or Pope, or one of their staff members) in the _New York Times_, and that its contents expressed an urgent plea that ran counter to the President's views, is reflective both of the military's central role throughout the Reconstruction era and the strong opinions held by senior military leaders during what were among the nation's most polarizing political days. Indeed, the military and its leaders were not merely observers, but rather firsthand participants in the events leading to and culminating in the President's impeachment in 1868. After Pope's letter was published, Johnson removed Pope and several other commanders of military districts, and replaced them with military officers who shared his outlook regarding Reconstruction. After being impeached early in 1868, Johnson would remain in office, holding on to the Presidency by a single vote. In the fall of that year, the nation elected the Republican candidate, General Ulysses S. Grant, to the first of two presidential terms.

A little more than a decade after Grant ascended to the Presidency, the military no longer occupied as central a role in the life of the nation. By the late 1870s, the combination of the end of Reconstruction, divisive domestic politics, and a lack of clarity regarding the role of military forces at a time when few if any clear international threats existed had rendered the military, and especially the Army, less critical to the country than it had been during the Civil War and in the immediate years that followed. The military and its senior leaders no longer could influence public opinion to the same degree that they could at the height of their prestige. Thus, civilian and military leaders themselves behaved politically in ways that were less extreme than they had immediately after the end of the Civil War. 

This chapter identifies, compares, and contrasts the patterns of political behavior engaged in by military and civilian actors from the end of the American Civil War in 1865 to roughly 15 years later, in 1880. The central argument made in this chapter is that during the post-US Civil War era, the character and severity of the political behaviors involving the military undertaken by civilian and military actors changed significantly. Early in this period, military and civilian actors both undertook highly visible and severe forms of political behavior that sought to leverage the military's political clout. Towards the end of the period, however, civilian and military leaders reacted to the military's diminishing prestige and its potential to sway public opinion by engaging in fewer and less severe forms of political behavior involving the military. 

The chapter proceeds in five parts. First, I describe how we should think about the post-Civil War era in terms of the central principles of civil-military relations. In chapter one, I derived the three central principles of civil-military relations, and in doing so, noted that the emergence and adoption of these principles occurred predominantly in the post-World War Two era. However, I will argue in this chapter that even in the mid-late 19th Century, these principles existed, albeit in a nascent and under-developed form compared to the mid-late 20th Century. More importantly, I argue in this chapter that high levels of polarization and military prestige nonetheless clearly contributed to civilian and military actors violating known standards of civil-military conduct, and for reasons that clearly accord with the theory laid out in chapter two of this dissertation. 

The second part of the chapter lays out the predictions regarding the patterns of political behavior in moving from a period of high polarization and high military prestige to one marked by high polarization and low military prestige. The third and fourth parts of this chapter comprise the heart of this case study. The third part explores the political behaviors engaged in by civilian and military actors during the earlier portion of this case study, from approximately 1865 - 1868. The fourth does the same for the latter portion of the case, from approximately 1877 - 1880. 

The fifth and final part of this chapter evaluates and compares the differences in the patterns of political behaviors that occurred during these two periods. Empirically, using mainly qualitative evidence, this case study contributes to firming up our understanding of the role of military prestige in shaping political behavior involving the military. Combined with the quantitative results obtained in the previous chapter, this chapter's results demonstrate that military prestige plays an important role in shaping how civilian and military leaders harness the military to shape public opinion.  

# Normative Political Principles and the Military in the Late 19th Century

The three central principles of civil-military relations, which were defined in chapter one as the principles of civilian control, non-partisanship of the military, and the non-interference of the military institution, did not exist qua firmly entrenched principles in the post-Civil War era, or for that matter, in any period before the post-World War Two era. These central principles, I noted in chapter one, derive mainly from post-World War Two scholars like Huntington and his critics, as well as military officers and statesmen, all of whom theorize about how a state's coercive and politically influential armed forces can secure the United States throughout the Cold War era while simultaneously obeying its civilian masters. During the late 19th Century and for a good portion of the early 20th Century, the United States was not a global superpower. This matters insofar as the military's latent political influence was not always as high as it has been throughout the post-World War Two era. Indeed, as this chapter shows, civilian leaders have not always had to contend with a military that possesses significant political capital.    

However, it would be wrong to conclude or think that there were no normative expectations regarding military and civilian leader interaction before World War Two, or during the mid-late 19th Century in particular. It would be far more accurate instead to state that the three central principles existed, but in nascent form. In particular, obeying civilian authority was considered a cardinal virtue for military officers, and not politicizing the military in overt ways was considered the same for civilian leaders well before the start of the Civil War. Both sets of actors also, at least in general terms, recognized that the other actor had a predominant sphere that, while perhaps not sacrosanct, was to be off limits except for the rarest of circumstances. In this way, the principles of civilian control, non-partisanship, and non-interference of the military each operated by the mid-late 19th Century. 

To substantiate these claims, I will rely on a mixture of academic expertise and several firsthand accounts from military actors who are the subject of this chapter. William Skelton, a prominent US military historian, notes that by the middle decades of the 1800s and well before the outbreak of the Civil War, "professional [military] ideology rejected partisanship and discouraged taking sides on purely civilian matters. The professed ideal was an apolitical officer corps and a rigid separation between the civilian and military spheres" [@skelton_american_1992, 283]. As early as the 1830s, and largely the result of the War of 1812, Skelton notes that, "though seldom expressed as a formal theory, officers developed a conception of the army as an apolitical instrument of public policy. As servants of the nation, they should stand aloof from party and sectional strife and avoid taking sides on civilian political issues" [@skelton_american_1992, 283]. Skelton's observations provide strong evidence for the existence of the principles of non-partisanship and non-interference in early form. 

For their part, civilian leaders expected that the military and its officers would obey orders and not obstruct their lawful authority. 19th and early 20th Century Presidents well understood that they were endowed by the Constitution as the Commander in Chief of the Armed Forces, even as the Presidency as an institution was not as powerful as it is in the contemporary era [@white_republican_1958, 20-44]. The conviction among military and civilian actors in the principle of civilian control of the military, even by the middle of the 19th Century, shines through in an elegant passage written by one of General Grant's staff officers, Adam Badeau, in 1887. In the larger passage, which cannot be fully quoted here, Badeau contrasts the differences in personalities between his immediate supervisor, General Ulysses Grant, and Grant's direct civilian boss, the Secretary of War, Edwin Stanton, both of whom are major characters in the earlier portion of this case study. Badeau reflects on their obvious personality differences, namely the fact that while Grant tended to be tender and sympathetic to others, Stanton was cold, hard, and direct. However, Badeau also notes that in spite of these differences, the two were united with respect to the principle of civilian control of the military. Badeau notes in particular that in the very few episodes in which Stanton did not allow Grant to proceed as the General wished, it was because Stanton "believed that he was asserting an important principle - the superiority of the civil over the military arm, the doctrine that even a victorious General-in-Chief is, under our institutions, the subordinate of the executive branch of our government" [@badeau_grant_1887, 82]. Clearly, civilian and military leaders at the time held the conviction that the principle of civilian control was to be respected. 

If the central principles of civil-military relations existed in nascent form, an important question then becomes whether the boundaries for acceptable versus inappropriate behaviors at the time were drawn in the same place as they are today? For example, was it acceptable for military or civilian actors to leak information, or for military officers in the 19th Century to write public opinion articles that blasted their civilian bosses? Similarly, was it acceptable for civilians to engage in behaviors that, like those explored in the previous chapter, used the military to bolster their chances at winning office? Answering this question for every type of political behavior explored in chapter two of this dissertation and listed in Tables \@ref(tab:pol-behav) and \@ref(tab:pol-behav-civ) would require significantly more space and is also, I would argue, not necessary. It is sufficient, rather, to state that the boundaries separating acceptable versus inappropriate types and forms of conduct were drawn in much the same place as they are today, with a few prominent exceptions, and to discuss a few of these in greater detail.

Military leaders expected a degree of professional autonomy free from civilian meddling, and resented being dragged into partisan politics. This sentiment shines through very clearly, for instance, in a private letter written by General Ulysses S. Grant to General William Sherman in which Grant explains his view on having his views publicized and aired in the press. Grant writes Sherman that, "military men had no objection to the publication of their views as expressed upon official matters properly brought before them, but that they did not like expressions of theirs which are calculated to array them on one or other side of antagonistic political parties to be brought before the public. That such a course would make or was calculated to make a whole party array itself in opposition to the officer and would weaken his influence for good" [Grant, 1866 quoted in @badeau_grant_1887,  40-41]. Grant's comments clearly indicate that it was not the act itself of having his views made public that bothered him, but rather the manner in which his views were publicized, and the ultimate effect this had, that was potentially problematic. More than 150 years ago, military leaders understood that the context in which political behaviors occurred often mattered more than some behaviors themselves, a point that this dissertation's theory made in chapter two.

Civilian leaders likewise expected that military leaders would obey their orders and not act in insubordinate ways. For example, when President Andrew Johnson relieved General Philip Sheridan as the Commander of the 5th Military District in August of 1867, the President grew increasingly frustrated with Grant for what the President deemed to be insubordinate behavior. Grant had written the President a private letter in which he objected to the removal of Sheridan, but Grant's letter also stated that the general "should be consulted" as to the details of future Presidential orders [Grant, 1867 as cited in @sioussat_notes_1913, 113]. The day after receiving the letter, Johnson summoned Grant, and the President told the General that his letter had "amounted to a sort of political essay," and that "if every order he gave was to provoke a political essay from the General, it would be impossible for the Executive and the head of the War Department to work together" [Johnson, 1867 as quoted in @sioussat_notes_1913, 113]. The President then told Grant that in the event the President leaked Grant's letter to the press, the President thought that the public would side against Grant, on the grounds that the general's behavior clearly bucked well understood norms of obedience at the time [@sioussat_notes_1913, 113]. Grant then asked to formally withdraw the letter, and Johnson agreed. This episode reveals not only that civilian leaders expected military leaders to fully obey and implement orders, but also that both sets of actors were sensitive to public opinion.  

These two episodes together strongly suggest that behaviors such as leaking information, giving interviews, writing op-eds, and otherwise making public appeals that openly disparaged either civilians or military actors by the other side was not considered an appropriate or acceptable behavior, as both sides sought to sustain their reputation with the public in an era when printed newspapers and other written outlets carried the headlines to America. Furthermore, military officers acting explicitly insubordinate, or selectively complying with orders issued by their civilian bosses, was understood to be normatively wrong. Civilians politicizing the military by thrusting the military into the political spotlight for expressly partisan purposes was likewise understood to be normatively inappropriate. 

The one behavior that deserves a bit of further discussion is that of military officers running for office, and in particular, running for President. Norms regarding this type of behavior in the 19th Century were simply not as developed as they are in the contemporary era. For instance, General Winfield Scott, a hero of the Mexican War, actively campaigned for President ahead of the 1852 election while in uniform, an action that modern day Americans would find reprehensible [@skelton_american_1992, 286-287]. The contemporary and late 20th Century American generals and admirals who have run for office - Dwight Eisenhower, Douglas MacArthur, and Wesley Clark - at least waited until their military service was complete before formally announcing their candidacy for office, even if their political ambitions were suspected sooner. 

In summary, the three central principles of civilian control, the non-partisanship of the military institution, and the non-interference of the military were not as firmly developed in the mid-late 19th Century as they would come to be 100 years later and afterward. In this way, we should acknowledge that the individuals examined in this chapter were not guilty of violating established principles that had been well worn for decades upon decades by both civilian and military actors. Yet even in their nascent form, both sets of actors clearly understood and believed in the underlying spirit behind each of the three central principles, with few exceptions. 

Therefore, in the same way undertaken in the previous two empirical chapters, this case study examines not only what types of political behaviors occurred, but what practical effect they had in terms of challenging the principles of civilian control, non-partisanship, and the principle of non-interference of the military. The reader should not interpret the rest of this chapter as an attempt to cloak 19th Century political behavior in a 20th Century framework, but rather as an evaluation of political behaviors that violated normative standards at the time. As in the previous chapters, it is exceedingly clear that high levels of polarization and military prestige impacted how and why both sets of actors, military and civilian, behaved politically. 

# Theoretical Predictions for the Post US Civil War Era, 1865-1881

This section of the chapter summarizes the predictions that stem from the theory laid out in chapter two of this dissertation. I begin with describing the political behaviors predicted to occur in a domestic environment characterized by high polarization and high prestige, and then describe those predicted to occur in a setting marked by high polarization and lower military prestige. 

## High Polarization and High Military Prestige

According to the theory laid out in this dissertation, significant ideological divisions will exist among military actors when polarization is high. The theory further predicts that high levels of polarization will provide a "motive" for military officers to intervene in politics. This stems from the idea that high levels of polarization point to ideological divisions not just about politics, but about the underlying moral issues and values that are at stake in the political sphere. 

The theory also predicts that as military actors encounter increasingly polarized environments, they increasingly encounter choices and decisions in which they must choose either to adhere to the principles of civil-military relations, or to behave in ways that violate these principles, but are aligned with the moral values and philosophies with which they identify. As polarization worsens, then, a greater number of military actors will choose moral principle over civil-military principle, and thus, increasing polarization leads to military actors increasingly engaging in behaviors that violate one or more of the principles of civil-military relations. This case study will show that even after the Civil War and even among officers who had fought together to preserve the Union, high polarization contributed to these officers adopting very different views on - and thus engaging in different behaviors related to - Reconstruction policies. Simply put, high polarization erodes the political basis for unity among military actors because the underlying moral foundation or source of that unity is no longer sure. Said differently, in periods of high polarization, not all military actors act from the conviction that the highest principles are the civil-military principles of civilian control, non-partisanship, and the non-interference of the military. 

When the level of military prestige is also high, however, the theory predicts that civilian and military actors act from the conviction that military actors can potentially influence public opinion. Chapter two rooted this expectation in the political communication literature, a portion of which argues that the public looks to voices who are considered to be "knowledgeable" and "trustworthy" before making decisions [@lupia_democratic_1998, 69-76; see also @zaller_nature_1992, 47]. It is therefore reasonable, the theory argued, to assume that as the level of military prestige rises, the public will view the military and its leaders as increasingly "knowledgeable" and "trustworthy." 

The sociological literature of professions and occupations states that two components, centrality and skill, shape the overall level of prestige commanded by the members of a profession [@davis_conceptual_1942; @jaco_prestige_1970]. The two previous empirical chapters examined empirical behaviors in the post-Cold War era, when the centrality of the military to the United States is relatively high and constant. Variation in military prestige in these two previous chapters stemmed largely from variation in society's evaluation of the military's skill and performance. In this chapter, the source of the variation in military prestige involves mainly the other component of prestige involving how central or important the military is viewed by the nation of which it is a part. 

Regardless of which component drives variation in military prestige, when prestige is high, the theory predicts that both civilian and military actors will engage in behaviors that are intended to sway or influence public opinion, even when these behaviors violate accepted norms of principles of civil-military relations. Furthermore, the theory predicts that during periods of high military prestige, the behaviors that civilian and military actors commit will be relatively more visible to the public - in an attempt to influence public opinion - than in periods of low military prestige. 

When we integrate simultaneously high polarization and military prestige, the theory predicts that civilian and military actors will engage in what can be considered a comprehensive form of military politicization. This means that civilian and military actors will undertake a rather extreme level of political activity when polarization and military prestige exist at high levels. 

```{r poss-environment2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env_revised <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "No politicization; low levels of political activity; low potential for civil-military conflict", 
  "Military (induced) politicization; moderate levels of political activity; moderate potential for civil-military conflict",  
  "High Prestige", "Military and Civilian (induced) politicization; moderate levels of political activity; high potential for civil-military conflict", 
  "Comprehensive politicization; extreme levels of political activity; significant potential for civil-military conflict" 
  )
 

  kable(text_env_revised, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2:3, width="18em") 

```

## High Polarization and Low(er) Military Prestige

If the level of polarization remains high but the level of military prestige declines, the most significant theoretical prediction stemming from this dissertation's theory is that the "opportunity" for civilian and military actors to involve the military in politics diminishes. Declining military prestige is associated with a reduction in the perceived potential of the military to influence public opinion, and therefore, the theory predicts that civilian and military actors will become less likely to undertake political behaviors that involve the military in politics in highly public ways when military prestige is low relative to when it is high.

Nonetheless, when polarization remains high, the theory predicts that military actors maintain a "motive" to intervene in politics. In a domestic environment categorized by "motive" without "opportunity," the theory predicts that military actors will be the primary type of actor inducing political activity, and that they will engage in less severe forms of political behavior than when military prestige is high. Moreover, in an environment marked by high polarization and low military prestige, the theory predicts that civilian actors will engage in far fewer (and far less overt) political behaviors involving the military.

Having laid out these theoretical predictions, this chapter now turns to examining the political behaviors undertaken by civilian and military actors in the years immediately following the end of the Civil War. 

# Part 1 - Johnson, Grant, Congress, and the Army, 1865 - 1868

On numerous occasions from 1865-1868, civilian actors sought to leverage the military's prestige for their own partisan gain. President Andrew Johnson as well as his opponents both attempted to pull Grant and other highly prestigious military actors into their respective political camps, and to portray to the American public that certain military leaders were indeed their political allies. 
At the same time, military leaders, including Grant himself, engaged in several instances of political behavior that, even by historical standards, challenged, if not violated, the normative standards of civil-military conduct that existed at the time. These military leaders did so, I argue, because of deep moral convictions related to significant political issues that arose during the period of Reconstruction. Senior military leaders, many of whom were considered by the public as heroes of the Civil War, were also keenly aware of the enormous influence they commanded with the public at the time. 

In this front section of the case study, I first set the scene by establishing that the levels of polarization and military prestige were both extremely high at end of the US Civil War in 1865. I then describe the political behaviors undertaken by civilian and military actors in the events leading to Johnson's impeachment and Grant's nomination to the Presidency. 

Before proceeding, it is important to point out that the impeachment of Andrew Johnson, and the Army's closeness to the events leading up to it, was undoubtedly (and thankfully) atypical. Yet this period also demonstrates, in my view like none other in American history to date, how high levels of polarization and military prestige can combine to generate a toxic political atmosphere with respect to civil-military conduct. It is also important to point out that in the historiography of this early period, clear biases are present. Some historians, such as Sioussat (1913) and Milton (1930), are sympathetic to President Johnson, and thus very critical of Grant and others who opposed the President. Others, such as Badeau (1887), White (2016) and Chernow (2017) are extremely critical towards Johnson, and thus ebullient in their praise of Grant. Others, in my view, including Hesseltine (1935) and Simpson (1988) seem to fall somewhere in the middle. Throughout this section of the chapter, I seek to remain as neutral as possible, and to simply present the events that occurred, and how high levels of polarization and military prestige contributed to these events taking place. 

## High Level of Polarization, 1865-1868

It is difficult to state the precise level of polarization that prevailed at the end of the Civil War, but there is no doubt that the level was significantly high. The fact that a Civil War - the only one in US history - had just occurred, primarily over the issue of slavery, provides an immediate basis for this assertion. Standard measurements of polarization in Congress, however, usually begin with the 45th Congress, which met from 1877-1879 [@jeffrey_b_lewis_voteview_2020]. This is again because of the Civil War and its aftermath. When the first state, South Carolina, seceded from the Union in December 1860, until the Compromise of 1877 effectively ended Reconstruction, several southern states either lacked Congressional representation altogether or had been prevented from electing members of Congress on their own, absent of federal interference.^[The Compromise of 1877 was, in essence, a compromise about the 1876 Presidential Election of 1876. Democrats more or less agreed to a compromise such that the Republican candidate, Rutherford B. Hayes, would become the President, while Republicans in Congress agreed that the last of federal troops would depart Louisiana and South Carolina, effectively ending Reconstruction. See @hoogenboom_rutherford_1995, 274-294 for an account of this compromise by one of Hayes' biographers.] Still, the lack of a precise measure of polarization until the late 1870s need not leave the reader wondering whether polarization was high at the end of the Civil War in 1865.  

Polarization was high at the end of the Civil War as the country sought to "bind up the nation's wounds" [@lincoln_lincolns_1865]. To begin with, the nation had to face and answer questions concerning how best to bestow new political freedoms, to include the right to vote and own land, to more than four million former slaves [@green_reconstructing_2010, 186-193].^[For just a brief sampling of the rich literature that describes the enormous political questions facing the nation at the end of the Civil War, see @mckitrick_andrew_1988; @green_reconstructing_2010; @sheehan-dean_northern_2014; @smith_stormy_2017; @sheehan-dean_northern_2019.] 

The US government also faced other significant legal and moral questions pertaining to how to handle and relate to the states that had seceded. Indeed, McKitrick (1988) argues that Reconstruction, at its heart, boiled down to two central questions: "what had the war done to the rebellious states? And what was the new relation of those states - or districts - to the federal government" [@mckitrick_andrew_1988, 96]? While this chapter cannot unpack all of the differing views on these questions, it is worth summarizing a few of the opinions held by prominent civilian leaders at the time to illustrate how widely opinions varied on these topics. 

Some politicians, such as Thaddeus Stevens, a Pennsylvania Congressman who bitterly opposed President Johnson and who was known as a Republican 'Radical,' believed that the southern states should be administered as "conquered provinces," much in the same way a victor in war might exercise rule over a formerly hostile foreign power and its inhabitants [@mckitrick_andrew_1988, 99-101]. Stevens and others who held to this view believed that the former Confederate states had forfeited any of the legal rights that they had once possessed under the Constitution, that the entire way of life in the south would have to be forcefully remade, and perhaps most controversially, that private property belonging to southerners should be seized, confiscated, and redistributed [@mckitrick_andrew_1988, 99-101]. 

On the other end of the spectrum of views regarding reconstruction is what McKitrick (1988) calls the "southern theory" of Reconstruction [@mckitrick_andrew_1988, 97-98]. This view advocated for a relatively quick restoration of full rights to the former Confederate states on the basis that these states had tried to secede, but had failed, and thus should be welcomed back into the Union with little or no hindrance [@mckitrick_andrew_1988, 97-98]. Now that the war was over, advocate of this view argued that there was little need for punitive action.^[McKitrick (1988) points out that it was not only those from the south who held this view. Among the most ardent supporters of this "southern theory" of reconstruction was none other than General William T. Sherman, who, after Lee's surrender at Appomattox, infamously offered generous peace terms to Confederate General Albert Sidney Johnston. These terms of peace included a general pronouncement of amnesty for former Confederate officers. But Sherman quickly encountered political turmoil after offering these terms. When General Grant, as Sherman's superior, learned what had happened, he intervened and revoked the terms. Grant understood that Sherman, by brokering terms of political restoration, had not only crossed into a realm that only statesmen should occupy, but had offered terms that were politically unpalatable at the time. See @mckitrick_andrew_1988, 98; and @simpson_ulysses_1988, 95-100; for more details.] 

In short, each of the different views regarding how the south should be reconstructed involved a host of significant moral and practical beliefs about what actions the federal government should take with respect to the southern states and its inhabitants, including both former confederates and former slaves. Should former confederate officers and soldiers be granted amnesty for their past actions in taking up arms against the union? Should such persons be allowed to hold political office, and if so, would they have to denounce their past behavior or declare loyalty to the federal government? How much help should the federal government provide, and how quickly should they provide it, to ensure the implementation of new freedoms to former slaves?^[See @mckitrick_andrew_1988, 120-152 and @green_reconstructing_2010, 186-194 for a concise treatment of these issues]. 

Adding to the weight of these significant questions was another crucial issue, which was _who_ was ultimately responsible for overseeing and implementing the nation's reunification. Was the President ultimately responsible, or should Congress set the terms for welcoming the recalcitrant southern states back into the Union after four years of Civil War [@mckitrick_andrew_1988, 103-109]? Although President Lincoln had claimed significant executive powers, some of which rested on perhaps questionable invocations of wartime necessity, Republicans in Congress were not willing to grant Lincoln's successor the same benefit of the doubt, especially in the wake of Lincoln's assassination and because the war, at least in terms of open battles fought, had formally ended [@mckitrick_andrew_1988, 106-107; @hearn_impeachment_2000, 29-32].  

The Army and its senior leaders came to adopt strong opinions about these and related issues as the Army became a main actor involved with Reconstruction. Indeed, the Army came to embark on what should be considered nothing short of an occupation of the American south that would last more than a decade. For much of this unprecedented period, Army forces in the south and their commanders took on incredible responsibilities for overseeing civilian government, administering martial law, and exercising authorities similar to those that exist in wartime [@downs_after_2015; @lang_wake_2017]. 
Like the public, however, military leaders were divided about how Reconstruction should proceed, and what specific policies Reconstruction should entail. Some leaders, including Generals Ulysses S. Grant and Phil Sheridan, came to adopt views that were consistent with Republicans in Congress, favoring strong involvement by the Army and the federal government. Other military figures, including Generals William Sherman and Winfield Scott Hancock, held views that were similar to the President, and favored states' rights, and thus a more limited role for the Army in the South [@welles_diary_1911; @milton_age_1930; @hesseltine_ulysses_1935; @jordan_winfield_1988;@simpson_let_1991; @marszalek_sherman_2007; @white_american_2016; @chernow_grant_2017]. 

That differing views regarding Reconstruction split the Army's officer corps is perhaps best personified by the relationship between and subsequent political futures of General Ulysses S. Grant and Major General Winfield Scott Hancock, two generals who appear in this case study. The former became President, serving as a Republican for two terms from 1868 - 1876. The latter nearly obtained the Democratic Presidential nomination in 1868, and obtained it outright in 1880. Though he never won the Presidency, Hancock years later noted that it was his behavior during his brief tenure in New Orleans as the commander of the 5th Military District - an appointment Grant had worked hard to prevent - that propelled him to instant acclaim among Democrats. "Of the war men available," Hancock noted, he had, unlike Grant and other generals who opposed the President, exhibited "deference to civil law and the Constitution" [Hancock, as quoted in @jordan_winfield_1988, 218]. 

## Extremely High Level of Military Prestige, 1865-1868

The level of military prestige was also high at the end of the Civil War, and for one central reason: the Army had achieved victory in preventing the dissolution of the nation. This was especially true of many general and admirals who had achieved nationwide fame as a result of their martial exploits.  

While it is not possible to list all of the individuals who achieved heroic stature as a result of their martial exploits during the Civil War, it is worth pointing out several, who, in the years after the war, remained in the Army. Among these included General Philip Sheridan, the famed cavalry officer; General William Sherman, known for his ruthless march through Georgia near the close of the war; General Winfield Scott Hancock, who repelled the Confederate charge at Cemetery Ridge at the battle of Gettysburg; and many, many others.^[These include figures such as John M. Schofield, who would play a significant role in the Army's reform efforts late in the 19th Century.] However, while these officers were well known and popular figures, the prestige that accompanied General Ulysses S. Grant, who became the Army's commanding general at the close of the war, was second to none. 

Grant was the last of Lincoln's commanding generals of the U.S. Army and the military leader who ultimately brought about the defeat of Lee's Army. A West Point graduate and veteran of the Mexican War, Grant had resigned his officer commission in 1854 as a Captain. However, after several years in failed private business ventures, Grant was recommissioned as a Colonel in the Illinois Militia when the Civil War broke out in 1861, and he quickly established himself as an able fighter and commander [@white_american_2016].

So popular and so recognized by the American public was Grant at the end of the Civil War that Hesseltine (1935) argues Grant "was rapidly on his way to becoming a saint" [@hesseltine_ulysses_1935, 47]. Furthermore, multiple biographers of Grant recount that the general spent several weeks during the late summer of 1865 on a traveling tour of the northern states, where he was fervently met at every stop, and often regaled with large and lavish receptions, parties, and other gifts [@hesseltine_ulysses_1935, 57-62; @chernow_grant_2017, 558-560]. A significant portion of the northern press was also infatuated with Grant. For example, a _New York Daily Tribune_ article written on April 20, 1865, said the following of the general: "Three captains created the art of war as it exists today in Europe - Marlborough, Frederick, and Napoleon. Grant's last campaign entitles him to rank with either of the two former, and in several important elements of military greatness places him on a level even with the great French Captain" [@noauthor_grants_1865]. Adam Badeau, one of Grant's subordinate staff officers who spent many years observing Grant as an eyewitness to the events written about in this chapter, would write in 1887, "It is impossible to understand either Johnson's forbearance or Grant's authority all through this epoch without bearing constantly in mind that Grant was the most popular man in America" [@badeau_grant_1887, 68-69].

Grant's popularity translated into political capital, and civilians understood this reality. Horace White, the editor for the _Chicago Tribune_, for instance, wrote to Republican Congressman Elihu Washburn at the end of 1865 that, "I am persuaded that Grant is to be the next President if he gives us half a chance to nominate him" [White quoted in @hesseltine_ulysses_1935, 63]. Chernow (2017) goes so far as saying that in the years after the Civil War, Grant "became the supreme prize in American politics, sought by all parties" [@chernow_grant_2017, 597]. With his headquarters in Washington D.C., Ulysses S. Grant found himself squarely in the middle of high stakes political debates about how Reconstruction should proceed, and what policies Reconstruction should entail. Later in this chapter, it will become clear that the same was also true of many of Grant's subordinate officers, many of whom were also highly prestigious and popular figures.  

For the first several weeks, and perhaps months, the relationship between Grant and President Johnson was relatively free of friction, even, as Hesseltine describes, a "political cauldron" was boiling all around [@hesseltine_ulysses_1935, 54]. Over time, however, Grant and several of his subordinate commanders would no longer be able, nor desire, to remain outside of politics. 

In the next section, I narrate and describe the political behaviors engaged in by civilian and military leaders from roughly May 1865 - May 1868 that, even by historical standards, violated the principles of civil-military conduct. The levels of polarization and military prestige were both extremely high, and combined in such a way so as to generate extreme and frequent political behavior by both civilian and military leaders. I describe these behaviors generally in chronological order in the following section.

## Political Behaviors Involving the Military, 1865-1868

As divisions over reconstruction became evident in 1865 between President Johnson and his opponents in Congress, both entities sought to bring the military into their respective camps. Johnson and members of his administration, as well as members of Congress, sought General Grant's approval on the specific measures that Reconstruction should entail. Indeed, in the months that followed, as the divisions between Johnson and Congress became increasingly irreconcilable, both Johnson and his opponents in Congress would increasingly wonder, as White (2016) notes, "where did Grant stand" [@white_american_2016, 429]? Badeau observed that as early as 1865, "both parties to the contest wanted to use the prestige of his (Grant's) name; both laid their arguments before him and sought to secure his support" [@badeau_grant_1887, 35, parentheses mine].

That civilian leaders of both political parties maneuvered and jockeyed for Grant fully aligns with the predictions made by the theory advanced in this dissertation. Civilian leaders acted from the belief that General Grant, as the nation's most prestigious figure, commanded not merely the Army, but the potential to shape and influence public opinion in a way that was politically advantageous. Adam Badeau's observations of the President's early overtures towards Grant clearly signify the link between Grant and his utility to civilian leaders to shape public perception. Reflecting back on the first few months of Grant's professional relationship with the President, Badeau notes that "the great popularity of Grant at this period made it important to win him over to the support of the enterprise...It was, however, not so much Grant's real concurrence as the appearance of it before the world that Johnson probably sought" [@badeau_grant_1887, 32, 39]. Grant personally understood that political parties saw him as politically powerful and useful. He wrote his friend, General William Sherman, in 1867 that, "...there is little difference between parties. No matter how close I keep my tongue each tries to interpret from the little I let drop that I am with them" [Ulysses Grant, 1867, cited in @badeau_grant_1887, 73]".

### The Military as Sales Reps and Public Appeals

One of the first political behaviors the President engaged in was sending Grant on a tour of the South in late 1865. The purpose of sending Grant to the South was to refute a previous report made by Carl Schurz, a journalist and former general whom Johnson had previously sent to assess the general political atmosphere of the southern states. Schurz's report was politically problematic for Johnson, however, because it portrayed southern attitudes as "angry and defiant," and indicated that former slaves who had been freed were "languishing in wretched conditions of poverty, reinforced by Black Codes that trapped them in a new subservience" [@chernow_grant_2017, 563]. In short, Schurz's report did not portray a south that was remorseful and prepared to rapidly reunify with the nation, but rather one that required a lengthier and more robust federal presence [@chernow_grant_2017, 563]. 

Johnson hoped that Grant, after touring the south, would write a report that instead described the political conditions there as conducive to a relatively quick restoration, in line with Johnson's views on Reconstruction [@hesseltine_ulysses_1935, 60; see also @chernow_grant_2017, 563-564]. Such a report, written by a highly influential military figure like Grant, would arm the President with a political weapon to employ against his opponents in Congress. After Grant returned from his visit, Johnson summoned Grant to a cabinet meeting where the general "expressed his conviction that every consideration called for the early reestablishment of the Union [@hesseltine_ulysses_1935, 61]. Grant's comments were immediately sent off to the Senate, and used by Johnson's administration to defend against attacks made by Radical Republicans [@hesseltine_ulysses_1935, 61; @chernow_grant_2017, 565-566].^[Chernow (2017) argues that Grant's visit was superficial in that it was rushed and largely featured sham interactions with southern whites who praised Grant. Chernow also notes that Grant regretted his report soon after furnishing it. See @chernow_grant_2017, 565-566 for more details. Furthermore, both Hesseltine (1935) and Milton (1930) argue that in the wake of Grant furnishing his report, Radical Republicans fervently tried to get Grant to denounce it. See @hesseltine_ulysses_1935, 65; @milton_age_1930, 290 for more details.] 

The President's behavior is synonymous with both employing the military as a sales representative, and harnessing the military to make a public appeal, both of which were discussed in chapter two. The main audience Johnson sought to influence was not the south, but rather his opponents in Congress. Furthermore, in sending Grant to the southern states for the purposes of having Grant make a politically advantageous and public report, the President challenged at least two of the normative principles of civil-military conduct that existed at the time. 

First, Johnson's behavior violated the principle of non-partisanship because there was effectively no possible way that Grant could generate a report that did not anger either Johnson's supporters or his opponents in Congress. Furthermore, Johnson's chief purpose in sending Grant was to provide both the American public and Congress the impression that Grant's views on Reconstruction favored his own. Johnson knew that validation of his views by Grant would disadvantage his opponents in Congress, if even for a short period of time.  

Second, Johnson's dispatching of Grant to the south for the purposes of Grant writing a subsequent public report also violated the principle of non-interference because, in effect, Johnson made a sitting military commander investigate, evaluate, and publicize prevailing political conditions in ways that were more comparable to what a reporter or journalist, rather than a military officer, would do. To be clear, there is nothing inappropriate for a President to send a military officer to learn more about a situation, or to report back with intelligence on a number of possible scenarios. But that is not what happened here. Grant did not possess any superior reporting or communication skills than the author of the initial report, Carl Schurz. And in fact, it is almost certain that Grant's popularity prevented him from accessing people during the trip who would likely have conveyed to Grant far less rosy of a situation  [@hesseltine_ulysses_1935, 60]. 

Schurz was a career journalist who had volunteered to serve in the Union Army, and would become the editor of the _Detroit Post_ in 1866. This matters because Schurz's initial report was written from the standpoint of a professional reporter or journalist [@simpson_let_1991, 116-188]. Johnson could have dispatched others to refute Schurz's report, including perhaps other press correspondents who likened to his views. Instead, in sending Grant, Johnson sought to harness Grant's enormous prestige, knowing that the public would be interested, and that Congressional Republicans would have to contend with, Grant's evaluation. Still, In asking Grant to fill a role that was better suited for a press correspondent or journalist, President Johnson compelled the nation's top Army officer to assume a role that was tangential, if not altogether unrelated, to his role as the general-in-chief of the Army. 

After Grant's report became public knowledge, the final months of 1865 and early months of 1866 evinced even greater divisions between Johnson and his Republican opponents in Congress. As time passed, it became increasingly clear that Johnson's conception of Reconstruction clearly favored southern whites with ties to the former confederacy, and at the same time, disadvantaged both former slaves and whites who had supported the union. For instance, newly freed blacks frequently faced increasing danger as southern state and local governments enacted a series of "black codes" that authorized a variety of discriminatory measures against blacks and former slaves [see for example, @mckitrick_andrew_1988, 204-205; @chernow_grant_2017, 571]. 

Though the 13th Amendment (which formally abolished slavery in the United States) went into effect in December of 1865, Johnson continually vetoed legislation that sought to expand the rights and political freedoms of the former slaves. These included bills which sought to extend the life of the Freedmen's Bureau, an organization created in 1865 to "provide food, shelter, clothing, medical services, and land to displaced southerners, including newly freed African Americans" [@noauthor_us_nodate], and later, a civil rights bill which, while "silent on voting rights," nonetheless sought to make the federal government the "guarantor of basic liberties" [@chernow_grant_2017, 570]. Both of these pieces of legislation would eventually end up passing as Congress voted to override vetoes by Johnson.

### Showcasing the Troops and Electoral Swaggering 

With legislative battles between Congress and the President unfolding, Johnson continued to harness General Grant's prestige and that of other military leaders for partisan reasons in the months leading up to the 1866 midterm elections. These actions occurred in relatively smaller and larger forms. On several occasions, civilian actors deliberately orchestrated their public appearances in such a way so as to physically appear alongside popular military figures in order to give the impression that these military leaders approved of their policies and views on Reconstruction. This type of behavior constitutes showcasing the troops and engaging in electoral swaggering, both of which were discussed in chapter two. Furthermore, this type of behavior is congruent with the civilian act of featuring military actors to appear in presidential campaign ads, which was examined in the previous chapter.

On August 18, 1866, President Johnson hosted "a delegation from the National Union Convention in Philadelphia" at the White House [@chernow_grant_2017, 576]. In spite of the clearly partisan nature of the gathering, Grant reluctantly accepted Johnson's invitation to attend, as he "felt duty-bound" to do so [@chernow_grant_2017, 576]. Before addressing his supporters in the East Room of the White House, the President summoned Grant, where the two stood next to each other during all of the President's remarks [@chernow_grant_2017, 577]. After the event, Grant returned to his headquarters, where he was "full of indignation" at having been manipulated by the President [Badeau, 1887, as cited in @chernow_grant_2017, 577]. Ultimately, Grant's physical proximity to the President on that particular August evening insinuated, at least to one reporter present at the event, that Grant fervently endorsed Johnson's policies [McFeely, 2002, as cited in @chernow_grant_2017, 577]. That particular evening, the President had showcased the nation's most highly regarded soldier. 

An even greater instance of Johnson harnessing Grant's prestige and that of other military leaders occurred at the end of the summer of 1866, however, when Johnson insisted that Grant and several other military figures accompany him on a two week long public relations tour of several Northern states and populous cities that came to be known as the "Swing Around the Circle." Traveling by train and designed to drum up support ahead of the midterms, the President brought with him several of his cabinet members, but also several military figures, including General Grant, Admiral David Farragut, General George Meade, and others. 

Johnson's touring with prestigious military figures intended to convey to the public that prominent military figures approved of Johnson's policies, which included at the time a strong denunciation of the 14th Amendment, which had been passed by the Senate in June [@hesseltine_ulysses_1935, 74-76; @chernow_grant_2017, 577-581].^[The 14th Amendment granted citizenship to all persons born in the United States, including former slaves. Except for Tennessee, the southern states rejected the 14th Amendment initially, but the Military Restoration Acts of 1867 later required ratification of both the 13th and 14th Amendments for states to be restored to the Union.] The President traveling and appearing with several military figures with him on a speaking tour of the country ahead of the elections was synonymous with using military figures to appear not only in campaign ads, but perhaps in even more obvious political forums, such as attending or speaking at a partisan political convention.  

Most historical accounts of the President's trip note that the tour was a public relations disaster. Johnson argued with the crowds at almost every stop, and many of those who turned out asked instead to hear from Grant and Farragut rather than from the President [@hesseltine_ulysses_1935, 75; @chernow_grant_2017, 577]. Importantly, Milton (1930, pp. 362-369), Hesseltine (1935, pp. 74-75), Simpson (1991, p. 148), and Welles (1911, 591-596) all argue that Johnson's political opponents - those allied with the Radical Republicans in Congress - went to great lengths to orchestrate the local crowds at several stops along the tour to cheer loudly for Grant and Farragut, and to heckle the President. Intended to ultimately "prevent the President from speaking or to embarrass him in his remarks," these attempts by Johnson's opponents to use the prestige of Grant and other military leaders against the President clearly had a partisan tone [@welles_diary_1911, 593]. Furthermore, that both Johnson and his opponents both sought to pit Grant against each other is in full alignment with the predictions made by this dissertation's theory. 

### Brass in the Cabinet and Explicit Insubordination

The nation's political climate continued to worsen in 1866 after race riots, first in Memphis and then in New Orleans, had killed dozens of freed blacks and sympathetic whites [@mckitrick_andrew_1988, 422-427; @hyman_santon_1960, 90]. Events such as these further cemented partisan battle lines, even among Army leaders: along with the Secretary of War, Edwin Stanton, Generals Grant, Sheridan, Pope, and Meigs came to associate themselves with a theory of Reconstruction congruent with most Republicans in Congress, while Generals Sherman and Hancock favored Johnson's [@hyman_santon_1960, 93]. As the political climate worsened, the severity of the political behaviors that occurred with respect to the military continued to escalate. 

Johnson first tried to bypass the combined team of Grant and Stanton, who by late 1866 had both come to believe that Congressional Republicans, rather than the President, possessed a view of Reconstruction that was best for the nation and for the Army [@hyman_santon_1960]. Too prestigious to dismiss outright, Johnson instead sought to send Grant to Mexico as part of a diplomatic mission, which would at least temporarily remove the General from Washington, and to replace Stanton with General Sherman. 

The high prestige of both Generals Grant and Sherman played an important role in both the President's decision and also Grant's subsequent reaction to the ordeal. To begin with, as Colonel W.G. Moore, Johnson's private secretary, notes, Johnson chose to send Grant to Mexico "in order that prestige might be given to his mission" [Moore, 1866 quoted in @sioussat_notes_1913, 99]. Furthermore, the President considered implementing this scheme in conjunction with leaking a letter he had received from Sherman in which the general had expressed support for the President's policies [@hesseltine_ulysses_1935, 78]. In short, if replacing Stanton and removing Grant temporarily had even a remote chance of working, the President would have to do two things. First, the President would have to make the public think that sending Grant to Mexico was a vital mission and worthy of directing someone of Grant's stature to go. And second, the President would have to replace Stanton with someone who was highly regarded by the public, such as General Sherman.

However, Grant refused Johnson's directive to go to Mexico several times, the last of which occurred in a rather public and dramatic manner. In front his assembled cabinet, the President asked the Attorney General whether his orders to send Grant to Mexico were illegal, ostensibly in order to publicly shame Grant, who was present for the meeting, into compliance. Hesseltine (1935) depicts Grant as then standing up before the Attorney General could respond, stating, "I can answer that question Mr. President, without referring to the Attorney General. I am an officer of the army, and bound to obey your military orders. But this is a civil office, a purely diplomatic duty that you offer me, and I cannot be compelled to undertake itno power on earth can compel me to it" [Badeau, 1887 as cited in @hesseltine_ulysses_1935, 79]. Importantly, Moore (cited in Sioussat, 1913, pp. 101-102) also notes that Grant was very much aware of just how beloved Sherman was by the country, and that this likely played a factor in Grant's refusal to obey the President's order. In fact, Moore notes that the President read Grant the private letter Sherman had sent, and that the President told Grant that "he (the President) thought of publishing it - a suggestion which...the General did not appear to relish" [Moore, 1866 quoted in @sioussat_notes_1913, 101]. This means that Grant's refusal to obey the President's order - unwise, crafty, manipulative, yet still legal -  and thus to engage in an act of explicit disobedience, resulted at least partially from Grant's awareness that if Sherman were to replace Stanton, Sherman's high personal prestige in the cabinet could lead to the very outcomes with respect to Reconstruction that Grant and Stanton were trying so very hard to avoid. 

This entire episode of Johnson's attempt to send Grant to Mexico and replace Stanton with Sherman is interwoven with the variables of high levels of polarization and military prestige. The high level of polarization prevalent at the time explains why Johnson tried to rid himself of Grant and Stanton in the first place, and Grant's very real concern that harm could arise if he left Washington. The high level of military prestige explains both how and why the scheme unfolded in the manner that it did. First, the President's scheme could only be successful if he chose to replace Stanton with Sherman, whom Hyman (1960) argues was the "nation's third most popular man" at the time [@hyman_santon_1960, 94]. Second, Grant knew that Sherman was sufficiently prestigious such that the President's scheme could actually work, and he feared the consequences of its implementation. Finally, Grant also knew that he was himself prestigious, and that he could refuse the President's directive and not be opposed in doing so. This incident clearly shows that not only did civilian leaders attempt to harness the prestige of military leaders, but that in this instance, Grant's high prestige actually prevented the President from exercising his lawful, albeit unwise and manipulative, authority.

### Partisan Stacking and Selective Compliance

The battle between Congress and the President continued to escalate, and on March 2nd, 1867, Congress passed (again, over a Presidential veto) the first of several Military Reconstruction Acts. Among other impacts, this legislation split the south into five military districts, each of which would be overseen by an army general. The legislation intended for these military commanders to hold enormous sway over the daily conduct of state and local governments, which were to be regarded by the commanders as "provisional" in nature pending their "full" restoration to the union [Grant, 1867 as cited in @chernow_grant_2017, 587]. Importantly, the legislation also stated that all orders to the Army from the President had to be routed through Grant [@chernow_grant_2017, 587].^[The Military Reconstruction Acts also empowered Grant to appoint these five generals himself, but even Grant had reservations about this, believing that the President, as the Commander in Chief, should have a strong say in who these generals were. See @hesseltine_ulysses_1935, 83 and @chernow_grant_2017, 585 for more details.] 

At the same time, Congress also enacted the Tenure of Office Act, which sought to preserve Stanton's place in Johnson's cabinet, and thus his link to Republicans in Congress [@hearn_impeachment_2000, 110-119; @chernow_grant_2017, 586-587]. This legislation specifically declared that the President could not remove cabinet officials that he had not appointed himself without the Senate's approval [@chernow_grant_2017, 586]. Indeed, Johnson had inherited Stanton, who had also served as Lincoln's Secretary of War. Together, the Military Reconstruction and the Tenure of Office Acts effectively cut off the President from exercising meaningful control of the Army in the south. Multiple scholars shed light on this central point from different angles. Downs (2015) makes the compelling point that in many ways, American Reconstruction was a fight over who - the President or Congress - has the power to end a special type of (Civil) war [@downs_after_2015, 7-8]. Hyman (1960) is perhaps even more stark. He argues that in essence, two armies existed when the Civil War ended; one was controlled by the President, and had duties on the American frontier and along the US border with Mexico. The other army that existed was the one that implemented Reconstruction in the south. This army, Hyman argues, is the one that Congress would never allow President Johnson to control. The stakes in Reconstruction were simply too high [@hyman_santon_1960, 86-87]. 

As the political situation escalated, so too did the political behaviors undertaken by civil and military actors. The President first tried to replace several military commanders who implemented Reconstruction policies that ran counter to the President's views. Of course, this in turn made the tension faced by Grant even more severe. On one hand, he strongly disagreed with the President about Reconstruction, but on the other hand, he strongly believed that the President should act and function as the Commander of the nation's Armed Forces.^[Chernow (2017, p. 586) notes that Grant personally felt the Tenure of Office Act was "ridiculous" in that it essentially rendered the President powerless, but that he nonetheless sought to obey the law that had been passed.] Before explaining these events in greater detail, it is worth pointing out that this dynamic - of a President cut off from exercising his constitutional powers - speaks to the power of high polarization in shaping the nation's civil-military dynamic. At this point in US history, in 1867, there was little if any difference between the nation's political and military policy. In other words, extremely high polarization had generated a political atmosphere such that civilian and military spheres very closely, if not entirely, overlapped. Grant was living in a world that might be shown by the right side of Figure \@ \@ref(fig:nonint), which was shown way back in chapter one of this dissertation. 

Today, it is difficult to generate a contemporary hypothetical scenario that corresponds to what occurred in 1867, but one such possibility might be if a future President, following several shootings at multiple abortion clinics across the country, ordered military forces to secure abortion clinics nationwide. Imagine, then, for a moment, that the President under such conditions began to appoint and fire four star generals and admirals solely on the basis of these officers' convictions about abortion rights. Though hypothetical, I suspect that were such a scenario to unfold, some percentage of military officers, probably those who hold to a "pro-life" view regarding abortion, would not carry out the President's orders, and would instead either resign or disobey. I do not know this for sure, and it is entirely speculative. The more crucial point to make, though, is that under such polarizing conditions, the President's choice of commanders would no longer be on the basis of how well a particular officer can train recruits, resupply military forces, or synchronize an attack of an enemy urban stronghold. Instead, the real question the President in this hypothetical scenario would want to know is, "will this officer fulfill my wishes on this issue?" In non-polarizing political times, this is a perfectly reasonable question for a President to ask. In extremely polarizing times such as in 1867, the question may still be appropriate, but the issue over which such a question is asked is less likely to result in compliance.  

This highly polarizing dynamic existed in the United States in 1867, and the President began to engage in behavior that amounted to partisan stacking, which was explained in chapter two. Purely partisan convictions, rather than military skill or excellence, became the basis of selecting and assigning military leaders. In August, Johnson informed Stanton that he was suspended (not "removed," so as to avoid being instantly in violation of the Tenure of Office Act), and proceeded to appoint Grant as the Secretary of War _ad interim_ until a permanent replacement might be named [@hyman_santon_1960, 98]. The President then removed General Philip Sheridan, who commanded the 5th Military District (Louisiana and Texas), after Sheridan removed the Governors of both Louisiana and Texas [@white_american_2016, 448-449; @hyman_santon_1960; 98-99]. Sheridan, a favored subordinate of Grant's who favored an aggressive Republican policy towards Reconstruction, had helped register thousands of blacks to vote in the district, earning widespread condemnation from Johnson and southern whites. Johnson also later removed the commanders of the 2nd and 3rd Military Districts, Generals Daniel Sickles and John Pope, for similar reasons.^[Johnson's replacement commanders were all known to be sympathetic to Johnson's views on Reconstruction, especially General Winfield Scott Hancock, who replaced Sheridan in the 5th Military District.]  

Before removing Sheridan, however, the President first asked Grant's opinion on the matter. Grant responded to the President in a private letter that it was "unmistakably the expressed wish of the country that General Sheridan not be removedthis is a Republic where the will of the people is the law of the land. I beg that their voice may be heard" [Grant, 1867 as cited in @milton_age_1930, 458]. Johnson then had his Attorney General, Henry Stanbery, produce an opinion strongly denouncing Sheridan's actions, and ordered Grant to distribute Stanbery's opinion to his commanders [@hesseltine_ulysses_1935, 85; @chernow_grant_2017, 589]. Grant complied, but not entirely, with Johnson's order. He distributed Stanbery's opinion, but simultaneously let his commanders know that "they could freely interpret" Stanbery's opinion as they saw fit [@chernow_grant_2017, 589]. 

Grant's behavior in this instance constitutes a form of selective compliance. Though Grant perhaps thought Sheridan's methods were a bit too strong, he did not fault the underlying premise of Sheridan's behavior: the Governor of Louisiana and other civilian officials whom Sheridan had previously removed had been responsible for preventing blacks from registering to vote, facilitating violence during race riots in New Orleans the previous year, and generally thwarting the implementation of key Reconstruction policies [@dawson_army_1994, 46-50]. Those who closely observed this episode noted that, "the President, of course, observed [Grant's] tacit disobedience, but he was powerless to control or punish his subordinates. He had disregarded the will of Congress, and in return the officers of the Army disregarded his. The situation was approaching mutiny on one side, or else treason on the other" [@badeau_grant_1887, 71].

### More Public Appeals

After publication of Stanbery's opinion, Sheridan then wrote Grant a letter in which Sheridan chastised the President [@hesseltine_ulysses_1935, 85-86]. The letter was leaked and published, though it is not clear who was responsible for the leak. The President ultimately decided not to punish Sheridan for the letter on the grounds that it had been written, at least plausibly from Sheridan's perspective, as a private rather than public letter [@hesseltine_ulysses_1935, 85-86]. However, Republicans in Congress were greatly happy with Sheridan's criticism of the President, and used it as grounds to make further overtures towards Grant [@hesseltine_ulysses_1935, 85-86]. Grant also furnished the contents of a previously private letter he had sent to the President about the possibility of removing Sheridan to several press outlets, including the _Army and Navy Journal_, which was popular with military audiences [@chernow_grant_2017, 596]. Grant's behaviors - directing his subordinates not to reinstate any politicians Sheridan had removed, and then leaking a letter to the press - challenged both the principles of civilian control and non-partisanship. 

In early January of 1868, the situation involving Johnson, Grant, and the Army finally moved towards a climax. The Senate first released a report refusing to concur with Stanton's suspension, which in turn raised immediate questions as to whether Grant's serving as the _ad interim_ War Secretary was legal or not [@mckitrick_andrew_1988, 501-505]. Grant did not wish to find out the answer, and thus he attempted to vacate the office, which Johnson would not let him easily do [@mckitrick_andrew_1988, 501-502].

Exactly what transpired next is not so easy to discern, even for contemporary historians [@chernow_grant_2017, 602]. The critical question is what sort of mutual understanding, if any, Grant and Johnson had regarding Grant's continued service as the _ad interim_ War Secretary, and when such an understanding was reached. Those who are sympathetic towards Johnson allege that the two men agreed that Grant would either keep the office, even if the Senate or the Supreme Court would hold proceedings to decide the legality of Stanton's removal or that of the Tenure of Office Act altogether, or that Grant would vacate the office sufficiently ahead of time so as to enable Johnson to name a replacement, and thus, to not be found in violation of the law [@milton_age_1930, 481-485]. 

Other historians who are charitable towards Grant take a different view, and argue that Grant made it clear to the President that if Stanton were to be reinstated by the Senate, Grant would comply by returning the office to Stanton immediately [@chernow_grant_2017, 602-607]. Regardless of which view is historically accurate, it is clear that after a tense and heated cabinet meeting in January of 1868, Johnson's administration engaged in multiple efforts to discredit Grant in the press and to portray him as willfully acting in a "duplicitous" manner [@chernow_grant_2017, 605; @milton_age_1930, 483-485; @mckitrick_andrew_1988, 503; @hesseltine_ulysses_1935, 106-107]. For several weeks, the nation hung on press reports fueled by leaked letters, interviews, and other details provided by the Johnson Administration. 

The civilian acts of leaking information and giving interviews to the press in this instance violated the principles of civilian control and non-partisanship. They violated the principle of civilian control because they reflected the fact that by this point, the Administration was no longer attempting to exercise oversight of the military through any established mechanisms. Indeed, to be fair to Johnson and also historically accurate, many of these regular mechanisms had been removed by Congress through the Military Reconstruction Acts, which effectually made the Commander in Chief an observer of, rather than a participant of, the state's civil-military relationship. Leaking and giving interviews to the press to discredit Grant also violated the principle of non-partisanship, however. While it was well known that Grant differed with the Commander in Chief over Reconstruction policies by this point, the public nature of the leaks and interviews given by the Johnson Administration only served to publicize these differences. 

## Summary of the 1865-1868 Period

In summary, civilian and military actors engaged in several extreme forms of political behavior from 1865 - 1868. These behaviors ranged considerably, from Johnson sending Grant to refute Carl Schurz's initial observations of the south in 1865, to Johnson's harnessing of several military actors on his "Swing Around the Circle Tour" ahead of the 1866 midterm elections, to Grant's refusal in front of an assembled cabinet to be sent to Mexico. These behaviors were extreme forms of behavior, especially considering that many were undertaken by the nation's top civilian and military actor. 

Table \@ref(tab:pol-behav-cs1) summarizes the political behaviors undertaken by civilian and military actors from 1865 - 1868. 

```{r pol-behav-cs1, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl1 <- tribble(~"Actor", ~"Tactic of Political Behavior", ~"Date", ~"Instance", ~"Normative Standard Violated",
  "President Andrew Johnson", "Military as Sales Reps", "1865", "Johnson dispatches Grant to refute a previous report made by Carl Schurz regarding political conditions in the south", "Non-partisanship; Non-interference",
  "President Andrew Johnson", "Electoral Swaggering", "1866", "Andrew Johnson directs Generals Grant and Meade, along with Admiral Farragut, to accompany him on a tour of the nation ahead of the 1866 midterm elections", "Non-partisanship",
  "President Andrew Johnson", "Brass in the Cabinet", "1866", "Johnson seeks to replace Stanton as Secretary of War with General Sherman", "Non-partisanship",
  "General Grant", "Explicit Insubordination", "1866", "Grant refuses to be sent to Mexico by Johnson", "Civilian control",
  "President Andrew Johnson", "Partisan Stacking", "1867", "Andrew Johnson replaces Generals Sheridan, Sickles, and Pope as Military District Commanders", "Non-partisanship",
  "General Grant", "Selective Obedience/Compliance", "1867", "Grant publishes Attorney General Stanbery's legal opinion of the Military Reconstruction Acts in conjunction with President Johnson's wishes, but also instructs his commanders to make their own interpretations", "Civilian control",
  "General Sheridan", "Public Appeal (perhaps inadvertent)", "1867", "Sheridan writes a letter to Grant in which he upbraids the President's views on Reconstruction", "Civilian control; non-partisanship",
  "General Grant", "Public Appeal", "1867", "After Johnson removes Sheridan, Grant leaks a private letter he had sent to Johnson in which he protested the move", "Civilian control; non-partisanship",
  "President Andrew Johnson", "Public Appeal", "1868", "Johnson leaks the content of a heated cabinet meeting held in 1868 for the purposes of discrediting Grant", "Civilian control; Non-partisanship"
  )
 

text_tbl1 %>% kable(booktabs=T, caption="Political Behaviors Involving the Military, 1865-1868") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="8em") %>% 
  column_spec(3, width="4em") %>% 
  column_spec(4, width="14em") %>% 
  column_spec(5, width="8em")
  

```

Several of these behaviors, especially partisan stacking and explicit insubordination, constitute extreme forms of political behavior in the sense that they occurred at the very highest levels of civilian government and the military. That the President and the nation's top Army general engaged in these behaviors underscores both the high level of political polarization and military prestige that existed at the time. 

# Part 2 - The Army, Domestic Disturbance, and Failed Reform Efforts, 1877 - 1880

This chapter now turns to examining the political behaviors undertaken by civilian and military actors beginning roughly a decade later, from approximately 1877 - 1880. Similar to the first part of the case study, I first set the scene by arguing that although the level of polarization remained high from 1865 - 1880, the relative prestige of the military as an institution declined from 1868-1877. Furthermore, this decline in military prestige corresponds mainly with a reduction in the component of military prestige associated with the centrality or importance of the military, rather than with the component of prestige associated with the perceived skill or excellence of the military [@davis_conceptual_1942; @jaco_prestige_1970]. 

I then narrate several significant events that involved the Army from 1877 - 1880, including the use of the Army in several domestic labor disputes and bureaucratic efforts concerning Army reform. I argue that the general qualities of the political behaviors engaged in by civilian and military actors from 1877 - 1880 were noticeably different than those which both types of actors had engaged in from 1865 - 1868. In particular, civilians did not exhibit the same level of intensity to harness the prestige of the military because the military no longer afforded as significant a political advantage as it had immediately at the end of the Civil War. Furthermore, senior military leaders in the late 1870s themselves were fully cognizant that they now represented an institution that carried less political clout than it once did. This is not to say that military leaders were utterly powerless or lacked the ability to influence civilian leaders or the public altogether, but simply that in relative terms, the Army of 1877 was not as politically influential as the Army of 1867.

## High Level of Polarization, 1877-1880

The level of polarization remained quite high as Reconstruction ended. When we view Figure \@ref(fig:polar-chamber), we can see that after a couple of very small drops in the level of polarization in the US House, polarization only climbed beginning in the mid 1880s all the way until after the start of the 20th Century. Polarization in the Senate reflects a very similar trajectory. 

Scholars have noted that polarization in the 21st century shares more than just a couple of passing similarities with the polarization that prevailed in the late 19th century. These scholars make the point that in the late 19th Century, issues related to "race, morality, immigration, and themes related to populism" contributed to American polarization [@azari_back_2016, 93]. These themes remained salient as the United States emerged from Reconstruction and embarked on a process of significant industrialization throughout the decades of the late 19th Century [@skowronek_building_1982]. In short, the level of polarization remained relatively high well into the start of the 20th Century.  

## Declining Military Prestige, 1868-1877 

However, the relative level of military prestige declined during the period 1868 - 1877. The decline in the level of military prestige, furthermore, was not the result of one factor, but rather the confluence of at least three broad factors that emerged throughout the Reconstruction period. These interrelated factors are, first, the significant downsizing of the Army after the Civil War; second, a lack of clarity concerning the Army's _raison d'etre_ in an era in which the US faced no clear international threats; and third, a general sense of apathy towards the Army as a result of its involvement in the inherent political tasks associated with Reconstruction. 

The first two of these factors primarily weakened the prevailing sense of the Army's centrality to the life of the nation. The third factor, the perception held by society of the Army's role in Reconstruction, is more closely associated with the component of prestige that relates to society's evaluation of the military's skill or professional excellence. Together, all three factors combined to render the military less prestigious than it had been at the conclusion of the Civil War. I explain each of these factors below in greater detail, and defend the claim of the military possessing less relative prestige in 1877 than in 1865 by presenting the viewpoints held by several military officers who served in the late 1870s. 

The first factor that contributed to the Army's diminishing prestige was the significant reduction in manpower the service endured during the decade following the end of the Civil War. Within six months of Lee's surrender at Appomattox, "800,000 of the 1 million men in the Union's Grand Army of the Republic were mustered out of federal service" [@laurie_role_1997, 21]. Four years later, by 1869, the Army's authorized end strength stood at 37,313 [Utley, 1974 and Cooper, 1986 as cited in @clark_preparing_2017, 99]. By the end of 1875, the Army was capped at 25,000 men, including officers and enlisted men [@laurie_role_1997, 21].^[Laurie and Cole (1997, p. 22) also note that only 2,105 officers remained in the Army after 1871 - a figure that did not change by more than 100 men until the close of the century.]

Reductions in manpower involved issues beyond mere end strength, however. Woven into the question of how large the Army should be was the question of _how_ the Army should be sourced, and what kind of an Army was necessary. Did the nation need a standing Army, or were volunteer forces sufficient? Many Americans and Congressional leaders felt that the Army of the Civil War, which was comprised of a conglomerate of regular, volunteer, and militia forces, had served the nation well, and thus that a large, professional, standing Army simply was not necessary.^[Clark (2017) tells the full story of the Army's efforts to reform after the Civil War in his book, _Preparing for War: The Emergence of the Modern U.S. Army, 1815-1917_. See especially his fourth chapter on this point, "The Civil War's Legacy."] 

A second factor that contributed to the Army's diminished importance to the nation and thus, its diminished level of prestige, involved the lack of clarity regarding the primary purpose for which regular Army forces existed in an era of relative peace. Until the Spanish-American War, and with the exception of engaging in several specific conflicts with American Indians in Western territories of the United States, there were few significant threats the Army faced. The Army had also become increasingly scattered in a geographical sense. Laurie and Cole (1997, p.22) note, for instance, that upon the end of Reconstruction, "all of the cavalry forces and three fourths of the infantry" had moved to geographical outposts "west of the Mississippi Rive to battle hostile Indians and otherwise police the frontier." In the absence of a compelling international threat, Army leaders in the 1870s and 1880s were challenged to convince Congress that any meaningful reforms to the Army were necessary.^[Grandstaff (1998) and Clark (2017) conclude that the late 1890s and early 1900s, after the Spanish-American War, is the period in which the Army became more successful at gaining support for their intended reforms. See @grandstaff_preserving_1998, 537 and @clark_preparing_2017 for more details.] 

The third factor that contributed to the decline in the Army's prestige in the decade after the Civil War was its involvement in the inherently political tasks of Reconstruction. Though it is difficult to precisely gauge the impact of Reconstruction on the military's relationship with American society, we can reasonably discern that the Army came to be especially disliked and distrusted by Southern White Democrats [@byler_civil-military_2006, 6-11]. After all, Southern White Democrats had favored President Johnson's views on Reconstruction, and had also cheered the behavior of military figures such as General Winfield Scott Hancock, who, unlike Generals Grant and Sheridan, sought a limited role for military forces in the southern states.

The antagonism of Southern White Democrats only increased as Reconstruction continued for more than a decade. Byler (2006) notes that of particular angst to southerners was the way that federal forces had been used at voting locations during elections throughout the Reconstruction era, and particularly after 1870, when the military districts created by the Military Reconstruction Acts of 1867 ceased to formally function [@byler_civil-military_2006, 8]. So eager were Southern Democrats to rid their states from what they perceived to be heavy handed involvement by federal troops that in the aftermath of the contested Presidential Election of 1876, they essentially agreed to award the Presidency to the Republican Candidate, Rutherford B. Hayes, in exchange for a guarantee from President Grant that federal troops would depart the southern states [@woodward_reunion_1991, 6-8; @marszalek_sherman_2007, 428; @wooster_united_2021, 242].

These three factors - the significant downsizing of the Army and whether a professional Army was even required, the lack of any clear international threat, and significant partisan resentment over how the Army had been used during Reconstruction - combined to render the military less central and vital to the nation than it had been at the conclusion of the Civil War and in the immediate years which followed. Whereas General Grant, the victorious military commander of a brutal and lengthy Civil War, was highly sought after by all parties from 1865-1868, military leaders in the late 1870s and early 1880s, many of whom were certainly popular and well known figures, no longer afforded civilian leaders as significant a political advantage. Even as the level of polarization remained high, the context had changed insofar as the military was no longer as vital to the nation as it had been. 

To see this, consider that Army leaders themselves were cognizant, by the 1870s, that the institution of which they were a part occupied a less prominent role in the nation's life. For instance, Major General John Pope, whose 1867 letter to General Grant was likely leaked and published in _The New York Times_ and was referenced in this chapter's introductory paragraph, told a gathering of Civil War veterans in 1873 that he was concerned that the Army might "properly and naturally become an object of suspicion and dislike" in the years to come [Cozzens, 2000 as cited in @byler_civil-military_2006, 1]. In one private letter written in 1877, General William Sherman, the commanding general of the Army from 1869-1883 and himself a major hero of the Civil War, wrote, "Our enemy is not now the Indians but Congress" [Sherman, 1877 as cited in @marszalek_sherman_2007, 432]. Finally, Francis V. Greene, who graduated from West Point in 1870 and would fight in the Spanish - American War, would reflect in 1919 on the general atmosphere that prevailed in the late 1870s and early 1880s as follows: "in the midst of such pressing problems as the building of railroads, the resumption of specie payments, the silver question, and the tariff, there was no time to think about the needs of the Army. There was no 'military situation in the United States' worth thinking about" [Greene, 1919 as cited in @fitzpatrick_emory_2017, 223]. These are just three snapshots of three different officers, but each one clearly points to an Army that did not occupy a place of prominence in the hearts and minds of its citizens. 

## Political Behaviors Involving the Military, 1877-1880

In this section, I narrate two related episodes that occurred in the period from 1877 - 1878, and describe the political behaviors that military and civilian actors either undertook or did _not_ undertake as these episodes unfolded. The first was the use of regular Army forces to quell what would end up being several domestic disturbances in the late 1870s - 1890s.^[See @cooper_army_1980 and @laurie_role_1997 for excellent and in-depth accounts of these events.] The first of these domestic disturbances, known as _The Great Railway Strike of 1877_, began in Maryland in July of that year before spreading to West Virginia, Pennsylvania, and later, Chicago. The second episode came on the heels of this strike, and involved a congressional effort that tried (and ultimately failed) to reform the Army the following year in 1878. The political behaviors that occurred throughout these episodes were characteristically different in several important ways, which I describe below, than those that had occurred in the earlier portion of this case study, from 1865 - 1868.   

### Opportunities Not Seized to Make Public Appeals

_The Great Railway Strike of 1877_ and other strikes which followed occurred as workers, particularly in the rail and coal industries, protested what they perceived were unfair and inequitable working conditions, including the enaction of several wage cuts by their employers after the economic downturn of 1873 [@cooper_army_1980, 43-45]. At first, state and local officials tried to respond to these strikes with local law enforcement entities and state National Guard units, but these proved insufficient to handle the strikers, who had not only ceased all train traffic, but in some cases, had engaged in acts of violence [@cooper_army_1980, 43-45; @laurie_role_1997, 30-32]. Despite concerns about involving the Army in what might be perceived as an economic "class war", President Hayes dispatched the Regular Army in the summer of 1877 to maintain law and order [@laurie_role_1997, 32].

The timing of this strike is also notable. Congress had failed to pass an appropriations bill funding the Army past July 1 of 1877, resulting in an Army that was essentially serving without pay. Moreover, the failed appropriations bill was the result of bitter fighting between Southern Democrats in Congress who were angry with an Army that had taken on significant political roles during Reconstruction, and Congressional Republicans who had directed the use of the Army for precisely these controversial tasks. After Congressional Republicans failed to commit to any firm restrictions on the Army's domestic use after the controversial 1876 Presidential Election, democrats "blocked the 1877 appropriations bill" [@clark_preparing_2017, 115]. 

The Army was generally successful in quelling the _Great Strike of 1877_, and many citizens, popular press outlets, and Army leaders perceived the Army as having performed admirably [@cooper_army_1980, 60-61; @laurie_role_1997, 52-55; @skowronek_building_1982; 100-101]. And yet, Army leaders did not make significant public appeals commending or lauding the performance of the troops in the aftermath of these events. The lack of public appeals made by military leaders after the _Great Strike of 1877_ is even more noteworthy considering that the Army's tactical response to the disturbance had been commanded by General Winfield Scott Hancock, who had risen to national prominence, especially among Democrats, after replacing General Phil Sheridan as the 5th Military District Commander in New Orleans in 1867.

Hancock was no stranger to making public appeals. In fact, it was precisely the publication of his _General Orders Number 40_ upon assuming command in New Orleans in 1867 that had "stirred a storm throughout the nation," soured Hancock's relationship with General Grant, and brought Hancock "prominently to the attention of the Democratic Party" [@jordan_winfield_1988, 204]. Moreover, Hancock would go on to obtain the Democratic Presidential nomination in 1880, and thus he was certainly sufficiently well known in 1877. And yet, Hancock did not published an op-ed after the Army's response to the strikes. Given the lack of funding for the Army, the significant reductions in size it had endured, and other concerns about the Army's role in an era of peace, it at least seems reasonable that Hancock had an opportunity to write a public appeal in which he may have called for increased funding, troop strength, or simply gathered public attention for the Army's recent accomplishments.  

Furthermore, the historical evidence shows that Hancock was in fact very proud of how the troops had performed, and particularly the restraint they had shown in bringing the domestic disturbance to an end. In a private letter written to another fellow general, John Schofield, Hancock stated, "the troops have lost the government no prestige...it was the moral force of the United States government that was displayed - not its physical force" [Hancock, 1877 as cited in @cooper_army_1980, 61]. Yet Hancock seems not to have such sentiments to the public in a form of an appeal. 

At a minimum, such an outcome is consistent with the expectation that despite the Army's success in responding to the labor strikes, the broader political environment still remained one of general indifference or apathy towards the Army. In other words, the success of the Army in quelling this labor strike did not significantly change the perceived centrality or sense of importance of the Army. The labor disputes were certainly not minor incidents, but they were not major wars resulting in tens of thousands of casualties or the nation assuming a war footing. If anything, and as the second episode illustrates, the domestic use of the Army in any capacity, especially in the immediate wake of Reconstruction, remained an extremely sensitive topic that, unlike a decade prior, Army leaders were more or less content to let civilian politicians argue over.   

The army appropriations bill that eventually passed Congress in 1878 contained two critical provisions. First, at the behest of Congressional Democrats, the bill instituted the _posse comitatus_ law. In response to what they perceived to be the inappropriate use of the Army throughout the Reconstruction period, Democrats "forbade the use of the Army as a posse to aid local officials in dealing with domestic turmoil without the express order of the President, and only then for very specific purposes" [@cooper_army_1980, 83]. Second, the bill established a bipartisan committee known as the _Burnside Commission_ to examine a number of military reorganization and reform topics, including the Army's size as well as systems involving manning, promotion, retirement, and staffing [@clark_preparing_2017, 116].^[Connelly (2006), Fitzpatrick (2017),and Clark (2017) are among the contemporary historians who write about these particular reforms extensively.]

Historians note that Army leaders remained relatively "quiet" about the passage of the important _posse comitatus_ provision [@connelly_john_2006, 238]. Similar to General Hancock _not_ writing a public appeal after the Army's success in quelling the _Great Strike of 1877_, the silence of Army leaders on this important and critical topic strongly hints at how Army leaders viewed their potential to persuade the public at this particular juncture. Furthermore, this historical record again indicates that at least in private, Army leaders predominantly did not approve of providing open-ended support to civilian authorities, and thus favored the majority of the restrictions on the use of the regular Army contained in the _posse comitatus_ provision [@laurie_role_1997; @connelly_john_2006, 238-239]. 

The fact that senior Army leaders remained silent on this very important and consequential aspect of the appropriations bill, and that the majority of these Army leaders had served the previous decade and had either engaged in making public appeals themselves or been exposed to others who did, strongly suggests that the context had changed, and that in particular, military leaders could not as easily influence public opinion. This is consistent with the evidence showing that in terms of centrality or importance to the nation, the military's position had diminished relative to what it had been immediately at the end of the Civil War. The evidence is also consistent with the theoretical expectation that as military prestige diminishes, there is less of an opportunity to influence public opinion, and that as a result, military and civilian leaders are less likely to engage in public behaviors that violate civil-military relations.     

### Shoulder Tapping and Limited Public Appeals 

It is not as if the Army and its leaders engaged in no political behaviors whatsoever in this latter portion of the case study. Army leaders in fact did engage in political behaviors, but these were far less severe, and rarely violated the principles of civil-military relations. This is most clearly seen in the political behaviors engaged in by Army leaders to proposed military reforms resulting from the _Burnside Commission_, which was established by the 1878 appropriations bill. Chaired by Senator Ambrose Burnside, a former Union General, the commission ultimately recommended a number of Army reforms that were essentially a series of compromises: to placate Republicans, the legislation proposed that the overall size of the Army remain the same; to placate Democrats who largely wished to reduce the Army, the legislation proposed a reduction in the number of Army officers, including the number of generals, as well as a reduction in several types of Army units [@fitzpatrick_emory_2017; @clark_preparing_2017, 118-121; @skowronek_building_1982; 103]. 

The Army did not receive and react to the proposed legislation uniformly. Some leaders, including Generals Sherman and Schofield, cautiously and perhaps somewhat reluctantly approved of the legislation on the grounds that it likely served as the Army's best chance at obtaining any meaningful reforms in the foreseeable future, while others, such as Generals Sheridan and Hancock, did not support the legislation at all [@clark_preparing_2017, 118-121; @jordan_winfield_1988, 254].^[Clark (2017) notes that while Sheridan approved of many aspects of the legislation, including proposed changes to the delegated authorities exercised by the commanding general, Sheridan took issue with the legislation's plans to trim the Army's Cavalry formations.] But the curtailment of the Army's centrality to the nation impacted _how_ military and civilian leaders advocated in support of or against the proposed legislation. Military actors rarely addressed or appealed to the general public about the proposed reforms and instead, either addressed fellow military actors or select politicians directly. Additionally, the Army's senior leaders remained remarkably quiet about the legislation, and did not engage in significant efforts either in support of or against the bill's passage. This is a significant difference from, for example, General Pope's 1867 letter published in _The New York Times_ and referenced in the introduction of this chapter, which blatantly argued against President Johnson's stated policies on Reconstruction. 

For starters, a number of military officers wrote letters in response to the _Burnside Commission_ asking for their opinions on the proposed changes. However, these letters should not be construed as inappropriate, as they were written in response to official inquiries made by the commission. Some officers, however, engaged in more overt forms of behavior. For instance, Fitzpatrick (2006) notes that several staff bureau officers, who in particular feared that the legislation would reduce their ability to influence Congress, produced and distributed informational pamphlets to members of Congress [@fitzpatrick_emory_2017, 221]. 

Military officers handing out informational pamphlets in order to persuade members of congress to denounce the legislation in question amount to acts of "shoulder tapping" and "alliance building", behaviors which were outlined in chapter two of this dissertation [@brooks_militaries_2009]. Making and distributing pamphlets - unlike writing a response to an inquiry made by Congress - violated the principles of civilian control in that they caused the _Burnside Commission_ to contend with a set of military officers who were functioning as a disparate interest group. Burnside himself was frustrated that he had to contend with the staff officers advocacy efforts, and noted, "I must say that some of these staff officers have gone beyond the line of duty, particularly...in Washington" [Burnside, 1878 as cited in @fitzpatrick_emory_2017, 221]. 

In addition to these acts of "shoulder tapping", military actors also made a number of semi-public appeals regarding the proposed reforms. However, these appeals differed in two important ways from the leaks and appeals made by Army leaders in 1865 - 1868. First, rather than publishing op-eds in major US newspapers such as _The New York Times_, military leaders aired their views in military periodicals such as the _Army and Navy Journal_, a periodical primarily read and consumed by military actors, to include soldiers, sailors, and veterans. And second, the content of what came to be published did not firmly support or denounce the proposed legislation, but rather commented on various aspects of the legislation. General Sherman, for instance, provided letters to the _Army and Navy Journal_ written by several leaders, including Generals Schofield, Sheridan, Pope, and Lieutenant Colonel Emory Upton [@connelly_john_2006, 243]. Some of these officers supported, and some opposed, the recommended reforms. This suggests that in providing these letters to the _Army and Navy Journal_, Sherman was not leaking information to score partisan political points, but rather to inform military audiences of the content of the legislation. This seems the most convincing explanation given Sherman's ambivalent disposition regarding the Burnside Commission's proposals, and marks a noticeable difference in behavior from the leaking and threat of leaking information between President Johnson and Grant that that marked the President's final days in office in 1868.^[Marszalek (2007, p. 435) quotes Sherman as having stated, "I shall not advocate the Bill or oppose it, but shall conform my action, official and personal, to the conclusion of the Law Making Powers of Our Government."] 

### Civilian Political Behavior, 1877 - 1880

It is also important to examine the behavior of civilian actors during this latter portion of the case study. If in fact the prestige of the military had diminished relative to the level of military prestige that had existed at the end of the Civil War, the theory predicts that civilian leaders especially would have been less inclined to use the military to influence public opinion relative to their attempts to harness military figures for political advantage from 1865-1868. Does this central prediction fit the historical events that occurred from 1877 - 1880? 

In one sense, the theoretical prediction holds, but in another sense, it does not, at least not at first glance. The theoretical prediction holds in that civilian leaders did not seek to leverage the military for their own partisan gain in ways that were as obvious as they did from 1865-1868, nor did these behaviors take on as severe a form of behavior as was evident in the prior period explored in this case study. 

The historical record, for instance, does not indicate that any sort of parallel behavior to Johnson's "Swing Around the Circle Tour," in which the President had several currently serving senior military leaders accompany him to several American cities for the sole purpose of bolstering his party's chances in the 1866 midterm elections, occurred under President Hayes. The memoirs of Sherman and other military leaders who occupied senior positions in the late 1870s do not show evidence that politicians shaped their public appearances so as to literally stand next to military leaders in public forums for the purposes of harnessing the popularity held by these military figures, as President Johnson did vis-a-vis General Grant at the White House in 1866 and discussed previously in this chapter. 

This is not to say that civilian actors never engaged in political behaviors with respect to the military from 1877-1880. They always do. The question is what kind of character these political behaviors take. For instance, Representative James Garfield, who later became President and who had served as a Union General during the Civil War, wrote a two part series in _The North American Review_ in early 1878 in which he advocated that Congress take seriously the topic of Army reform. In these publications, Garfield included a portion of, or in some cases, the entirety of, a number of letters written by senior Army leaders addressing the size of the peacetime Army, the role of the staff vis-a-vis the commanding general and the Secretary of War, and other pertinent items. 

Garfield's articles, however, do not read as a blatantly partisan document. His articles rather function as a sort of alarm bell for the nation, and urge the American people (and Congress in particular) to treat the Army seriously in spite of the very real transition from war to peace that had occurred since the end of the Civil War [@garfield_army_1878; @garfield_army_1878-1]. Thus, the theory holds in that civilian policymakers from 1877 - 1880 for the most part engaged in political behaviors that were consistent with the routine duties of political actors, and in ways that did not violate normative principles of civil-military conduct. 
The theory holds less well with respect to the specific political behavior of military figures running for office, and in particular, the Presidency. To see this, consider that during the seven Presidential Elections held from 1868 to 1892, only the election between Grover Cleveland and James Blaine in 1884 featured two non-veteran candidates [@teigen_why_2018, 104-106]. All other elections featured at least one veteran as one of the party's major final candidates. These candidates running for President in elections held between 1868 and 1892 (except for 1884) are displayed in Table \@ref(tab:ch5-candidates). 

```{r ch5-candidates, include=T, echo=F, error=F, message=F}

ch5_text_tbl2 <- tribble(~"Election Year", ~"Candidate", ~"Party", ~"Veteran Type",
  "1868", "Ulysses S. Grant", "Republican", "Career General",
  "1872", "Ulysses S. Grant", "Republican", "Career General",
  "1876", "Rutherford Hayes", "Republican", "Battlefield General",
  "1880", "Winfield Scott Hancock (and James Garfield)", "Democrat (Republican)", "Career General (Battlefield General)",
  "1888", "Benjamin Harrison", "Republican", "Battlefield General",
  "1892", "Benjamin Harrison", "Republican", "Battlefield General"
  )
 
ch5_text_tbl2 %>% kable(booktabs=T, caption="Presidential Candidates, Post Civil-War Era") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="10em") %>% 
  column_spec(2, width="12em") %>% 
  column_spec(3, width="8em") %>% 
  column_spec(4, width="14em") %>% add_footnote(label = "Reproduced from Table 5.1 in Jeremy Teigen's (2018) book, Why Veterans Run (p. 105).")
```

This then begs the question of how military figures could become the final candidate of a major political party if military prestige truly declined in the decade after the Civil War. Specifically, we have to ask the question, if military prestige truly declined after the Civil War, how did General Winfield Scott Hancock still manage to obtain the Democratic nomination in 1880? Does this historical fact cast doubt on what has been argued here - that military prestige truly did decline relative to 1865 by 1880 - and if it does, what are the implications for this dissertation?

One plausible way of reconciling what appears at first glance to be a contradiction involves a theoretical relationship between individual and institutional dimensions of prestige. The theory that I offered in chapter two of this dissertation hinted at these distinctions by drawing on the two separate but related components of prestige involving first, the centrality or the importance of a profession or occupation, and second, the skill or excellence of the members who serve in the profession or occupation. These separate but related components were first theorized in the sociological literature regarding the stratification of occupations [@davis_conceptual_1942; @jaco_prestige_1970]. 

In this vein, it as at least plausible to interpret Hancock's obtaining the Democratic Party nomination in 1880 as evidence that under certain conditions, individuals can retain a high level of personal prestige even when the profession of which they are a part operates at a relatively lower level of prestige. Applying this dynamic to the two dimensions of prestige would suggest that some military leaders who served in the late 1870s were considered by society to have held a high level of skill or excellence, the second component of prestige. This certainly fits, conceptually, with the understanding that many of these military leaders serving in the late 1870s had achieved heroic status as a result of their martial exploits during the Civil War. At the same time, though, these senior military leaders were in fact serving in a profession that, when viewed through the lens of centrality, the first component of prestige, was no longer as central to the life of the nation compared to a decade prior. 

While future research is required to further unpack the institutional versus the personal dimensions of military prestige as a variable, such a dynamic is at least consistent with a very plausible explanation for the selection of Hancock as the Democratic Party nominee in 1880. Hancock, Sherman, Sheridan, Schofield, and other generals were in fact, by the late 1870s, serving in an Army that was more than a decade removed from the conclusion of the Civil War. The nation was no longer on a martial footing in 1877-1880 in the way that it had been during the Civil War and in the early period of Reconstruction. While this change in contexts certainly seems to have produced less extreme political behaviors involving the military, there were still signs that for some occasions, civilian actors were not dissuaded from harnessing every last ounce of military prestige available. 

## Summary of the 1877-1880 Period

In the decade after the Civil War, multiple factors combined to render the Army less prestigious than it had been at the close of the Civil War. The political nature of Reconstruction had been unpopular with many Americans, and particularly with Southerners. As the nation reunified, the Army became a perennial target by Democrats who sought to trim the Army's size and proscribe the terms of its domestic involvement. In the absence of a clear international threat to prepare against, the Army was largely at the mercy of a divided Congress that was not convinces to take up a number of reforms in the late 1870s. 

The Army's reduced centrality to the nation in 1877-1880 greatly impacted the character of the political behaviors military and civilian leaders undertook compared to those undertaken immediately after the Civil War ended. Unlike the political behaviors undertaken by General Grant and President Johnson, which included acts of explicit insubordination (Grant), partisan stacking (Johnson), and the routine threat of leaking information (both actors), the political behaviors undertaken between 1877-1880 by military and civilian actors were far less extreme. 

The political behaviors that did occur involving the military from 1877 - 1880 included relatively normal acts such as shoulder tapping (some military actors) and public appeals that were by and large not explicitly partisan in tone (military and civilian). Furthermore, military leaders, including those who had made public appeals previously in their careers from 1865-1868, did not make public appeals and instead to stay silent after several important events, especially following the Army's tactical success in quelling the first of many domestic labor disputes in the late 19th Century, _The Great Railway Strike of 1877_. 

The political behaviors engaged in by civilian and military actors from 1877-1880 are displayed in Table \@ref(tab:pol-behav-cs1-part2).

```{r pol-behav-cs1-part2, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_tbl3 <- tribble(~"Actor", ~"Tactic of Political Behavior", ~"Date", ~"Instance", ~"Normative Standard Violated",
  "Military Staff Officers", "Shoulder Tapping", "1878", "Staff officers distribute informational pamphlets to members of Congress arguing against adopting a number of provisions proposed by the Burnside Commission", "Civilian Control",
  "General William Sherman", "Public Appeals", "1878", "General Sherman sent letters written by fellow military commanders to be published by The Army and Navy Journal regarding various reforms under consideration by the Burnside Commission", "No principles violated; the public appeal was geared to the military community and conducted for the purposes of informing rather than influencing",
  "Representative James Garfield (Civilian)", "Public Appeal", "1878", "Garfield publishes two articles in The North American Review on the topic of Army reform in early 1878", "No principles violated; the articles were written in general terms and did not seek to leverage the Army for personal political or partisan gain",
  "General Winfield Scott Hancock", "Running for Office", "1880", "General Hancock runs for political candidacy even as he is an active member of the Army in 1880", "Non-partisanship"
  )
 

text_tbl3 %>% kable(booktabs=T, caption="Political Behaviors Involving the Military, 1877-1881") %>%
  kable_styling(full_width = F, font_size=8, latex_options="striped" ) %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2, width="8em") %>% 
  column_spec(3, width="4em") %>% 
  column_spec(4, width="14em") %>% 
  column_spec(5, width="8em")
  
```

## Alternate Explanations

Besides the relative decline in the importance of the Army to the nation in the years following the Civil War, a number of other factors might explain the changes in how military and civilian leaders engaged in political activity involving the military. One possible alternate explanation centers on individual personality differences between senior army leaders between 1865 and 1877. There may be some merit to this explanation. Grant was no lover of politics as a General, but it is also well known among historians that Sherman and Grant approached politics rather differently. Marszalek (2007) for instance notes that, "...politics was anathema to Sherman. He viewed politicians as venal adversaries he was forced to tolerate but refused to work with" [@marszalek_sherman_2007, 425].

But in my view, personality differences alone are not sufficient in explaining why Sherman behaved so differently in the late 1870s than Grant did in the late 1860s. Even if Sherman was naturally inclined against engaging in openly political behaviors, the historical literature clearly indicates that Sherman strongly lamented the fact that he was presiding over the Army in an era in which the Army was no longer considered vital to the country. 

Another potential alternate explanation involves the change in the conflict or threat environment that occurred between the period of 1865-1868 and 1877-1880. In my view, there is merit to this explanation, but we have to carefully think _how_ the change in the threat environment impacted the variables that are part of this dissertation's theory. 

It is factually incorrect to simply state that Lee's surrender in 1865 ushered in an era of instant peace, and that the change in political behaviors that occurred by the late 1870s simply reflects this change from war to peace. The reason such a view is incorrect is, as this case study has shown, the Reconstruction period involved a significant military undertaking - an occupation - of the southern states, and involved military forces and their commanders exercising a number of incredible wartime authorities throughout this period [@downs_after_2015]. Additionally, over the nearly twelve year period of Reconstruction, the nation's political circumstances also changed as the nation reunited, but the tasks associated with Reconstruction had rendered the Army less popular, especially in the southern states. 

It would be far more accurate to conclude that a combination of factors, to include the significant reduction in the size of the Army, the lack of any clear international threat for the Army to prepare against, and a decline in the perception of the Army as a result of the types of tasks it carried out during Reconstruction itself, rendered the Army less central to the life of the nation in 1880 than it had been in 1865. In other words, I acknowledge that the change in the threat environment that occurred over the course of 1865-1880 was a factor that impacted the political behavior civilian and military leaders undertook by the end of the same period, but contend that this change in the threat environment nonetheless operated through the wider variable of military prestige. In this way, I see no contradiction between claiming that a reduction in military prestige _and_ a change in the threat environment contributed to the change in political behaviors undertaken by civilian and military actors between 1865-1868 and 1877-1880. 

A final alternate explanation is to conclude that what changed in this case study was not the level of military prestige, but rather the presence of a political crisis. In this vein, the years 1865 - 1868 constituted a political crisis that compelled Grant to act the way that he did. Furthermore, through such a lens, the level of military prestige did not significantly diminish between 1868 and 1877. What really happened is that the political crisis of the early years of Reconstruction were eventually replaced by a more normal type of 19th Century politics.

Admittedly, there is much about this view that I find accurate and true. The years from 1865 - 1868 were indeed atypical years, and they certainly resulted in more than one crisis, including one of the few impeachments of a sitting US President that has occurred in US history. The difficulty associated with this alternate view arises from the difficulty in disentangling these years of political crisis from the high level of polarization that existed at the same time. The empirical record shows that political polarization was high throughout the late 19th Century, which leads us to infer that polarization did not change over the period examined here. At the very best, then, we have to ask how or why a political crisis emerged during the years 1865 - 1868 to begin with. 

As I have attempted to argue here, these years were unique in the nation's history that coincided with the end of the Civil War and the immediate task of restoring the former confederate states to the Union. Military leaders were incredibly popular and well known in these years, and ultimately were given the task of occupying the south as part of Congress's plan for Reconstruction, a plan that ran counter to that of the President's. In this vein, the crisis years of 1865 - 1868 were made more acute than they would have been if the military and its leaders would not have held the degree of prestige that they did.   

# Conclusion

This chapter finds moderate support for H4, which stated that military and civilian actors engage in relatively more visible and public forms of political behavior that involve the military institution when military prestige is high relative to periods of low military prestige. It also finds strong support for H5, which stated that the most extreme levels of overall political activity occur in environments that are characterized by simultaneously high levels of political polarization and military prestige. 

The strongest evidence for these claims arises from the severity of political behaviors that occurred in the United States between 1865-1868, and the actors who committed them. At times, during these years of crisis, President Johnson and General Grant, the nation's chief executive and the nation's top military officer, openly and blatantly challenged what even by historical standards constituted the most basic and fundamental norms of civil-military conduct. Both actors did so because of the interplay of exceedingly high levels of political polarization and military prestige. The nation was in the early process of recovering from a bitter and destructive Civil War, and making consequential and weighty decisions related to Reconstruction that touched on a host of significant moral-ethical factors. At the same time, the level of prestige afforded to senior military leaders, and to General Grant in particular, was incredibly high, especially in the immediate aftermath of the Civil War. 

By the late 1870s, even as polarization remained high, the character of the political behaviors undertaken by civilian and military actors lost their extreme flavor, and rarely violated standards of civil-military conduct. Senior military leaders serving at the time, including Generals Hancock and Pope, both of whom had had engaged in obvious forms of political behavior in the years from 1865-1868, were cognizant that they now served in an Army that was no longer viewed as vital to the life of the nation. The Army had downsized significantly, largely because of the fact that few if any international threats existed. Accordingly, the military and its leaders were not often the target of obvious schemes to influence public opinion. Simply put, the military and its leaders from 1877-1880 did not offer as strong of a political advantage to civilian actors as they had in the period from 1865-1868. 

This dissertation now moves to a second case study that examines an era in which the level of polarization predominantly changes. In order to do this, we now jump forward roughly 140 years and examine political behaviors conducted by civilian and military actors during the Iraq War. 
\pagebreak