
# Chapter Six {-}

On September 11, 2001, terrorists hijacked several commercial airplanes and purposely crashed them into several sites, including the twin towers of the World Trade Center in New York City, the Pentagon in Washington D.C., and a field in Western Pennsylvania.^[Heroic Americans prevented this plane from reaching its intended target, the US Capitol or the White House, in Washington D.C. [@kean_911_2004, 14].] The evening of the 9/11 attacks, members of Congress gathered on the steps of the US capitol to express solidarity and resolve. After a moment of silence and short remarks by then speaker of the House Dennis Hastert, those gathered sang the song, "God Bless America" [@us_congress_members_2001].

Seventeen months later, in February 2003, the Senate Armed Services Committee held a hearing regarding a likely US invasion of Iraq, which would begin the following month. The committee was interested in hearing the views of the respective military service chiefs regarding plans for the upcoming war. One exchange was particularly noteworthy. General Eric Shinseki, the Army Chief of Staff, told the committee that he thought "something on the order of several hundred thousand..." was a likely figure for the number of troops required to stabilize Iraq after the invasion [Shinseki, 2003 quoted in @fallows_karmic_2008]. Shinseki's remarks, however, went against the official position held by the Defense Department and it's civilian head, Donald Rumsfeld, and his second in command, Paul Wolfowitz, both of whom believed that a smaller, lighter military footprint was more than adequate for the pending invasion and its aftermath. In the days that followed, both Wolfowitz and Rumsfeld publicly criticized Shinseki's remarks [@schmitt_pentagon_2003]. Shinseki, for his part, remained largely silent, and retired from the Army shortly afterward.  

Eighteen and a half years later, the United States withdrew from Afghanistan after the resurgent Taliban rapidly displaced the Afghan National Government in the summer of 2021. During a Congressional hearing held on 28 September 2021 about the US withdrawal, Tom Cotton, a veteran and the junior Republican Senator from Arkansas, asked the Chairman of the Joint Chiefs of Staff, General Mark Milley, "why haven't you resigned [@singman_milley_2021]?" The Senator's question alluded to the fact that General Milley, weeks before the collapse of the Afghan Government, had recommended to President Biden that he maintain approximately 2,500 US Troops in Afghanistan, a recommendation the President did not heed. General Milley told Cotton that, "It would be an incredible act of political defiance for a commissioned officer to just resign because my advice is not taken...this country doesn't want generals figuring out what orders we are going to accept and do or not. That's not our job [@singman_milley_2021]." Furthermore, in the same hearing, before ever even discussing the issue of Afghanistan, General Milley addressed allegations that he had inappropriately conversed with both a Chinese general and with the Speaker of the House, Representative Nancy Pelosi, during the transition between President Trump and Biden [@baldor_watch_2021]. These allegations had provoked strong partisan reactions in the weeks leading up to the hearing. 

Cotton's question by itself was not necessarily inappropriate, but it is highly likely that Cotton wished that Milley would either have resigned or at least intimated to the public that he considered resigning as a result of the chaotic withdrawal from Afghanistan. Such a signal from the sitting Chairman of the Joint Chiefs of Staff would have generated significant political points for Congressional Republicans, and likely have caused a fairly significant rift in civil-military relations. 

This chapter identifies, compares, and contrasts the patterns of political behavior engaged in by military and civilian actors in the post-9/11 era. From approximately 2001 - 2021, the level of military prestige remained relatively high, but the level of political polarization rose sharply. The chapter proceeds in four parts. First, I describe the predictions concerning the patterns of political behavior as a state moves from a domestic environment marked by low polarization and high military prestige to one marked by high polarization and high military prestige. In this section, I also substantiate the claim that such a movement approximates the change in the domestic environment that occurred in the United States from 2001 - 2021. 

In the second part of this chapter, I describe several political behaviors associated with one significant civil-military relations episode that occurred during the first half of this nearly twenty year period. The episode that I focus on is the so-called 2006 "Revolt of the Generals," which involved several retired senior military officers publicly castigating the Secretary of Defense, Donald Rumsfeld. The third part of the chapter explores an episode that occurred during the second half of the period. I focus on several political behaviors involving the military undertaken in the summer and fall of 2020, which involved periods of domestic civil unrest and the turbulent 2020 Presidential Election. The fourth and final part of this case study evaluates, compares, and contrasts these two episodes. 

Similar to the previous chapter, the central focus is on the tenor, characteristics, tone, and underlying reasons for the occurrence of the examined political behaviors. The chief argument made in this chapter is that, in accordance with this dissertation's theoretical predictions, the rise in the level of political polarization from 2001 - 2021 contributed to noteworthy changes in _how_ civilian and military actors politicize the military.

# Theoretical Predictions for the Post-9/11 Era, 2001-2021

This section describes the predictions regarding the patterns of political behavior involving the military as a state's domestic environment changes. In particular, I will describe how the political behaviors change as a state's domestic environment shifts from one marked by relatively low polarization and high prestige to one characterized by high polarization and high military prestige. 

## Low Polarization and High Military Prestige

The theory presented in chapter two of this dissertation argued that a state's domestic environment strongly shapes the patterns of political behavior that occur involving the military. When the level of military prestige is high, both civilian and military actors are more likely to believe that the military can influence public opinion relative to when the level of military prestige is low. Furthermore, the theory argued that both types of actors will engage in behaviors that seek to harness or leverage the prestige of the military in order to appeal to the public when military prestige is high, even if doing so violates one or more of the principles of civil-military relations. 

Chapter two noted that this theoretical expectation is rooted in the political communication literature, which argues that the public listens to voices who they consider to be "knowledgeable" and "trustworthy" [@lupia_democratic_1998, 69-76; see also @zaller_nature_1992, 47]. The theory argued that as the level of military prestige rises, the public increasingly considers the military and its members to possess these traits. 

When the level of polarization is low, the theory also predicts that there will be relatively little basis for any substantive ideological divisions between and among military actors. This is because periods of low political polarization are associated with few if any differences that touch on issues of moral values or worldviews between and among military actors. Importantly, this is not to say that military actors are entirely unified or that there is little chance of conflict between military actors when polarization is low. Rather, the theory predicts that absent high levels of polarization, there is little if any ideological basis, or _motive_, for military actors to behave in ways that violate the central principles of civil-military relations.

In an environment marked by low polarization and high military prestige, the theory ultimately predicts that either military or civilian actors (or both) may induce politicization of the military, and engage in behaviors that violate the principles of civil-military relations. The theory even noted that such an environment is far from tranquil, and that there is a high potential for civil-military conflict altogether, as shown in the bottom left quadrant of Table \@ref(tab:poss-environment3). 

```{r poss-environment3, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env_revised2 <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "No politicization; low levels of political activity; low potential for civil-military conflict", 
  "Military (induced) politicization; moderate levels of political activity; moderate potential for civil-military conflict",  
  "High Prestige", "Military and Civilian (induced) politicization; moderate levels of political activity; high potential for civil-military conflict", 
  "Comprehensive politicization; extreme levels of political activity; significant potential for civil-military conflict" 
  )
 

  kable(text_env_revised2, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors and the Conduct of Civil-Military Relations") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2:3, width="18em") 
 
```

Throughout US history, there have been several significant civil-military relations disputes that have occurred under the domestic conditions of low polarization and high military prestige. For example, disputes between General Douglas MacArthur and President Harry Truman over Korea, General Matthew Ridgeway and President Dwight Eisenhower over Ike's New Deal policies in the early 1950s, and Colin Powell and President Bill Clinton over several post-Cold War crises in the early 1990s, all occurred in domestic environments marked by relatively low polarization and high military prestige. Therefore, we should not conclude that the conduct of civil-military relations is free of tension or absent the potential for conflict between civilian and military leaders when when polarization is low and when military prestige is high. 

## High Polarization and High Military Prestige

The theory does predict, however, that as polarization rises, the potential for civil-military conflict rises and that military actors in particular are even more likely to engage in behaviors that violate the principles of civil-military relations. These predictions are so because rising political polarization is associated with increasing ideological division occurring within and among military actors who, like the rest of society, become more polarized as the level of polarization grows. 

Rising levels of polarization attest to a loss of consensus or agreement concerning the moral values and ideals that should hold sway in society [@marsden_twilight_2014; @mann_its_2012; @allen_why_2020]. In highly polarizing times such as the contemporary age in the United States, this dissertation's theory noted that the conflicts that emerge are not simply political. That is, many contemporary polarizing conflicts do not simply involve questions concerning whether society should enact policy "X" or policy "Y", but rather involve conflicts over the worldviews and moral values that underlie policy "X" or policy "Y". In making these claims, chapter two highlighted several polarizing issues in the contemporary United States, including the question of what it means to be an American, whether America should be proud of its history, and whether the construct of gender is important anymore. What these and many other polarizing issues have in common is that they cannot be solved nor even discussed without an individual referencing or alluding to some sort of moral-worldview commitment. 

Furthermore, chapter two argued that the reason the level of polarization impacts the conduct of civil-military relations is because a military actor's commitment to the worldviews or values underlying any number of polarizing issues may in fact be stronger than that same actor's level of commitment to the central principles of civil-military relations. An officer in the United States military likely feels some commitment or bond, perhaps even a very strong bond, to the principles of civilian control, non-partisanship of the military institution, and non-interference. But the same officer likely also has a worldview that relies on moral values or commitments of some kind, even if that officer cannot articulate what the worldview is [@sire_universe_2020]. Chapter two argued that in highly polarizing environments, military actors more frequently encounter situations in which they have to choose between adhering to the principles of civil-military relations or advancing the values underlying any number of polarizing issues that are at stake in the political sphere. When confronted with such a decision, especially when the disparity in resulting outcomes is great, military actors will increasingly side with the moral values associated with their worldview, and thus, behave in ways that violate the principles of civil-military relations.   

As Table \@ref(tab:poss-environment3) notes, high polarization and high military prestige thus leads to an environment in which there is comprehensive politicization, extreme levels of political activity involving the military, and significant potential for civil-military conflict. These characteristics are even more severe than those that exist in a domestic environment marked by low polarization and high military prestige. 

## Substantiation of Environmental Characteristics, 2001 - 2021

To substantiate the claim that the level of polarization increased and the level of military prestige generally remained constant (but relatively high) over the period of 2001 - 2021, I rely on measurable quantitative data. This is different from the previous chapter that explored the post-US Civil War era, which relied on mainly qualitative data. Figure \@ref(case2-polar) plots the level of polarization in both the US House and the US Senate, as well as the level of trust and confidence in the US military, from 2001 - 2021. Data for polarization uses roll-call voting data on members of Congress over time [@jeffrey_b_lewis_voteview_2020], and data for trust in the military comes from Gallup public opinion data regarding the level of public trust and confidence in the military [@noauthor_confidence_2020].  

```{r case2-polar-prep, include=FALSE, warning=FALSE, echo=FALSE}

df_trust_polar_cs2 <- read_csv(here("data","polar-inst-trust-time.csv"))

df_trust_polar_cs2 <- df_trust_polar_cs2 %>% filter(Year >= 2001) %>% select(-Congress) %>% 
  mutate(
    house_polar_dim1 = house_polar_dim1 *100,
    sen_polar_dim1 = sen_polar_dim1 *100,
    ) %>% pivot_longer(Military:sen_polar_dim1, names_to = "value", values_to = "Score") %>% distinct()
```

```{r case2-polar, include=T, warning=FALSE, echo=F, message=F, fig.height=4, fig.width=6, fig.cap = "Public Trust (L) and Polarization (R), 2001-2021" }
df_trust_polar_cs2 %>% ggplot(aes(x=Year, y=Score, color=value)) + geom_line() + geom_point(size=0.7) + labs(x="Year", y="Trust in Institutions") + scale_color_discrete(name="Trust/\nPolarization",
                         breaks=c("Military", "house_polar_dim1", "sen_polar_dim1"),
                         labels=c("Trust in\nMilitary","House\nPolarization", "Senate\nPolarization")) + scale_y_continuous(sec.axis = sec_axis(~., name = "Polarization (Scaled)")) + scale_x_continuous(breaks=seq(2001, 2021, 2)) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE))
```

Figure \@ref(case2-polar) shows that each variable, polarization and military prestige, has changed differently over the period from 2001-2021. With the exception of a select few years, polarization levels in both the US House and Senate have consistently increased over time. 

The level of public trust in the military, on the other hand, has exhibited more variation between years. For example, a relative spike in public trust occurred in 2003, perhaps reflecting a boost in patriotic fervor as the war in Iraq started the same year. Similarly, local minima occurred in 2007, perhaps coinciding with impatience over mounting US casualties in Iraq, and again in 2021, perhaps reflecting public frustration over the US withdrawal from Afghanistan. Yet even in the presence of this variability, the level of military prestige that has endured for the past twenty years is relatively high, and certainly higher than the level that existed in the late 1970s and early 1980s after the Vietnam War. Since data has been kept, for example, the lowest recorded level of trust in the military occurred in 1981, when only 50% of surveyed Americans said they trusted the military a "great deal or quite a lot" [@noauthor_confidence_2020]. In short, the period 2001 - 2021 reflects a domestic environment in which the level of military prestige remains relatively high, but in which the level of polarization increases. 

Note, however, that Figure \@ref(case2-polar) does not, and indeed cannot, definitively conclude when the shift from a relatively moderate to relatively high level of polarization occurred. What we do see, however, is a fairly noteworthy spike in the level of polarization in the US House (red line) beginning in 2011, and that this spike was accompanied by a smaller yet still detectable spike in polarization in the US Senate (blue line) the same year. Furthermore, we know that this occurred midway through the first term of President Barack Obama. 

From a methodological standpoint, then, the analysis which follows takes the year 2011 as a sort of natural breaking point. I begin by describing the 2006 "Revolt of the Generals," which occurred several years before 2011. I then describe a number of political behaviors undertaken by military and civilian actors involving the military in 2020. Finally, I compare and contrast these behaviors, and argue that the behaviors undertaken in 2020 reflect the rise in the level of polarization that occurred in the United States since the 2006 "Revolt of the Generals." 

\singlespace
# Part 1 - The War in Iraq and The Revolt of the Generals, 2004 - 2006
\doublespace

In the first section of this case study, I describe the 2006 "Revolt of the Generals." After describing the political behaviors that occurred as part of this episode, I clarify the reasons why this episode occurred in the first place, and distill the underlying issues that were ultimately at stake in this episode. While this dissertation certainly speaks to the normative implications of the "Revolt of the Generals" with respect to civil-military conduct, my primary task here is not to render a definitive evaluation of the conduct that occurred as part of this episode. Others [@cook_revolt_2008; @snider_strategic_2014] have written extensively on this. Rather, my primary task is to show how the relatively low level of polarization and a relatively high level of military prestige enabled the "Revolt of the Generals" to occur. 

## Background

The 2006 "Revolt of the Generals" narrowly refers to the behavior of several recently retired senior military officers who, in the Spring of 2006, publicly criticized the Secretary of Defense, Donald Rumsfeld, concerning Rumsfeld's leadership of the war in Iraq. By 2006, the war in Iraq was three years old, and more importantly, it was clear that any hopes for a quick and relatively easy war had been dashed. 

In a broader sense, the "Revolt of the Generals" should be viewed as the military's backlash to their perception of not being heard, listened to, or valued by their civilian chain of command within the Department of Defense [@snider_strategic_2014; @binkley_revisiting_2020]. In a very real way, the "Revolt of the Generals" signified a reaction by several retired military officers to the way that civilians had disparaged General Shinseki for presenting Congress with his views on force levels required to stabilize Iraq (as noted in this chapter's introduction).  

As is always the case, the timing of political behaviors undertaken matters, and those associated with the "Revolt of the Generals" are no exception. Not only was the war in Iraq three years old in the Spring of 2006, but also, later that same year, Congressional midterm elections were held. The behaviors conducted as part of the "Revolt of the Generals" ultimately elevated the status of the war in Iraq, and in all likelihood, partially contributed to the eventual dismissal of Rumsfeld by the President, George W. Bush.   

## Political Behaviors Undertaken

The immediate slate of military actors who participated in the "Revolt of the Generals" include a total of six retired military officers, including Army Major General Paul Eaton, Marine Lieutenant General Greg Newbold, Army Major General John Batiste, Marine Lieutenant General Paul Van Riper, Army Major General John Riggs, and Army Major General Charles Swannack. In chapter three of this dissertation, which examined public opinion commentary authored by retired military actors, I discussed the behavior of retired military officers in detail, and so I will not readdress those considerations here. But, it should be noted that these six retired military officers were very recently retired. For example, Eaton had been responsible for training the Iraqi military from 2003 - 2004, and both Swannack and Batiste had served as Division Commanders on the ground in Iraq [@cloud_more_2006].

I also describe a number of political behaviors undertaken by other military and civilian actors in response to the behavior of these six primary military actors. By considering the behaviors undertaken both as part of the "Revolt" _and_ those undertaken directly in response to it, the reader is better positioned to understand the full impacts of the episode. 

### Public Appeals

The primary political behavior undertaken by military actors during the "Revolt of the Generals" was that of making public appeals. These took various forms. For example, in March and April of 2006, Generals Eaton, Batiste, and Newbold penned articles that appeared in _The New York Times_, _The Washington Post_, and _Time_ magazine, respectively [@eaton_top-down_2006; @batiste_case_2006; @newbold_why_2006]. These public appeals levied multiple criticisms of Rumsfeld, including that he had micromanaged the war effort, failed to heed the advice of senior military leaders during the planning and execution of the war, and insisted on implementing a strategy that was bound for failure. Shortly afterward, Generals Van Riper, Riggs, and Swannack gave interviews in which they "concurred with the criticisms levied by Eaton, Newbold, and Batiste" [@binkley_revisiting_2020, 23].

In response to the "Revolt of the Generals," some military actors then defended Rumsfeld. And while Rumsfeld's defenders did not necessarily constitute a large number of military actors, they too made public appeals. For instance, four retired generals - Vallely, McInerney, Crosby, and Moore - penned an op-ed in _The Wall Street Journal_ in April of 2006 entitled, "In Defense of Donald Rumsfeld" [@john_crosby_defense_2006]. 

### Grandstanding

While public appeals were perhaps the most obvious form of political behavior involved in the "Revolt of the Generals," the episode also indirectly touched on the behavior of grandstanding. In chapter two, I noted that Risa Brooks has defined grandstanding as the "threat or actual resignation in protest of policy" [@brooks_militaries_2009, 219]. To my knowledge, no military officer publicly resigned or publicly threatened to resign as a direct part of or in direct response to the "Revolt of the Generals." However, the entire episode sparked somewhat of a debate among civil-military relations scholars as well as among military officers regarding the behavior of military officer resignation and the conditions under which such a behavior should be considered. Furthermore, there were strong indications that several of the participating generals had resigned from active military service - albeit quietly and without much public fanfare - in order to then publicly criticize the Secretary of Defense. For example, strong rumors circulated that Marine Lieutenant General Newbold had resigned despite being considered a strong candidate for the position of Marine Corps Commandant [@kaplan_revolt_2006]. Similarly, Army Division Commanders are not usually positions from which officers retire directly, as Generals Swannack and Batiste did. 

Here, I do not take up the contours of the resignation debate. Chapter two discussed resignation in greater detail, and emphasized that even among scholars, there is a lack of consensus about if and when resignation should be considered a viable option by military officers.^[For example, see @burke_responsible_2009; @dubik_principled_2014; @snider_strategic_2014; and @milburn_breaking_2010 for general arguments in favor of military officers exercising dissent under strict conditions. See also, however, @m_shields_introduction_2017 and @golby_beyond_2015 for scholars who present alternate views.] However, it is important to acknowledge the debate over resignation that occurred in the wake of the "Revolt of the Generals" because the episode pointed back to the instance involving General Shinseki, which was referenced in this chapter's introduction. Those who participated in or supported the "Revolt of the Generals" argued that General Shinseki would have been justified to have resigned in protest in 2003 ahead of the Iraq War, rather than acquiesce begrudgingly to a strategic plan that he believed was bound for failure.

Finally, it should be noted that the "Revolt of the Generals" was unique from other civil-military disputes in US history in that it essentially marked a coordinated effort by multiple retired generals to oust the Secretary of Defense. All of the revolting generals called on Rumsfeld to step down from his role. After naming Rumsfeld explicitly, Batiste, for instance, argued that, "To move forward, we need a leader with the character and skills necessary to lead" [@batiste_case_2006]. Newbold, for his part, stated that, "We need fresh ideas and fresh faces. That means, as a first step, replacing Rumsfeld and many others unwilling to fundamentally change their approach" [@newbold_why_2006]. Eaton also stated, perhaps more bluntly, that "Rumseld must go" [@eaton_top-down_2006]. Though Rumsfeld did not heed such requests, this aspect - military actors explicitly calling for their former civilian boss to depart - marked the "Revolt" as different from previous civil-military disputes that had occurred in American history.  

### Impact of Undertaken Behaviors

The entire "Revolt of the Generals" episode, and the political behaviors undertaken by the military actors involved in it, were fairly visible to the American public. Especially in a midterm election year, these behaviors likely contributed to shaping the views of Americans on the war in Iraq. 

Rumsfeld went on to remain at his post until one week after the mid-term elections in which the Republicans incurred heavy losses. In all, Republicans lost more than twenty seats in the House, and more than five seats in the Senate, resulting in the Democratic Party gaining control of the House of Representatives for the first time since 1994 [@hulse_wave_2006].

In his memoirs, President George W. Bush briefly commented on the "Revolt of the Generals." Bush wrote, "there was no way I was going to let a group of retired officers bully me into pushing out the Secretary of Defense. It would have looked like a military coup and would have set a disastrous precedent" [@bush_decision_2010, 93]. 

At the same time, it's clear that the behavior of the revolting generals forced politicians to evaluate and ultimately to stake claims, both on the personal leadership and abilities of Rumsfeld and the broader war effort in Iraq. Republicans mainly responded to the "Revolt of the Generals" by backing Rumsfeld while also suggesting the incident marked a good opportunity to evaluate the war effort more broadly, while some Democrats explicitly signaled their support for Rumsfeld stepping down [@rutenberg_general_2006-1]. 

## How Levels of Polarization and Prestige Shaped the Revolt of the Generals

Previous analyses of the "Revolt of the Generals" have examined the episode from the perspective of military dissent, the conditions under which dissent may be appropriate, and the behavioral forms that military dissent should and should not take [@cook_revolt_2008]. These are very important questions in their own right. However, previous analyses of the "Revolt" have not, in my view, fully explored the role of polarization and military prestige in enabling the "Revolt," both to occur in the first place _and_ the manner in which it unfolded. I take up this important task in the following paragraphs. 

As previously noted, the level of polarization in 2006 was not necessarily low in absolute terms, but it was relatively low compared to the levels that would emerge in the years afterward. Furthermore, the main substance of the "Revolt" involved topics and issues that are squarely related to the professional knowledge and expertise of senior military officers. These topics include issues pertaining to military strategy, and the planning and execution of military operations such as wars. 

I am not suggesting here that the planning and execution of wars are issues that cannot become polarized political topics. They most certainly can. But I am emphasizing that the planning and execution of wars and military operations are issues and topic areas that are entirely in line with the expertise of senior military officers. In fact, it is difficult to imagine a more appropriate topical area or issue for senior military officers to address than those addressed by the participants of the "Revolt of the Generals," even if one strongly condemns the manner in which these retired officers weighed in on these topics.   

The level of military prestige was relatively high in 2006. This helps explain the fact that the revolting officers undertook behaviors that were highly visible, such as the authoring of public appeals in popular newspapers/magazines, as opposed to engaging in less visible forms of behavior. It also explains why the press was interested in interviewing and subsequently publicizing the remarks of officers who did not write public appeals themselves, as in the case of Generals Riggs, Swannack, and Van Riper. Furthermore, the high level of military prestige explains the fact that politicians noted and reacted to the dissenting officers. Even President Bush's short acknowledgement in his memoirs, in my view, suggests that he felt pressure. Had the dissenting officers been part of an institution that was less prestigious than it was, it is more likely that the President would more easily have brushed off calls for taking action vis-a-vis Rumsfeld.  

### The Revolt and the Central Principles of Civil-Military Relations

With respect to the central principles of civil-military relations, the "Revolt of the Generals" clearly violated the principle of civilian control of the military. Furthermore, the episode violated the principle of non-partisanship, but indirectly. The episode did not violate the principle of non-interference of the military whatsoever.

The "Revolt of the Generals" violated the principle of civilian control. In levying blatant and public criticism against Secretary of Defense Rumsfeld, the Bush Administration had to reconsider its policy approach to the war in Iraq. This was, after all, the purpose of the "Revolt." Those who applaud the "revolting" generals are likely to point out that even as the participating military actors engaged in behavior that breached a significant principle of civil-military relations, their behavior nonetheless contributed to the achievement of a greater good, namely, the evaluation and alteration of a failing strategic policy. One such commentator of the revolt, Frederick Kagan, noted for instance that the generals who revolted helped "ensure that the people of a democracy [were] sufficiently well informed to make sound decisions about their leaders in a time of war" [@kagan_let_2006, 15]. 

The "Revolt of the Generals" indirectly violated the principle of non-partisanship, however. Here I have in mind the fact that on one hand, the participating generals did not levy any explicitly partisan remarks, while on the other, their actions nonetheless had clear partisan consequences. Remarkably, the public appeals written by Eaton, Batiste, and Newbold do not contain any explicit criticism or advocacy either in support of or against Republicans or Democrats. The criticism these authors levy against Rumsfeld is instead directed at his personality and characteristics.

Yet Donald Rumsfeld was not an apolitical figure, nor his position simply the equivalent to the CEO of a major company. Rumsfeld was the Secretary of Defense serving in the second administration of the Republican President, George W. Bush. Furthermore, as mentioned previously, the "Revolt" occurred ahead of the 2006 midterm elections, and at a time when the American people were in the process of evaluating the war in Iraq. And, also as mentioned previously, the relatively high level of military prestige that existed at the time at least contributed to members of Congress and Congressional candidates taking clear positions, both on Rumsfeld but also a number of other broader topics related to Iraq, such as possible war with Iran and the broader effort against terrorism [@howell_while_2007, 238-242]. 

In early 2007, just weeks after Rumsfeld's dismissal, the Bush Administration announced a new strategy for Iraq involving a substantial "surge" of military forces. Senator Lindsey Graham, a Republican, noted at the time that, "we never had enough troops to begin with...a month or two ago, we found out the Army is broken, and they [the American people] agreed that Shinseki was right" [Graham, 2007, quoted in @shanker_new_2007]. Perhaps it is impossible to accurately measure the influence of the "Revolt of the Generals" on the outcome of the 2006 midterm elections, or on the Bush decision to institute a surge the following year in 2007. But at the very least, the "Revolt" certainly impacted both of these (and other) events.

The "Revolt" did not violate the principle of non-interference, however. The military officers who participated in the revolt addressed topics that were in accord with reasonable expectations of military expertise, as noted in the previous section. 

### Summary

The 2006 "Revolt of the Generals" was a significant episode in US civil-military relations. Among the high profile civil-military conflicts that have occurred in US history, the "Revolt" was unique in that several recently retired military generals, all of whom had served either in Iraq or been directly involved in its planning, called on the Secretary of Defense to step down. The episode involved six retired officers making public appeals, and sparked other officers to also make additional public appeals. Additionally, the episode sparked renewed interest in the topic of military officer resignation and the conditions under which resignation was ever appropriate. To this day, the topic of military officer resignation remains one that is hotly debated by scholars [@m_shields_introduction_2017].  

The "Revolt of the Generals" violated the principles of civilian control, and to a lesser degree, the principle of non-partisanship. However, the episode did not violate the principle of non-interference. Critics of the "revolting" generals generally take issue not with the substance of the critiques levied by the outspoken officers, but with the manner and forms of behaviors used by these officers [@cook_revolt_2008; @binkley_revisiting_2020]. Supporters of these officers, on the other hand, tend to acknowledge the severity of the behaviors employed by these military actors, but instead argue that the national good was ultimately better served as a result of the episode having occurred [@kagan_let_2006]. 

\singlespace
# Part 2 - Domestic Unrest, Presidential Transition, and the Withdrawal from Afghanistan, 2020-2021
\doublespace

## Background

## Political Behaviors Undertaken

### Impact of Undertaken Behaviors

## UnderlyingIssues at Stake

### Summary

# Evaluation, Comparison and Contrast of Behaviors

## The Role of Polarization - Does Your Theory Fit? Why?

## Alternate Explanations

## Conclusion

\pagebreak