---
geometry: margin=1in
output: 
  bookdown::pdf_document2:
    number_sections: TRUE
    #indent: true
    toc: false
  #bookdown::pdf_document2: default
reference-section-title: References
bibliography: 
  - dissertation.bib
fontsize: 12pt 
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{setspace}
- \usepackage{sectsty}
- \usepackage{paralist}
- \usepackage{fancyhdr}
- \usepackage{lastpage}
- \usepackage{indentfirst}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption}
- \usepackage{rotating, graphicx}
- \usepackage{dcolumn}
- \usepackage{ragged2e}
---

```{r loading-library0, include=FALSE, echo=FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("pacman")
library("ggrepel")
library("writexl")
library("lme4")
library("bucky")
library("scales")
library("stargazer")
library("rstatix")
library("brglm")
library("margins")
library("IRdisplay")
library("texreg")
library("cowplot")
library("mice")
library("miceadds")
library("bookdown")
library("sandwich")
library("mitools")
library("modelsummary")
```

\doublespace

# Chapter Six {-}

On September 11, 2001, terrorists hijacked several commercial airplanes and purposely crashed them into several sites, including the twin towers of the World Trade Center in New York City, the Pentagon in Washington D.C., and a field in Western Pennsylvania.^[Heroic Americans prevented this plane from reaching its intended target, the US Capitol or the White House, in Washington D.C. [@kean_911_2004, 14].] The evening of the 9/11 attacks, members of Congress gathered on the steps of the US capitol to express solidarity and resolve. After a moment of silence and short remarks by then speaker of the House Dennis Hastert, those gathered sang the song, "God Bless America" [@us_congress_members_2001].

Seventeen months later, in February 2003, the Senate Armed Services Committee held a hearing regarding a likely US invasion of Iraq, which would begin the next months. The committee was interested in hearing the views of the respective military service chiefs regarding plans for the upcoming war. One exchange was particularly noteworthy. General Eric Shinseki, the Army Chief of Staff, told the committee that he thought that "something on the order of several hundred thousand soldiers" was a likely figure for stabilizing Iraq after the invasion [Shinseki, 2003 quoted in @fallows_karmic_2008]. Shinseki's remarks, however, went against the official position held by the Defense Department and it's civilian head, Donald Rumsfeld, and his second in command, Paul Wolfowitz, both of whom believed that a smaller, lighter military footprint was more than adequate for the pending task. In the days that followed, both Wolfowitz and Rumsfeld publicly criticized Shinseki's remarks. Shinseki, for his part, remained largely silent, and retired from the Army shortly afterward.  

Eighteen and a half years later, the United States withdrew from Afghanistan after the resurgent Taliban quickly displaced the Afghan National Government in the summer of 2021. During a Congressional hearing held on 28 September 2021 about the withdrawal from Afghanistan, Tom Cotton, a veteran and the junior Republican Senator from Arkansas, asked the Chairman of the Joint Chiefs of Staff, General Mark Milley, "why haven't you resigned [@singman_milley_2021]?" The Senator's question alluded to the fact that General Milley, weeks before the collapse of the Afghan Government, had recommended to President Biden that the US maintain approximately 2,500 US Troops in Afghanistan, a recommendation that the President did not heed. General Milley responded to Cotton by noting that, "It would be an incredible act of political defiance for a commissioned officer to just resign because my advice is not taken...this country doesn't want generals figuring out what orders we are going to accept and do or not. That's not our job [@singman_milley_2021]." Furthermore, it is important to note that in the same hearing, before ever even discussing Afghanistan, General Milley briefly addressed allegations that he had inappropriately conversed with both a Chinese general and with the Speaker of the House, Nancy Pelosi, during the transition between President Trump and Biden [@baldor_watch_2021]. These allegations had provoked strong partisan reactions in the weeks leading up to the hearing. 

Cotton's question by itself was not necessarily inappropriate, but it is highly likely that Cotton wished that Milley either would have resigned, or that Milley would at least have signaled to the public that he considered resigning as a result of the chaotic withdrawal from Afghanistan. Such an intimation from the mouth of a sitting Chairman of the Joint Chiefs of Staff would have generated significant political points for Congressional Republicans, as well as cause a fairly significant rift in civil-military relations at an already tense moment. 

This chapter identifies, compares, and contrasts the patterns of political behavior engaged in by military and civilian actors throughout the post-9/11 era. From approximately 2001 - 2021, the level of military prestige remained high and relatively constant, but the level of political polarization rose sharply. As the preceding paragraphs note, civil-military tension exists when polarization is relatively low, as was the case in 2003, and when it is much higher, as in 2021. This chapter, however, explores how the political behaviors that occur in low versus highly polarized settings, but that involve a relatively prestigious military, are nonetheless different.  
The chapter proceeds in four parts. First, I describe the predictions concerning the patterns of political behavior as a state moves from a domestic environment marked by low polarization and high military prestige to one marked by high polarization and high military prestige. In this section, I also substantiate the claim that such a movement approximates the change in the domestic environment that occurred in the United States from 2001 - 2021. In the second part of this chapter, I describe several political behaviors associated with one significant civil-military relations episode that occurred during the first half of this nearly twenty year period. The episode that I focus on is the so-called 2006 "Revolt of the Generals," which involved several retired senior military officers publicly castigating the Secretary of Defense, Donald Rumsfeld. 

The third part of the chapter explores an episode that occurred during the second half of the period. I focus on the political behaviors involving the military undertaken in the summer of 2020 and throughout 2021, which involved periods of domestic civil unrest, a turbulent presidential election, and the US withdrawal of Afghanistan. The fourth and final part of this case study evaluates, compares, and contrasts these two episodes. 

Similar to the previous chapter, this chapter cannot, and therefore does not attempt, to recount or list every political behavior that occurred during each episode. Instead, my central focus is on the tenor, characteristics, tone, and underlying reasons for some of the most glaring or noteworthy political behaviors associated with each episode. The chief argument made in this chapter is that there is moderate evidence to support the claim, in accordance with this dissertation's theoretical predictions, that the rise in the level of political polarization occurring from 2001 - 2021 has contributed to increases in both the politicization of the military and the undertaking of more extreme forms of political behavior by military and civilian actors.

# Theoretical Predictions for the Post-9/11 Era, 2001-2021

This section of the chapter describes the theoretical predictions for the patterns of political behavior involving the military that occur as the domestic environment changes. In particular, I will describe how the political behaviors change as the domestic environment shifts from one marked by relatively low polarization and high prestige to one characterized by high polarization and high military prestige. 

## Low Polarization and High Military Prestige

The theory presented in chapter two of this dissertation argued that the state's domestic environment strongly shapes the patterns of political behavior that occur involving the military. When the level of military prestige is high, both civilian and military actors are more likely to believe that the military can influence public opinion. For this reason, the theory argued that both types of actors will behave in ways that harness or leverage the prestige of the military to appeal to the public. 

This theoretical expectation is rooted in the political communication literature, which argues that the public listens to those voices which are considered "knowledgeable" and "trustworthy" [@lupia_democratic_1998, 69-76; see also @zaller_nature_1992, 47]. The theory argued that as the level of military prestige rises, the public increasingly considers the military and its members to possess these traits. 

When the level of polarization is relatively low, however, the theory predicts that there is relatively little basis for ideological division between and among military actors. Unlike when polarization exists at a high level, there are few if any major moral-worldview differences at work among military actors. This is not to say that military actors are entirely unified, or that there are no major disagreements whatsoever between and among military actors, when polarization is low. But the theory does predict that absent high polarization, there is little if any ideological basis, or _motive_ that exists, for military actors to behave in ways that violate the central principles of civil-military relations.

In an environment marked by low polarization and high military prestige, the theory ultimately predicts that either military or civilian actors may induce politicization of the military, and engage in behaviors that violate the principles of civil-military relations. The theory even noted that such an environment is far from tranquil, and that there is a high potential for civil-military conflict altogether, as shown in the bottom left quadrant of Table \@ref(tab:poss-environment3). 

```{r poss-environment3, echo=FALSE, include=TRUE, error=FALSE, message=FALSE}

text_env_revised2 <- tribble(~"Environment", ~"Low Polarization", ~"High Polarization",
  "Low Prestige", "No politicization; low levels of political activity; low potential for civil-military conflict", 
  "Military (induced) politicization; moderate levels of political activity; moderate potential for civil-military conflict",  
  "High Prestige", "Military and Civilian (induced) politicization; moderate levels of political activity; high potential for civil-military conflict", 
  "Comprehensive politicization; extreme levels of political activity; significant potential for civil-military conflict" 
  )
 

  kable(text_env_revised2, booktabs=T, caption="Types of Domestic Environments for Civil-Military Relations and their Impact on Political Behaviors and the Conduct of Civil-Military Relations") %>%
  kable_styling(full_width = F, font_size=10, latex_options="striped") %>% 
  column_spec(1, width="8em") %>% 
  column_spec(2:3, width="18em") 
 
```

Throughout US history, there have been several significant civil-military relations disputes that have occurred under conditions of low polarization and high military prestige. For example, disputes between General Douglas MacArthur and President Harry Truman over Korea, General Matthew Ridgeway and President Dwight Eisenhower over Ike's New Deal policies in the early 1950s, and Colin Powell and President Bill Clinton over several post-Cold War crises in the early 1990s all occurred in domestic environments marked by relatively low polarization and high military prestige. Therefore, we should not automatically conclude that the conduct of civil-military relations will be smooth or without tension when polarization is low and when military prestige is high. 

## High Polarization and High Military Prestige

As the level of polarization rises, however, the theory predicts that the potential for civil-military conflict increases. This occurs, moreover, because rising political polarization is associated with increasing ideological division occurring within and among military actors who, like the rest of society, become more polarized as the level of polarization grows. 

At the core, rising levels of polarization attest to a loss of consensus or agreement concerning the moral values and ideals that should hold sway in society [@marsden_twilight_2014; @mann_its_2012; @allen_why_2020]. In highly polarizing times such as the contemporary age in the United States, the conflicts that emerge are not simply political in the sense that they involve whether society should enact policy "X" or policy "Y" with respect to any number of polarizing issues, but rather a conflict over the worldviews that underlie policy "X" or policy "Y". Chapter two noted several instances of polarizing issues in the contemporary United States, including the question of what it means to be an American, whether America should be proud of its history, and whether the construct of gender is important anymore. What these and many other polarizing issues have in common is that they cannot be solved nor even intelligently discussed without an individual referencing or alluding to some sort of moral-worldview commitment. 

The reason that polarization matters for the conduct of civil-military relations, chapter two argued, is because one's commitment to any number of polarizing issues may in fact be stronger than one's commitment to the central principles of civil-military relations. Chapter two noted in particular that in highly polarizing environments, military actors more frequently encounter situations in which they have to choose between adhering to the principles of civil-military relations or advancing one or more polarizing issues with which they strongly identify or believe in. 

As Table \@ref(tab:poss-environment3) notes, high polarization and high military prestige thus leads to an environment in which there is comprehensive politicization, extreme levels of political activity involving the military, and significant potential for civil-military conflict. These characteristics are even more severe than those that exist in a domestic environment marked by low polarization and high military prestige. 

## Substantiation of Environmental Characteristics, 2001 - 2021

To substantiate the claim that the level of polarization increased and the level of military prestige generally remained constant (but relatively high) over the period of 2001 - 2021, i rely on measurable quantitative data. This is different from the previous chapter that explored the post-US Civil War era, which relied on mainly qualitative data. Figure \@ref(case2-polar) plots the level of polarization in both the US House and the US Senate, as well as the level of trust and confidence in the US military, from 2001 - 2021. Data for polarization uses roll-call voting data on members of Congress over time [@jeffrey_b_lewis_voteview_2020], and data for trust in the military comes from Gallup public opinion data regarding the level of public trust and confidence in the military [@noauthor_confidence_2020].  

```{r case2-polar-prep, include=FALSE, warning=FALSE, echo=FALSE}

df_trust_polar_cs2 <- read_csv(here("data","polar-inst-trust-time.csv"))

df_trust_polar_cs2 <- df_trust_polar_cs2 %>% filter(Year >= 2001) %>% select(-Congress) %>% 
  mutate(
    house_polar_dim1 = house_polar_dim1 *100,
    sen_polar_dim1 = sen_polar_dim1 *100,
    ) %>% pivot_longer(Military:sen_polar_dim1, names_to = "value", values_to = "Score") %>% distinct()
```

```{r case2-polar, include=T, warning=FALSE, echo=F, message=F, fig.height=4, fig.width=6, fig.cap = "Public Trust (L) and Polarization (R), 2001-2021" }
df_trust_polar_cs2 %>% ggplot(aes(x=Year, y=Score, color=value)) + geom_line() + geom_point(size=0.7) + labs(x="Year", y="Trust in Institutions") + scale_color_discrete(name="Trust/\nPolarization",
                         breaks=c("Military", "house_polar_dim1", "sen_polar_dim1"),
                         labels=c("Trust in\nMilitary","House\nPolarization", "Senate\nPolarization")) + scale_y_continuous(sec.axis = sec_axis(~., name = "Polarization (Scaled)")) + scale_x_continuous(breaks=seq(2001, 2021, 2)) + theme(legend.position="bottom") + guides(color = guide_legend(ncol=3,nrow=2,byrow=TRUE))
```

Figure \@ref(case2-polar) shows that each variable, polarization and military prestige, has changed differently over the period from 2001-2021. With the exception of a few years, polarization levels in both the US House and Senate have consistently increased. The level of public trust in the military, on the other hand, has exhibited more variation between years. 

For example, a relative spike in public trust occurred in 2003, perhaps reflecting a boost in patriotic fervor as the war in Iraq started the same year. Similarly, local minima occurred in 2007, perhaps coinciding with impatience over mounting US casualties in Iraq, and again in 2021, perhaps reflecting public frustration over the US withdrawal from Afghanistan. Yet even in the presence of this variability, the level of military prestige that has endured for the past twenty years is relatively high, and certainly higher than the level that existed in the late 1970s and early 1980s after the Vietnam War. Since data has been kept, for example, the lowest recorded level of trust in the military occurred in 1981, when only 50% of surveyed Americans said they trusted the military a "great deal or quite a lot" [@noauthor_confidence_2020]. In short, the period 2001 - 2021 reflects a domestic environment in which the level of military prestige remains relatively high, but in which the level of polarization increases. 

Note, however, that Figure \@ref(case2-polar) does not, and indeed cannot, definitively conclude when the shift from a relatively moderate to relatively high level of polarization occurred. What we do see, however, is a fairly noteworthy spike in the level of polarization in the US House (red line) beginning in 2011, and that this spike was accompanied by a smaller yet still detectable spike in polarization in the US Senate (blue line) the same year. Furthermore, we know that this occurred midway through the first term of President Barack Obama. 

From a methodological standpoint, then, the analysis which follows takes the year 2011 as a sort of natural breaking point. I begin by describing several political behaviors involving the military that occurred prior to 2011, and then shift to examining several that occurred after 2011. I then compare and contrast these behaviors, and explain how rising polarization shaped those behaviors that occurred in the second half of the period. 

\singlespace
# Part 1 - The War in Iraq and The Revolt of the Generals, 2004 - 2006
\doublespace

In the first section of this case study, I describe the 2006 "Revolt of the Generals." After describing the political behaviors that occurred as part of this episode, I clarify the reasons why this episode occurred in the first place, and distill the underlying issues that were ultimately at stake in this episode. While this dissertation certainly speaks to the normative implications of the "Revolt of the Generals" with respect to civil-military conduct, my primary task here is not to render a definitive evaluation of the conduct that occurred as part of this episode. Others [@cook_revolt_2008; @snider_strategic_2014] have written extensively on this. Rather, my primary task is to show how the a relatively low level of polarization and a relatively high level of military prestige enabled the "Revolt of the Generals" to occur. 

## Background

The 2006 "Revolt of the Generals" narrowly refers to the behavior of several recently retired senior military officers who, in the Spring of 2006, publicly criticized the Secretary of Defense, Donald Rumsfeld, concerning Rumsfeld's leadership of the war in Iraq. By 2006, the war in Iraq was three years old, and more importantly, it was clear that hopes for a quick and relatively easy war had been dashed. 

In a broader sense, the "Revolt of the Generals" must be viewed as the military's backlash to their perception of not being heard, listened to, or valued by their civilian chain of command within the Department of Defense [@snider_strategic_2014; @binkley_revisiting_2020]. In a way, the "Revolt of the Generals" signified a reaction by several retired military officers to the way that civilians had disparaged General Shinseki prior to the war's beginning, and the way that he was treated by his civilian superiors after he candidly presented Congress with his views on force levels required to stabilize Iraq (as noted in this chapter's introduction).  

As is always the case, the timing of political behaviors undertaken matters, and those associated with the "Revolt of the Generals" are no exception. Not only was the war in Iraq three years old in the Spring of 2006, but also later that year, Congressional midterm elections were held. The behavior of a small but influential number of retired senior military officers ultimately elevated the status of the war in Iraq, and would contribute to the ultimate dismissal of Rumsfeld by the President, George W. Bush.   

## Political Behaviors Undertaken

The immediate slate of military actors who participated in the "Revolt of the Generals" include a total of six retired military officers, including Army Major General Paul Eaton, Marine Lieutenant General Greg Newbold, Army Major General John Batiste, Marine Lieutenant General Paul Van Riper, Army Major General John Riggs, and Army Major General Charles Swannack. In the paragraphs that follow, I describe the behaviors undertaken by these six officers. 

However, I then describe a number of political behaviors undertaken by other military and civilian actors in response to the behavior of these six primary military actors. By considering the behaviors undertaken both as part of the "Revolt" _and_ those undertaken directly in response to it, the reader is better positioned to understand the full impacts of the episode. 

### Public Appeals

The primary political behavior undertaken by military actors during the "Revolt of the Generals" was that of making public appeals. In March and April of 2006, Generals Eaton, Batiste, and Newbold penned articles that appeared in _The New York Times_, _The Washington Post_, and _Time_ magazine, respectively [@eaton_top-down_2006; @batiste_case_2006; @newbold_why_2006]. Shortly afterward, Generals Van Riper, Riggs, and Swannack gave interviews in which they "concurred with the criticisms levied by Eaton, Newbold, and Batiste" [@binkley_revisiting_2020, 23].

The public appeals made by the military actors who participated in the "Revolt" charged Rumsfeld with micromanaging the war effort, failing to heed the advice of military professionals during the planning and execution of the war, and insisting on a strategic course of action bound for failure. In addition to the outspoken and highly visible nature of the public appeals made, the public appeals undertaken were unique in that they explicitly stated that in order for the US to prevail in Iraq, "Rumseld must go" [@eaton_top-down_2006]. 

Somewhat unsurprisingly, some military actors then defended Rumsfeld, and while his defenders likewise did not necessarily constitute a large number of actors, they nonetheless also engaged in making public appeals. For instance, four retired generals - Vallely, McInerney, Crosby, and Moore - penned an op-ed in _The Wall Street Journal_ in April of 2006 entitled, "In Defense of Donald Rumsfeld" [@john_crosby_defense_2006].

### Grandstanding

A secondary type of political behavior involved in the "Revolt of the Generals" was grandstanding, which Risa Brooks has defined as the "threat or actual resignation in protest of policy" [@brooks_militaries_2009, 219]. While no officers (to my knowledge) actually resigned or threatened to resign as a result of or in conjunction with the "Revolt of the Generals," the entire episode sparked somewhat of a debate among civil-military relations scholars as well as among military officers regarding the role, if any, of the behavior of military officer resignation and the conditions under which such a behavior should be considered. 

Here, I do not take up the contours of that debate. Chapter two discussed this behavior in detail, and made the point that even among scholars, there is a lack of consensus about if and when resignation should be considered a viable option by military actors.^[For example, see @dubik_principled_2014; @snider_strategic_2014; and @milburn_breaking_2010 for arguments who support the act of resigning under very strict conditions. See also, however, @m_shields_introduction_2017 and @golby_beyond_2015 for alternate views.] However, it is rather important to at least acknowledge the debate over resignation that occurred in the wake of the "Revolt of the Generals" because many felt that General Shinseki, three years prior, would have been entirely justified to have resigned in protest in 2003 rather than acquiesce begrudgingly to a war plan that he believed would result in failure.

Finally, it should be said that the military actors who participated in the "Revolt of the Generals" were essentially insisting that their former civilian boss, the Secretary of Defense, resign. Though Rumsfeld did not heed such requests, this aspect marked the "Revolt" as different from previous civil-military disputes that had occurred in American history.  

### Impact of Undertaken Behaviors

The entire "Revolt of the Generals" episode, and the political behaviors undertaken by the military actors involved in it, were rather visible to the American public. Especially in a midterm election year, these behaviors likely contributed to shaping the views of Americans on the war in Iraq. 

Donald Rumsfeld, the Secretary of Defense, remained at his post until one week after the mid-term elections in which the Republicans incurred heavy losses. In all, Republicans lost more than twenty seats in the House, and more than five seats in the Senate, resulting in the Democratic Party gaining control of the House of Representatives for the first time since 1994 [@hulse_wave_2006].

In his memoirs, President George W. Bush briefly commented on the "Revolt of the Generals." Bush wrote, "there was no way I was going to let a group of retired officers bully me into pushing out the Secretary of Defense. It would have looked like a military coup and would have set a disastrous precedent" [@bush_decision_2010, 93]. 

## How Levels of Polarization and Prestige Shaped the Revolt of the Generals

At its core, the entire "Revolt of the Generals" episode deals with the question of how military officers should dissent with their civilian leaders, the circumstances under which dissent is appropriate, and what forms military officer dissent should (and conversely, should not) take. Those who participated in the "Revolt of the Generals," and those who applauded the generals involved in it, argue that the behavior of the revolting generals was a great public service in that it ensured "that the people of a democracy [were] sufficiently well informed to make sound decisions about their leaders in a time of war" [@kagan_let_2006, 15]. Critics of the outspoken generals, on the other hand, raised a number of concerns, ranging from criticizing civilian leadership altogether, to leveling public and visible criticism during a time of war, when US forces were engaged in hostile action [@snider_strategic_2014].

But something that previous analyses of the "Revolt" have not fully explored is the role of polarization and military prestige in enabling the "Revolt" to occur in the first place, and the manner in which it unfolded. As I have already noted, the level of polarization in 2006 was not low in absolute terms, but it was relatively low compared to the levels that would emerge in the years which followed, and relatively higher than the level of polarization that had existed, for example, a decade earlier. 

Yet the main substance of the "Revolt of the Generals" centered on topics that, although they certainly included some partisan aspects, were largely bipartisan in nature. These topics included the planning and execution of a war that had been initiated with a high degree of bipartisan support. Regardless of what one thinks about the normative appropriateness of the "Revolt of the Generals," it is vital to consider, at least for the purposes of this dissertation, that the substance of the "revolt" was the perceived inadequacy of wartime leadership of the civilian chain of command, namely, Donald Rumsfeld. Critics may, rightly in my view, argue that in expressing dissent, the revolting generals violated the principle of civilian control, but in my view, they did not violate either the principle of non-partisanship nor the principle of non-interference. 

Say some more about that here. And then cite Jon Pevehouse. 

### Summary

\singlespace
# Part 2 - Domestic Unrest, Presidential Transition, and the Withdrawal from Afghanistan, 2020-2021
\doublespace

## Background

## Political Behaviors Undertaken

### Impact of Undertaken Behaviors

## UnderlyingIssues at Stake

### Summary

# Evaluation, Comparison and Contrast of Behaviors

## The Role of Polarization - Does Your Theory Fit? Why?

## Alternate Explanations

## Conclusion

\pagebreak